### 2016-08-23 Project1/Project1HK.tex
First checkin

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{Computational Astrophysics Project 1 - Investigation of Lutz-Kelker bias, through Monte Carlo simulations, on estimation of stellar distance using trigonometric parallax}
\begin{document}
\maketitle
\begin{abstract}
Monte Carlo simulations were carried out with a randomly generated sample of 400,000 stars and derived absolute magnitudes and distances were retrieved from the sample with assumed parallax and photometric errors. The resulting derived data were compared with the 'true' data and the error dispersion was found to be consistent with the well known Lutz-Kelker bias.
\end{abstract}
{\bf Keywords:} Lutz-Kelker bias, Monte Carlo simulation, stars: distances, magnitude, parallax

\section{Introduction}
The Lutz-Kelker bias was first quantified \citep{lutz} using the absolute magnitude of stars estimated with trigonometric parallax. It occurs because the number of observable stars increases with the square of the distance. For any given parallax measurement there will be an error which will give an upper and lower bound of retrieved distance, and stars at a smaller distance and stars further away can scatter into this volume, however as there are more stars outside the sampled volume than inside, the sample will be biased. This effect causes a systematic bias such that measured parallaxes will on average yield too small distances \citep{oud}. Monte Carlo simulations are ideally placed to investigate this phenomenom and provide corrections and the necessary calibration to trigonometric parallax measurements.
\newpage
\section{Monte Carlo simulations}
Four Monte Carlo simulations were carried out, each with a sample of 400,000 stars having a true absolute magnitude of 0.6 (with a Gaussian dispersion of 0.001 mag.). The stars were assigned true distances of between 1pc and 8Kpc. A small $\sigma$ photometric error of 0.01 mag. in the measurement of the retrieved absolute magnitude was assumed in each case. Please note that the original LKbiasMSc.f program had a small bug associated with photometric error - this was corrected prior to running these simulations. Only stars within a retrieved distance of 4Kpc were considered.

\subsection{Task 1}
For this simulation a constant spatial density distribution of stars was assumed. This simulation employed a constant parallax error of 50 microarcsec. There was no apparent magnitude limit. The total mean and standard deviation of the retrieved absolute magnitudes are shown in Table \ref{tab:task1} together with the mean and standard deviations after the results were split into 1000pc bins. This is shown graphically in Figures \ref{fig:t1bin1} to \ref{fig:t1bin4}. A scatter plot of retrieved absolute magnitude as a function of retrieved distance is shown in Figure \ref{fig:t1graph1}. Figures \ref{fig:t1graph2} and \ref{fig:t1graph3} show the number distribution of stars for true and retrieved distance respectively.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
Distance (pc) & Mean & Std Deviation \\
\hline
Total & 1.0355 & 0.4797 \\
0-1000 & 0.6122 & 0.1055 \\
1001-2000 & 0.6945 & 0.2049 \\
2001-3000 & 0.8561 & 0.3629 \\
3001-4000 & 1.1057 & 0.4977 \\
\hline
\end{tabular}
\caption{\label{tab:task1}Mean retrieved absolute magnitudes (Task 1)}
\end{table}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task1/Bin1}
\caption{Task 1 -  Distribution of Retrieved Abs. Mag. (0-1000 pc)}
\label{fig:t1bin1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task1/Bin2}
\caption{Task 1 - Distribution of Retrieved Abs. Mag. (1001-2000 pc)}
\label{fig:t1bin2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task1/Bin3}
\caption{Task 1 - Distribution of Retrieved Abs. Mag. (2001-3000 pc)}
\label{fig:t1bin3}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task1/Bin4}
\caption{Task 1 - Distribution of Retrieved Abs. Mag. (3001-4000 pc)}
\label{fig:t1bin4}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task1/Graph1}
\caption{Task 1 - Retrived Absolute Magntude as a function of Retrieved Distance}
\label{fig:t1graph1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task1/Graph2}
\caption{Task 1 - Number distribution of stars by True Distance}
\label{fig:t1graph2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task1/Graph3}
\caption{Task 1 - Number distribution of stars by Retrieved Distance}
\label{fig:t1graph3}
\end{figure}
\newpage
\subsection{Task 2}
This simulation used the same parameters as Task 1 except an apparent magnitude limit of 14 was assumed. Table \ref{tab:task2} and figures \ref{fig:t2bin1} to \ref{fig:t2graph3} are the same plots as used for Task 1.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
Distance (pc) & Mean & Std Deviation \\
\hline
Total & 0.7992 & 0.3082 \\
0-1000 & 0.6069 & 0.1182 \\
1001-2000 & 0.6915 & 0.1963 \\
2001-3000 & 0.8363 & 0.3252 \\
3001-4000 & 0.7929 & 0.3061 \\
\hline
\end{tabular}
\caption{\label{tab:task2}Mean retrieved absolute magnitudes (Task 2)}
\end{table}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task2/Bin1}
\caption{Task 2 -  Distribution of Retrieved Abs. Mag. (0-1000 pc)}
\label{fig:t2bin1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task2/Bin2}
\caption{Task 2 - Distribution of Retrieved Abs. Mag. (1001-2000 pc)}
\label{fig:t2bin2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task2/Bin3}
\caption{Task 2 - Distribution of Retrieved Abs. Mag. (2001-3000 pc)}
\label{fig:t2bin3}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task2/Bin4}
\caption{Task 2 - Distribution of Retrieved Abs. Mag. (3001-4000 pc)}
\label{fig:t2bin4}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task2/Graph1}
\caption{Task 2 - Retrived Absolute Magntude as a function of Retrieved Distance}
\label{fig:t2graph1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task2/Graph2}
\caption{Task 2 - Number distribution of stars by True Distance}
\label{fig:t2graph2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task2/Graph3}
\caption{Task 2 - Number distribution of stars by Retrieved Distance}
\label{fig:t2graph3}
\end{figure}

\newpage
\subsection{Task 3}
This simulation used the same parameters as Task 1 except the spatial density of stars was no longer assumed to be constant but decreasing as $R^{-2}$ where R is the distance between 1pc and 8Kpc. Table \ref{tab:task3} and figures \ref{fig:t3bin1} to \ref{fig:t3graph3} are the same plots as used for Task 1.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
Distance (pc) & Mean & Std Deviation \\
\hline
Total & 0.7859 & 0.3785 \\
0-1000 & 0.6095 & 0.0792 \\
1001-2000 & 0.6528 & 0.1822 \\
2001-3000 & 0.7404 & 0.3158 \\
3001-4000 & 0.8842 & 0.4541 \\
\hline
\end{tabular}
\caption{\label{tab:task3}Mean retrieved absolute magnitudes (Task 3)}
\end{table}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task3/Bin1}
\caption{Task 3 -  Distribution of Retrieved Abs. Mag. (0-1000 pc)}
\label{fig:t3bin1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task3/Bin2}
\caption{Task 3 - Distribution of Retrieved Abs. Mag. (1001-2000 pc)}
\label{fig:t3bin2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task3/Bin3}
\caption{Task 3 - Distribution of Retrieved Abs. Mag. (2001-3000 pc)}
\label{fig:t3bin3}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task3/Bin4}
\caption{Task 3 - Distribution of Retrieved Abs. Mag. (3001-4000 pc)}
\label{fig:t3bin4}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task3/Graph1}
\caption{Task 3 - Retrived Absolute Magntude as a function of Retrieved Distance}
\label{fig:t3graph1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task3/Graph2}
\caption{Task 3 - Number distribution of stars by True Distance}
\label{fig:t3graph2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task3/Graph3}
\caption{Task 3 - Number distribution of stars by Retrieved Distance}
\label{fig:t3graph3}
\end{figure}

\newpage
\subsection{Task 4}
This simulation used the same parameters as Task 2 (an apparent magnitude limit of 14) except the parallax error was no longer assumed to be constant but had a GAIA-like error distribution. Table \ref{tab:task4} and figures \ref{fig:t4bin1} to \ref{fig:t4graph3} are the same plots as used for Task 1.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
Distance (pc) & Mean & Std Deviation \\
\hline
Total & 0.6286 & 0.0944 \\
0-1000 & 0.5866 & 0.0834 \\
1001-2000 & 0.6017 & 0.0239 \\
2001-3000 & 0.61 & 0.0554 \\
3001-4000 & 0.6376 & 0.1067 \\
\hline
\end{tabular}
\caption{\label{tab:task4}Mean retrieved absolute magnitudes (Task 4)}
\end{table}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task4/Bin1}
\caption{Task 4 -  Distribution of Retrieved Abs. Mag. (0-1000 pc)}
\label{fig:t4bin1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task4/Bin2}
\caption{Task 4 - Distribution of Retrieved Abs. Mag. (1001-2000 pc)}
\label{fig:t4bin2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task4/Bin3}
\caption{Task 4 - Distribution of Retrieved Abs. Mag. (2001-3000 pc)}
\label{fig:t4bin3}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task4/Bin4}
\caption{Task 4 - Distribution of Retrieved Abs. Mag. (3001-4000 pc)}
\label{fig:t4bin4}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task4/Graph1}
\caption{Task 4 - Retrived Absolute Magntude as a function of Retrieved Distance}
\label{fig:t4graph1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task4/Graph2}
\caption{Task 4 - Number distribution of stars by True Distance}
\label{fig:t4graph2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Task4/Graph3}
\caption{Task 4 - Number distribution of stars by Retrieved Distance}
\label{fig:t4graph3}
\end{figure}

\newpage
\section{Discussion}
The mean of the total sample, 1.0355 as shown in Table \ref{tab:task1}, has clearly diverged from the true mean absolute magnitude of 0.6. With a dispersion (standard deviation) of 0.4797 it is not a true representive of the sample: there is a bias. Table \ref{tab:task1} and Figures \ref{fig:t1bin1} to \ref{fig:t1bin4} from the Task 1 simulation show that the margin of error in retrieved absolute magnitude grows with the retrieved distance. Stars within 2Kpc are showing a mean absolute magnitude that is approximately 0.6 but for more distant stars the standard deviation grows sharply with distance. The scatter plot in Figure \ref{fig:t1graph1} illustrates this further with a strong tendency towards larger (less bright) absolute magnitudes with greater distance.

Figures \ref{fig:t1graph2} and \ref{fig:t1graph3} help explain the above error margin by showing that the number distribution of stars selected by retrieved distance is far greater than those selected by true distance. This is consistent with the Lutz-Kelker bias: the retrieved distances are seen to be smaller than they acually are, causing a greater number of more distant stars to be pulled into the sample than nearby stars, creating a bias.

In addition to the Lutz-Kelker bias another form of bias needs to be taken into account and that is caused by error propogation. In this case as $R=1/p$ (R=distance, p=parallax) there is an inverse relationship between the measurement (parallax) and the derived quantity (distance) which leads to an asymmetric error i.e. the error will become greater the more distant the star.

In Task 2 an apparent magnitude limit of 14 was assumed. This simulation resulted in a much improved total mean absolute magnitude and lower error dispersions in the various bins in Table \ref{tab:task2} and Figures \ref{fig:t2bin1} to \ref{fig:t2bin4}. This can be attributed to the number distribution of stars sampled for true and retrieved distances, as can be seen in Figures \ref{fig:t2graph2} and \ref{fig:t2graph3} the number of stars in both figures are similar. Lutz determined that in a magnitude limited sample the Lutz-Kelker corrections should not be applied \citep{lutz2}. There is however a bias still associated with a magnitude limited sample: the Malmquist bias, which is the preferential selection of intrinsically bright objects.

Task 3 assumed a stellar spatial density decreasing by $R^{-2}$ whereas the previous simulations all assumed a constant spatial density. The results, Table \ref{tab:task3} and Figures \ref{fig:t3bin1} to \ref{fig:t3graph3} again show a reduction in the error. So it seems that an exponentially declining spatial stellar density does go some way to counteract the Lutz-Kelker bias. The Milky Way is disk shaped so is clearly not uniform in structure and some Kutz-Kelker calibrations do attempt to take scale height, direction and large scale structure into account \citep{smith}. There is an exponential decrease of projected density with galactocentric radius for spiral galaxies \citep{smith}.

Finally, Task 4 introduced a Gaia-like error approximation, together with the apparent magnitude limit of 14 from Task 2. This simulation produced results with very low margins of error, see Table \ref{tab:task4} and Figures \ref{fig:t4bin1} to \ref{fig:t4graph3}, even at large distances. Lutz-Kelker corrections and calibration therefore can, and should, be taken into account when estimating distance with trigonometric parallax. However, as can be seen from the Gaia-like error corrections, these calibrations themselves also need revision as technological improvements increase the accuracy of parallax measurements.

\begin{thebibliography}{1}
\bibitem[Lutz(1983)]{lutz2}Lutz, T.E., Nearby Stars and the Stellar Luminosity Function, 1983, IAU Coll, 76
\bibitem[Lutz and Kelker(1973)]{lutz}Lutz, T.E., Kelker, D.H., On the use of Trigonometric Parallaxes for the Calibration of Luminosity Systems: Theory, 1973, Publications of the Astronomical Society of the Pacific, 85, 573
\bibitem[Oudmaijer et al.(1997)]{oud}Oudmaijer. R.D., Groenewegen, M.A.T., Schrijver, H., The Lutz-Kelker bias in trigonometric parallaxes, 1997, Monthly Notices of the Royal Astronomical Society, 294, 41-46 
\bibitem[Smith(1999)]{smith}Smith, H., The Lutz-Kelker effect in the Hipparcus era and beyond, 1998, Modern Astrometry and Astrodynamics, 1999
\end{thebibliography}
\end{document}

### 2016-08-23 Project3/example-codes_project3/Project3HK.tex
First checkin

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{Computational Astrophysics Project 3}
\begin{document}
\maketitle
\section{Problem 1B: Heat Conduction}
\subsection{Part a}
In order to solve this problem it is first necessary to set the initial and boundary conditions. Variables A, B and C were defined according to equation 15 in the notes. Variable A was set to $\mu$ because of $\mu u^n_{j-1}$ in equation 15. In Octave this is simply:
\begin{lstlisting}
A = mu;
\end{lstlisting}
Variable B was set to:
\begin{lstlisting}
B = 1-(2*mu*(1+(1/(j-1))));
\end{lstlisting}
This is in order to satisfy $1-2\mu(1+\frac{1}{j-1})u^n_j$ in equation 15. Finally C was set to:
\begin{lstlisting}
C = mu*(1+(2/(j-1)));
\end{lstlisting}
This satisfies the last part of equation 15: $\mu(1+\frac{2}{j-1})\mu^n_{j+1}$.
\newline
The boundary condition at the surface was set to zero:
\begin{lstlisting}
unew(N) = 0;
\end{lstlisting}
The boundary condition at the centre was set to the value above it on each iteration i.e. the boundary condition at the centre cannot be calculated so it is just assumed to be the same as the value above it. In Octave this was coded as:
\begin{lstlisting}
unew(1) = unew(2);
\end{lstlisting}
The problem required the distributions to be between $\tau=0$ and $\tau=0.2$ so I modified the total time steps to be 2400 because $\Delta_x=1/\left(N-1\right)=\frac{1}{49}$ and so $\Delta_{\tau}=\mu\Delta_x^2=\frac{1}{12005}$ which leads to $n=\frac{\tau_n}{\Delta_{\tau}}+1$ and so $n=2402$.

The plot of the temperature distribution is shown in Fig. \ref{fig:problem1_1}. Each line represents a timestep of 100. At time 1 the temperature gradient is almost vertical as there has been no time for cooling but at a time of 100 the temperature gradient forms a steep curve. The curve gets gradually less steep as time goes on until at time 2400 the curve has flattened considerably and the temperature at the centre has cooled to approximately 0.3.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem1_1}
\caption{Plot of temperature distribution}
\label{fig:problem1_1}
\end{figure}

\subsection{Part b}
The following lines of code in Octave retrieve the temperature when it has cooled to exactly half its initial value. The first line retrieves all values at position 2 i.e. the centre (the actual centre is position 1 but this was just set to be the same as position 2 in the boundary conditions). The second line finds the smallest timestep value (n) where temperature (u) is less than 1/2. 
\begin{lstlisting}
u1 = approx(2,:);
nc = min(find(u1<0.5))
\end{lstlisting}
The timestep value n retrieved from the above has a value of 1629. $\tau_n=(n-1)\Delta_{\tau}=\frac{1628}{12005}$. From the notes $\tau$ is in units of $t/\left(a^2/\eta\right)$. Therefore time in seconds=$\tau\left(a^2/\eta\right)$ where $a=100,000m$ and $\eta=10^{-6}m^2 sec$, giving the time it takes the centre to cool to half its initial temperature as $4.3\times 10^7$ years.

Full source code listing for Problem 1B is in Appendix 1.

\newpage
\section{Problem 2: Circular Binary}
\subsection{Part a} 
For a binary system where the centre of mass is at the origin then: $m_1\mathbf{r_1}=m_2\mathbf{r_2}$ and we can express the stars positions using just their mass. The more massive star would lie closer to the centre of mass at a radius that is inversely proportional to its mass.
The equation of motion (for the primary star) is given by:
\begin{equation}
m_p\frac{d^2\mathbf{x_p}}{dt^2}=\frac{Gm_pm_s}{|\mathbf{x_s}-\mathbf{x_p}|^3}\left(\mathbf{x_s}-\mathbf{x_p}\right)
\end{equation}
If we substitute $\mathbf{m_p}=m\mathbf{\tilde{m_p}}$, $\mathbf{x_p}=a\mathbf{\tilde{x_p}}$, $\mathbf{x_s}=a\mathbf{\tilde{x_s}}$ and $t=\tilde{t}\sqrt{a^3/Gm}$ into equation 1 then this leads to:
\begin{equation}
m\mathbf{\tilde{m_p}}\frac{d^2a\mathbf{\tilde{x_p}}}{d\left(\tilde{t}\sqrt{a^3/Gm}\right)^2}=\frac{Gm\mathbf{\tilde{m_p}}m\mathbf{\tilde{m_s}}}{|a\mathbf{\tilde{x_s}}-a\mathbf{\tilde{x_p}}|^3}(a\mathbf{\tilde{x_s}}-a\mathbf{\tilde{x_p}})
\end{equation}
Rearranging, we get:
\begin{equation}
\frac{a\mathbf{\tilde{m_p}}Gm^2d^2\mathbf{\tilde{x_p}}}{a^3d{\tilde{t}}^2}=\frac{Gm^2\mathbf{\tilde{m_p}}\mathbf{\tilde{m_s}}a}{|a\mathbf{\tilde{x_s}}-a\mathbf{\tilde{x_p}}|^3}(\mathbf{\tilde{x_s}}-\mathbf{\tilde{x_p}})
\end{equation}
The $\frac{a\mathbf{\tilde{m_p}}Gm^2}{a^3}$ cancel from both sides leaving:
\begin{equation}
\frac{d^2\mathbf{\tilde{x_p}}}{d\tilde{t}^2}=\frac{\mathbf{\tilde{m_s}}a}{|\mathbf{\tilde{x_s}}-\mathbf{\tilde{x_p}}|^3}(\mathbf{\tilde{x_s}}-\mathbf{\tilde{x_p}})
\end{equation}

The primary star is positioned on the negative side of the x axis, hence $m_scos(\pi)$, whilst the secondary star is on the positive side, hence $m_pcos(0)$. They both sit on the x axis initially so $y=0$. Both of their velocities are zero in the x direction i.e. $-m_ssin(\pi)$ and $-m_psin(0)$ respectively. The stars initial velocities are the time derivatives of the initial positions. The primary star has negative velocity i.e. downwards given by $m_scos(\pi)$ whilst the secondary star has a positive velocity: $m_pcos(0)$, so the orbit is anti-clockwise. 
The initial conditions for this problem are given below. These are assuming an initial phase of $\phi=0$.
\begin{equation}
\mathbf{r_p}=\left(m_scos\left(\pi\right),m_ssin\left(\pi\right)\right)
\end{equation}
\begin{equation}
\mathbf{r_s}=\left(m_pcos\left(0\right),m_psin\left(0\right)\right)
\end{equation}
\begin{equation}
\mathbf{\dot{r_p}}=\left(-m_ssin\left(\pi\right),m_scos\left(\pi\right)\right)
\end{equation}
\begin{equation}
\mathbf{\dot{r_s}}=\left(-m_psin\left(0\right),m_pcos\left(0\right)\right)
\end{equation}

The orbital period is given by:
\begin{equation}
T=2\pi\sqrt{\frac{a^3}{G\left(M_1+M_2\right)}}
\end{equation}
and so in dimensionless units of $\sqrt{\left(a^3/Gm\right)}$ the period is simply $2\pi$.

\subsection{Parts b + c}
Assuming that $\frac{m_p}{m_s}=3$, then $m_p=0.75$ and $m_s=0.25$. I set the final time to be $6\pi$ to allow the stars to orbit three times. In Octave the intial conditions were set as follows:
\begin{lstlisting}
mp    =  .75 ;  % primary mass
ms    =  .25 ;  % secondary mass
x(1)  = ms*cos(pi);  % primary x
x(2)  = ms*sin(pi);  % primary y
x(3)  = -ms*sin(pi);  % primary vx
x(4)  = ms*cos(pi);  % primary vy
x(5)  =  mp*cos(0);  % secondary x
x(6)  =  mp*sin(0);  % secondary y
x(7)  =  -mp*sin(0);  % secondary vx
x(8)  = mp*cos(0);  % secondary vy
tmax  =  6*pi;  % final time
\end{lstlisting}
The full source code listing for Problem 2 is in Appendix 2.

The plot of the two stars is shown in Fig. \ref{fig:problem2_1}. As can be seen the orbits are circular with the more massive primary star (in blue) orbiting closer to the centre of mass. The less massive secondary star (in red) lies much further out. 
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem2/problem2_1}
\caption{Plot of circular binary system}
\label{fig:problem2_1}
\end{figure}

\subsection{Part d}
The plot of energy in the system is shown in Fig. \ref{fig:problem2_2}. The plot shows that the derived energy in the system (blue circles) and the exact energy (red line) match. The total energy in the system is constant and negative. The exact energy in the system was calculated using $E_{exact}=\frac{-m_pm_s}{2}$. This is derived as below, for total energy in orbit (potential and kinetic):
\begin{equation}
E=-\frac{Gm_1m_2}{r}+\frac{1}{2}mv^2
\end{equation}
Eliminate v by equating it with gravity:
\begin{equation}
E=-\frac{Gm_1m_2}{r}+\frac{1}{2}G\frac{m_1m_2}{r}
\end{equation}
Kinetic energy therefore is -1/2 of potential energy, so this simplifies to:
\begin{equation}
E=-G\frac{m_1m_2}{2r}
\end{equation}
$E_{exact}=\frac{-m_pm_s}{2}$ is in dimensionless units of $Gm^2/a$.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem2/problem2_2}
\caption{Plot of energy in binary system}
\label{fig:problem2_2}
\end{figure}

\newpage
\section{Problem 3: Hypervelocity Stars}
\subsection{Part a}
In order to plot the stellar orbits in the black hole rest frame it is first necessary to set the initial conditions. As the binary system is moving in a parabola the equations of motion for the centre of mass of parabola are:
\begin{equation}
X_{cm}=(Rcos(f),Rsin(f)),  \qquad  R=\frac{2R_p}{1+cos(f)}     
\end{equation}
This leads to the following equation for the initial angle given that the initial radius is $10R_{tidal}$ and $D=R_p/R_{tidal}=3$:
\begin{equation}
f_0=-cos^{-1}\left(-1+\frac{D}{5}\right)
\end{equation}
because $\frac{2R_p}{10R_{tidal}}=1+cos(f)$.
This was encoded into Octave as:
\begin{lstlisting}
f=-acos(-1+D/5);
\end{lstlisting}
Initial time was set according to the following equation:
\begin{equation}
t_0=\frac{\sqrt{2}}{3}D^{3/2}tan(f_0/2)(3+tan^2(f_0/2))
\end{equation}
In Octave this was set as follows:
\begin{lstlisting}
t = (sqrt(2)/3)*(D^(3/2))*tan(f/2)*(3+(tan(f/2))^2);
\end{lstlisting}
The initial radius is equal to $10R_{tidal}$ and $R_{tidal}=(M/m)^{1/3}a$ so in units of a, R was calculated as follows in Octave:
\begin{lstlisting}
Rt=mb^(1/3);
R=Rt*10;
\end{lstlisting}
The initial positions of the stars relative to the black hole are given by the position of the centre of mass plus the positions of the stars as they orbit each other in circles. The initial position of the centre of mass was calculated as follows:
\begin{lstlisting}
Xx=R*cos(f);
Xy=R*sin(f);
\end{lstlisting}
The initial positions of the stars were calculated in a similar fashion to Problem 2, however a phase of $pi/2$ was used.
In order to calculate the initial velocities of the stars we first need the initial velocity of the centre of mass. This was achieved by taking the derivatives of equation 13 and by using the deriviative of the true anomaly, the angle f:
\begin{equation}
\dot{f}=\frac{\sqrt{2}}{4}D^{-3/2}(1+cos(f))^2
\end{equation}
The above equation is derived from:
\begin{equation}
\frac{df}{dt}=\sqrt{\frac{GM}{R_p^3}}\frac{\sqrt{2}}{4}(1+cos(f))^2
\end{equation}
Here the $\frac{GM}{R_p^3}=D^{-3/2}$ in dimensionless units. We note that $M=R_{tidal}^3$. So therefore it follows that:
\begin{equation}
\sqrt{\frac{GM}{R_p^3}}=\sqrt{\frac{GR_{tidal}^3}{R_p^3}}
\end{equation}
and so, dropping the G for dimensionless units, we have:
\begin{equation}
\sqrt{\frac{R_{tidal}^3}{r_p^3}}=\left(\frac{R_p}{R_{tidal}}\right)^{-3/2}=D^{-3/2}
\end{equation}

The derivative of equation 13 is as follows:
\begin{equation}
\dot{R}=\frac{2R_p\dot{f}sin(f)}{(1+cos(f))^2}
\end{equation}
which gives:
\begin{equation}
\dot{R}=\frac{2R_p\frac{\sqrt{2}}{4}D^{-3/2}sin(f)(1+cos(f))^2}{(1+cos(f))^2}
\end{equation}
This simplies immediately to:
\begin{equation}
\dot{R}=\frac{R_pD^{-3/2}sin(f)}{\sqrt{2}}
\end{equation}
This can be rewritten as:
\begin{equation}
\dot{R}=\frac{R_psin(f)}{D\sqrt{2D}}
\end{equation}
As $D=\frac{R_p}{R_{tidal}}$ and $R_{tidal}=M^{1/3}$ we finally get:
\begin{equation}
\dot{R}=\frac{M^{1/3}sin(f)}{\sqrt{2D}}
\end{equation}

Substituting $\dot{R}$ and $\dot{f}$ into equation 13 and applying the product rule gives us the following equations for $\dot{X_x}$ and $\dot{X_y}$:
\begin{equation}
\dot{X_x}=\dot{R}cos(f)-R\dot{f}sin(f)
\end{equation}
\begin{equation}
\dot{X_y}=R\dot{f}cos(f)+sin(f)\dot{R}
\end{equation}
This was encoded in to Octave as follows:
\begin{lstlisting}
rdot=(mb^(1/3)/(sqrt(2*D)))*sin(f);
fdot=(sqrt(2)/4)*(D^(-3/2))*(1+cos(f))^2;
xdotx=(rdot*cos(f))-(R*fdot*sin(f));
xdoty=(R*fdot*cos(f))+(sin(f)*rdot);
\end{lstlisting}
Finally the initial conditions are ready to be set as follows:
\begin{lstlisting}
x(1) = Xx+(ms*cos(pi/2+pi)); % x : primary star
x(2) = Xy+(ms*sin(pi/2+pi)); % y
x(3) = xdotx+(-ms*sin(pi/2+pi)); % dx/dt
x(4) = xdoty+(ms*cos(pi/2+pi)); % dy/dt
x(5) = Xx+(mp*cos(pi/2)); % x : secondary star
x(6) = Xy+(mp*sin(pi/2)); % y
x(7) = xdotx+(-mp*sin(pi/2)); % dx/dt
x(8) = xdoty+(mp*cos(pi/2)); % dy/dt
\end{lstlisting}

The full source code listing for Problem 3 is in Appendix 3.

Fig. \ref{fig:problem3_1} shows the plot of the stars as the approach the black hole in a parabola. With a penetration factor of $D=\frac{R_p}{R_{tidal}}=3$ they can be seen to be orbitting each other and successfully maintain their orbits as they pass the black hole at periastron.
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem3/problem3_1}
\caption{Plot of binary system in BH rest frame where D=3}
\label{fig:problem3_1}
\end{figure}

\subsection{Part b}
Fig. \ref{fig:problem3_2} shows a plot of the same data as in Problem 1 except the plot is in the comoving frame of the secondary star. This was achieved by subtracting the position of the primary star from the position of the secondary star. The was coded into Octave as follows:
\begin{lstlisting}
plot(xs-xp, ys-yp, 'r:')
\end{lstlisting}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem3/problem3_2}
\caption{Plot of primary star in comoving frame of secondary star D=3}
\label{fig:problem3_2}
\end{figure}
The binary system is clearly not destabilised by the black hole.
\subsection{Part c}
Fig. \ref{fig:problem3_3} is a plot of the energy of the binary system. It shows the energy is oscillating from positive to negative between the two stars as they orbit each other. The amplitude of the plot can be seen to be increasing as the stars approach the black hole reaching a maximum at periastron before starting to decrease as they move away from the influence of the black hole. The system has a total energy of zero and the energy of the one star is always the negative of the other.
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem3/problem3_3}
\caption{Plot of energy in binary system D=3}
\label{fig:problem3_3}
\end{figure}

\subsection{Part d}
The code was then re-run but this time with $R_p=0.1R_{tidal}$.
Fig. \ref{fig:problem3_4} is a plot again in the rest frame of the black hole.
The binary system is clearly disrupted by the black hole. The two stars are orbitting each until the closest approach to the black hole where they separate.
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem3/problem3_4}
\caption{Plot of binary system in BH rest frame where D=0.1}
\label{fig:problem3_4}
\end{figure}

A plot of the secondary star in the comoving frame of the primary star, Fig. \ref{fig:problem3_5}, shows the secondary star to be initially in orbit with the primary star but then leaves the binary system on a separate trajectory.
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem3/problem3_5}
\caption{Plot of primary star in comoving frame of secondary star D=0.1}
\label{fig:problem3_5}
\end{figure}

A plot of the energy of the system, Fig. \ref{fig:problem3_6}, shows that the energy is initially oscillating as in Part c but after the closest approach to the black hole the energy no longer oscillates between the two stars after the system is disrupted. The primary star (blue) ends up with a high negative energy whilst the secondary star (red) has a high positive energy.
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem3/problem3_6}
\caption{Plot of energy in binary system D=0.1}
\label{fig:problem3_6}
\end{figure}

\subsection{Part e}
The velocity of the ejected star (secondary star in this case) can be calculated by using $E_k=1/2mv^2$ so
\begin{equation}
v=\sqrt{2E/M}
\end{equation}
From Fig. \ref{fig:problem3_6}, and from the output file, the energy of the ejected star tends towards 19 as r approaches $\infty$. E here is in units of $Gm^2/a$ so, in kms units, $Gm^2/a=5.4\times10^{40}$ and so $E=10^{41}J$. From equation 27 we get 1,017,849m/s, and so the ejected star is travelling at approximately 1000km/s at $r=\infty$.
\subsection{Part f}
Throughout this project I had assumed that the lower mass star would always be ejected from the system, so I chose to investigate this further to see if my assumptions were correct.
I ran the above simulation 8 times whilst varying the initial phase of the binary system i.e. pi/4, pi/2, 3pi/4, pi, 5pi/4, 3pi/2, 7pi/4 and 2pi.
The results were surprsing because in exactly half the runs the lower mass star was ejected and half the larger mass star. With an initial phase of pi/4 to pi the lower mass star was ejected, whilst from 5pi/4 to 2pi the larger mass star was ejected, see Fig. \ref{fig:problem3_7}. This suggests that this process is independent of the mass of the star.
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./problem3/problem3_7}
\caption{Plot of binary system in BH rest frame where D=0.1 and Phase=2Pi}
\label{fig:problem3_7}
\end{figure}
This may be possibly due to the energy in the system. The total energy in the system is zero so the energy of the one star is always equal to the negative of the other. The star which is ejected may be the one that is in the phase with positive energy at the time of closest approach to the black hole.

Astronomical observations could confirm whether indeed we do detect similar numbers of high mass hypervelocity stars as we do low mass hypervelocity stars.
\newpage
\section{Appendix 1 - Problem 1B code}
\lstinputlisting[language=Octave]{Part2.m}
\newpage
\section{Appendix 2 - Problem 2 code}
\lstinputlisting[language=Octave]{./problem2/binary.m}
\lstinputlisting[language=Octave]{./problem2/f.m}
\lstinputlisting[language=Octave]{./problem2/energy.m}
\lstinputlisting[language=Octave]{./problem2/RK4.m}
\lstinputlisting[language=Octave]{./problem2/orbitplot.m}
\lstinputlisting[language=Octave]{./problem2/energyplot.m}
\newpage
\section{Appendix 3 - Problem 3 code}
\lstinputlisting[language=Octave]{./problem3/initialc.m}
\lstinputlisting[language=Octave]{./problem3/f.m}
\lstinputlisting[language=Octave]{./problem3/HVS.m}
\lstinputlisting[language=Octave]{./problem3/script4.m}
\lstinputlisting[language=Octave]{./problem3/RK4.m}
\lstinputlisting[language=Octave]{./problem3/energy.m}
\end{document}

### 2016-08-23 Project3/example-codes_project3/problem3/Project3HK.tex
First checkin

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{Computational Astrophysics Project 2 - Investigation of Stellar Orbits within Barred Spiral Galaxies}
\begin{document}
\maketitle
\section{Part 1}
The Fortran program orbit.f was used to solve the 2nd order differential equation of motion: 
\begin{equation} \label{motion}
   \frac{d^2\mathbf{r}(t)}{dt^2}=-\nabla\Phi\left(\mathbf{r}\right)-2\mathbf{\Omega_B}\times\mathbf{v}+\Omega_B^2\mathbf{r}
\end{equation}
where the gravitational potential of an elongated bar is represented by:
\begin{equation} \label{grav}
\Phi\left(x,y\right)=\frac{v_0^2}{2}ln\left(a^2+x^2+\frac{y^2}{q^2}\right)
\end{equation}
In order to solve a 2nd order equation it is necessary to first split it into two 1st order differential equations and solve for the x and y components:
\begin{equation} \label{dxdt}
\frac{dx}{dt}=v_x
\end{equation}
\begin{equation} \label{dvxdt}
\frac{dv_x}{dt}=\frac{-v_0^2x}{a^2+\frac{y^2}{q^2}+x^2}+2\Omega_B\frac{dy}{dt}+\Omega_B^2x
\end{equation}
\begin{equation} \label{dydt}
\frac{dy}{dt}=v_y
\end{equation}
\begin{equation} \label{dvydt}
\frac{dv_y}{dt}=\frac{-v_0^2y}{a^2q^2+q^2x^2+y^2}-2\Omega_B\frac{dx}{dt}+\Omega_B^2y
\end{equation}
\newpage
Equation \ref{dvxdt} was encoded into Fortran as:
\begin{lstlisting}
dydx(2) = (-v0**2*y(1))/(a**2+(y(3)**2/q**2)+y(1)**2)
         +2*om*y(4)+om**2*y(1)
\end{lstlisting}
Equation \ref{dvydt} was encoded as:
\begin{lstlisting}
 dydx(4) = (-v0**2*y(3))/(a**2*q**2+q**2*y(1)**2+y(3)**2)
         -2*om*y(2)+om**2*y(3) 
\end{lstlisting}

For the first run the initial conditions were set as $x=0$, $y=0.087$, $v_y=0$ and $v_x=-1.68$. The final time was set as 5 initially.
The accuracy of the integration was controlled by monitoring the Jacobi integral
which is conserved as a star moves in a rotating bar. An initial value for $\epsilon$ was set at $10^{-3}$. The plot of the Jacobi integral with time is shown in Fig. \ref{fig:jacobi1}. As can be seen it ranges from -.7747 to -.7738.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.68x5/jacobi1}
\caption{Plot of Jacobi integral with $\epsilon=10^{-3}$}
\label{fig:jacobi1}
\end{figure}

To improve accuracy the program was re-run with $\epsilon=10^{-5}$. Fig. \ref{fig:jacobi2} now shows a Jacobi integral values of -.774615 to -.774595, a substantial improvement.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.68x5/jacobi2}
\caption{Plot of Jacobi integral with $\epsilon=10^{-5}$}
\label{fig:jacobi2}
\end{figure}

Fig. \ref{fig:jacobi3} is a plot of the Jacobi integral this time where $\epsilon=10^{-7}$. This plot shows that most values of the integral hover around the -.774616 mark.
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.68x5/jacobi3}
\caption{Plot of Jacobi integral with $\epsilon=10^{-7}$}
\label{fig:jacobi3}
\end{figure}

An attempt to further improve accuracy with values of $\epsilon=10^{-8}$ or $\epsilon=10^{-9}$ show that the accuracy actually starts to deteriorate - see Fig. \ref{fig:jacobi4}. Therefore a value of $\epsilon=10^{-7}$ was used for all subsequent runs of the program.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.68x5/jacobi4}
\caption{Plot of Jacobi integral with $\epsilon=10^{-8}$}
\label{fig:jacobi4}
\end{figure}

Orbit trajectories were then calculated with a final time of 20. The first plot where $v_x=-1.68$ is shown in Fig. \ref{fig:orbit1}. The orbit is closed and has an elliptical shape. The star can be seen to be rotating around a bar-like shape. The plot also shows that the star does not orbit on a constant path, each subsequent orbit follows a slightly different path around the bar whilst maintaining the general elliptical shape. The path variations are greatest along the short axis.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.68x20/Graph2}
\caption{Plot of trajectory where $v_x=-1.68$}
\label{fig:orbit1}
\end{figure}

Figs. \ref{fig:orbit2} and \ref{fig:orbit3} show the same plots but with $v_x=-1.64$ and $v_x=-1.635$ respectively. The orbit here retains its essentially elliptical shape but the star now seems to orbit along a more constant path. This can be seen even more clearly in Fig. \ref{fig:orbit3} where subsequent orbits keep to the same general path around a bar-like shape. The orbit is also symmetrical about both the x and y axis.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.64x20/Graph2}
\caption{Plot of trajectory where $v_x=-1.64$}
\label{fig:orbit2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.635x20/Graph2}
\caption{Plot of trajectory where $v_x=-1.635$}
\label{fig:orbit3}
\end{figure}

Figs. \ref{fig:orbit4} to \ref{fig:orbit9} show the trajectory plots where the value of $v_x$ has values of -1.63, -1.58, -1.15, -.6, -.524 and -.4 respectivetly. They are again all closed orbits. As $v_x$ increases the orbits can be seen to gradually become more erratic. In Fig. \ref{fig:orbit5} the orbit begins to lose its elliptical shape and become slightly rounder but again showing the radius of the orbit varying widely along the shorter axis. Fig. \ref{fig:orbit6}, where $v_x=-1.15$, the orbits, although still surrounding a bar-like feature, no longer follow constant paths and the radial distance is not constant at any points. Further increases of $v_x$, as in Fig. \ref{fig:orbit7}, creates orbits which no longer orbit a bar-like shape, and takes on an almost torus-like shape with the star seeming to make small epicycles within the torus as it orbits the galaxy. When $v_x$ has been increased to -.524, as in Fig. \ref{fig:orbit8}, a more stable orbital structure appears, where again the star follows a fairly constant elliptical orbit. A final run of $v_x=-.4$ reveals an elliptical orbit although, similarly to Fig. \ref{fig:orbit7}, the star is again making epicycles forming a torus-like shape as it orbits the galaxy.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.63x20/Graph2}
\caption{Plot of trajectory where $v_x=-1.63$}
\label{fig:orbit4}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.58x20/Graph2}
\caption{Plot of trajectory where $v_x=-1.58$}
\label{fig:orbit5}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.15x20/Graph2}
\caption{Plot of trajectory where $v_x=-1.15$}
\label{fig:orbit6}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-.6x20/Graph2}
\caption{Plot of trajectory where $v_x=-0.6$}
\label{fig:orbit7}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-.524x20/Graph2}
\caption{Plot of trajectory where $v_x=-0.524$}
\label{fig:orbit8}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-.4x20/Graph2}
\caption{Plot of trajectory where $v_x=-0.4$}
\label{fig:orbit9}
\end{figure}

\newpage
\section{Part 2}
Fast Fourier Transforms were then carried out on the trajectories calculated above where $v_x=-1.635$ and the subsequent one-sided power spectrum density plotted. Fig. \ref{fig:psd1} shows the linear PSD, whilst Fig. \ref{fig:psd2} shows it with a logarithmic scale.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./F-1.635/Graph1}
\caption{Linear plot of PSD for $v_x=-1.635$}
\label{fig:psd1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./F-1.635/Graph2}
\caption{Logarithmic plot of PSD for $v_x=-1.635$}
\label{fig:psd2}
\end{figure}

Fig. \ref{fig:psd1} shows that 85.5\% of the PSD is oscillating at a frequency at 4.7 (to 1 dp). There are also much smaller peaks at 9.4 and 14.1. This is consistent with harmonics of the first peak. The peak at 4.7 is the fundamental frequency. The harmonics are repeated at integer multiples of the fundamental frequency of 4.7 e.g. 9.4, 14.1, 18.8 etc. This is illustrated far better in the logarithmic plot in Fig. \ref{fig:psd2} where again the fundamental frequency peak is at 4.7 with clear harmonics showing at regular intervals of 4.7 e.g. 9.4, 14.1, 18.8 etc. The power can be seen to be decreasing as the frequency increases. This is consistent with a radius that is oscillating at a constant frequency i.e. the initial fundamental frequency. In order to investigate this further I modified orbit.f to produce an additional column in its output: radius. This was done to see how the radius varied within the time domain. Fig. \ref{fig:radius1} is a plot of radius against time and clearly shows the radius oscillating at a constant frequency. The period is the reciprocal of the fundamental frequency so $\frac{2\pi}{4.7}=1.3$. This is consistent with Fig. \ref{fig:radius1}. After one period the star will have completed half an orbit because the orbit is elliptical and so symmetrical about the x and y axis i.e. the radius will have returned to its original value and the star will lie on the opposite side of its orbit.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.635x20/Graph4}
\caption{Plot of radius/time for $v_x=-1.635$}
\label{fig:radius1}
\end{figure}

Fig. \ref{fig:psd2} also shows that there is little or no aliasing in the results because there are no peaks at the Nyquist frequency: $\frac{\pi}{.02}=157.08$. Frequencies tending to zero at the Nyquist frequency are an indication that aliasing is unlikely. There are also no peaks at very low frequencies indicating a good choice of sampling interval.

The PSD was then plotted for $v_x=-1.64$ as shown in Figs. \ref{fig:psd3} and \ref{fig:psd4}.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./F-1.64/Graph1}
\caption{Linear plot of PSD for $v_x=-1.64$}
\label{fig:psd3}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./F-1.64/Graph2}
\caption{Logarithmic plot of PSD for $v_x=-1.64$}
\label{fig:psd4}
\end{figure}

In Fig. \ref{fig:psd3} the power of the main peak is lower than for when $v_x=-1.635$ at 71.4\% but is still clearly present at almost the same fundemantal frequency (4.6). The harmonics are again present at similar values as the previous plot. In Fig. \ref{fig:psd4} the peaks are less well defined than in Fig. \ref{fig:psd2}. There appears to be other peaks appearing that are not harmonics of the one fundamental frequency, for example there are independent peaks with powers of 11.1\%, 3.5\%, 1.9\% and 1.1\%. There are no peaks at low frequencies and the data tends to zero at the Nyquist frequency so aliasing again doesn't seem to be an issue. A plot of radius with time, Fig. \ref{fig:radius2}, shows that the radius is now varying by small amounts on each cycle - it is no longer constant.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.64x20/Graph4}
\caption{Plot of radius/time for $v_x=-1.64$}
\label{fig:radius2}
\end{figure}

The PSD was then plotted for $v_x=-1.68$ as shown in Figs. \ref{fig:psd5} and \ref{fig:psd6}.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./F-1.68/Graph1}
\caption{Linear plot of PSD for $v_x=-1.68$}
\label{fig:psd5}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./F-1.68/Graph2}
\caption{Logarithmic plot of PSD for $v_x=-1.68$}
\label{fig:psd6}
\end{figure}

Fig. \ref{fig:psd5} has a main peak at about 47.9\% of the power. Referring back to Fig. \ref{fig:orbit1} shows that this would represent the star when it is orbitting along the long axis of the ellipse i.e. the outer edges. The star is almost following a constant path here. There is also another independent frequency with a power of 16.5\%, and at least 16 peaks with powers of \textgreater=.1\%. These other peaks represent the various orbital paths around the galaxy.
In Fig. \ref{fig:psd6} many unresolved peaks at low frequencies become apparent. There is also significant aliasing as the radius is no longer oscillating at a constant frequency, see Fig. \ref{fig:radius3}, and so the sampling interval is no longer representative of the true orbital path.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.68x20/Graph4}
\caption{Plot of radius/time for $v_x=-1.68$}
\label{fig:radius3}
\end{figure}

The PSD was then plotted for $v_x=-1.15$ as shown in Figs. \ref{fig:psd7} and \ref{fig:psd8}. There is a major peak at a frequency of 11 with 38.8\% power with another smaller peak with a power of 10.8\%. As in the previous run, there are many peaks with powers of \textgreater=.1\%. The linear plot doesn't show much aliasing but when viewed on the logarithmic plot many low amplitude aliased peaks are revealed which spread across the whole sample up to the Nyquist frequency. The is due to the fact that the radius is oscillating at frequencies greater than the Nyquist frequency and so are appearing in the sample. A plot of radius against time, Fig. \ref{fig:radius4}, reveals that the radius oscillates erratically and so the samples no longer represent the true path which results in aliasing.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./F-1.15/Graph1}
\caption{Linear plot of PSD for $v_x=-1.15$}
\label{fig:psd7}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./F-1.15/Graph2}
\caption{Logarithmic plot of PSD for $v_x=-1.15$}
\label{fig:psd8}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./Vx-1.15x20/Graph4}
\caption{Plot of radius/time for $v_x=-1.15$}
\label{fig:radius4}
\end{figure}

\end{document}

### 2016-08-23 DoubleBars/DoubleBars.f

@@ -1,0 +1,216 @@
+      program DoubleBars
+      implicit none
+      integer NVAR,NMAX,KMAXX
+      parameter(NVAR=8,NMAX=50,KMAXX=10000) !same values as in odeint
+      integer kmax,kount,nok,nbad,i,j
+      REAL ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),dxsav,x1,x2,ej,om,v0,a,q
+     #,r(KMAXX),m1,m2,m3,m4
+      external derivs,rkqs
+      common /path/ kmax,kount,dxsav,xp,yp
+      common /params/ om,v0,a,q
+      x1=0.
+      x2=6*4*ATAN(1.)
+
+      m1 = 0.75
+      m2 = 0.25
+      m3 = 1.
+      m4 = 1.
+
+      ys(1) = -1*m2             ! m1,x
+      ys(2) = 0                 ! m1,vx
+      ys(3) = 0                 ! m1,y
+      ys(4) = 1*m2              ! m1,vy
+
+      ys(5) = 1*m1              ! m2,x
+      ys(6) = 0                 ! m2,vx
+      ys(7) = 0                 ! m2,y
+      ys(8) = -1*m1             ! m2,vy
+c
+      call odeint(ys,nvar,x1,x2,1.e-7,1.e-8,0.0,nok,nbad,derivs,rkqs)
+c
+      do 10 i=1,kmax
+         ej = 0.5*(yp(2,i)**2+yp(4,i)**2)-0.5*om*(yp(1,i)**2+yp(3,i)**2)
+     #      + 0.5*v0**2*alog(a**2+yp(1,i)**2+(yp(3,i)/q)**2)
+         r(i)=sqrt(yp(1,i)**2+yp(3,i)**2)
+         write(9,33) xp(i), (yp(j,i),j=1,NVAR), ej,i,r(i)
+         if(xp(i).eq.x2) goto 99
+ 10   continue
+ 99   continue
+ 33   format(9f10.4,f13.8,i5,f10.4)
+      end
+c
+      subroutine derivs(x,y,dydx)
+      implicit none
+      integer NVAR
+      PARAMETER(NVAR=8)         !same values as in main program
+      real x,y(NVAR),dydx(NVAR),om,v0,a,q
+      real dist1,dist2,dist3, dist4, dist5, dist6
+      real r10,m1,m2
+      COMMON /params/ om,v0,a,q
[…]

### 2016-08-23 Project1/LKbiasMSc.f

@@ -1,0 +1,305 @@
+C-----------------------------------------------------------------------
+C
+      program LKbias
+C
+
+      implicit double precision (a-h,o-z)
+
+C     ---   Declaration of variables   ---
+
+c      real ran  
+c      character*4 ngc
+c      character*72 fili
+c      double precision lg
+
+
+C     ---   Parametri   ---
+
+c      parameter (nstar=500000)  ! number of objects
+c      parameter (exp=2.0e0)     ! exponent of distance distribution
+c      parameter (sigma=5.0d-5)  ! parallax error (in arscsec)
+c      parameter (sigmam=0.2d0)  ! magnitude spread 
+       
+C     ---   Commons   ---
+
+      common /reed/ iseed1,iseed2
+ 
+      open(unit=2, file='LK.dat')  ! output file
+
+      write(*,*) 'Output file LK.dat'
+
+      write(*,*) '-----------------------------------------------'
+      write(*,*) 'Select number of stars (1-500000)'
+      read(*,*) nstar
+      write(*,*) 'Select space distribution (exponent n of R^n' 
+      write(*,*) 'distribution with R=distance between 1 pc and 8 Kpc)'
+      write(*,*) 'n=3 --> constant spatial density of stars' 
+      write(*,*) 'n=2 --> spatial density of stars decreasing as R^(-1)' 
+      write(*,*) 'n=1 --> spatial density of stars decreasing as R^(-2)' 
+      read(*,*) exp    
+      write(*,*) 'Select type parallax error'
+      write(*,*) 'GAIA-like error law =0 - Constant value=1'
+      read(*,*)  indicepi
+      if(indicepi.gt.0) then
+      write(*,*) 'Select constant parallax error (arcsec)'
+      read(*,*) sigma
+      end if
+      write(*,*) 'Select true mean absolute magnitude stars (mag)'
+      read (*,*) zmvnp
+      write(*,*) 'Select distribution true magnitudes'
+      write(*,*) 'Gaussian=0 - Uniform=1'
+      read(*,*) indice
+      write(*,*) 'Select 1sigma true dispersion of the star' 
+      write(*,*) 'absolute magnitude' 
+      read(*,*) sigmam
+      write(*,*) 'Select 1sigma photometric error (mag)' 
+      write(*,*) 'on the measured apparent magnitude'
+      read (*,*) photerr
[…]

### 2016-08-23 Project1/analysis1.f

@@ -1,0 +1,211 @@
+      program LKBiasAnalyse
+
+      implicit real*8 (a-h,o-z)
+
+      open(unit=2, file='LK.dat')
+      open(unit=3, file='LKRDist.dat',status='REPLACE')
+      open(unit=15, file='LKTrueDist.dat',status='REPLACE')
+      do 10 i=0, 12
+         read(2,*)
+ 10      continue
+         
+      rmean=0
+      count=0
+      total=0
+      temp=0
+      stdev=0
+      count1=0
+      count2=0
+      count3=0
+      count4=0
+c      count5=0
+c      count6=0
+      total1=0
+      total2=0
+      total3=0
+      total4=0
+c      total5=0
+c      total6=0
+      rmean1=0
+      rmean2=0
+      rmean3=0
+      rmean4=0
+c      rmean5=0
+c      rmean6=0
+      stdev1=0
+      stdev2=0
+      stdev3=0
+      stdev4=0
+c      stdev5=0
+c      stdev6=0
+         
+      do 20 i=0, 400001
+         read(2,*,IOSTAT=io) tpara, tdist, appmag, obpara, paraerr, 
+     +rdist, tmag, rmag, errphoto
+      if(io<0) then
+         exit
+      elseif(rmag.lt.99) then
+         if(rdist.lt.4000) then
+            count=count+1
+            total=total+rmag
+            write(3,*) tpara, tdist, appmag, obpara, paraerr,
+     +           rdist, tmag, rmag, errphoto
+         
+            if(tdist.lt.4000) then
+               write(15,*) tpara, tdist, appmag, obpara, paraerr,
+     +              rdist, tmag, rmag, errphoto  
+            endif
+        endif
+      endif
+ 20   continue   
+      close(2)
+      close(3)
+      close(15)
+      
+      open(unit=4, file='LKRDist.dat')
+      open(unit=5, file='Bin1000.dat',status='REPLACE')
+      open(unit=7, file='Bin2000.dat',status='REPLACE')
+      open(unit=8, file='Bin3000.dat',status='REPLACE')
+      open(unit=9, file='Bin4000.dat',status='REPLACE')
+c      open(unit=16, file='Bin5000.dat',status='REPLACE')  
+c      open(unit=17, file='Bin6000.dat',status='REPLACE')    
+     
+      rmean=total/count
+      do 30 i=0, count-1
[…]

### 2016-08-23 Project2/fourier.f

@@ -1,0 +1,344 @@
+      program follow
+      implicit none
+      integer NVAR,NMAX,KMAXX
+      PARAMETER(NVAR=4,NMAX=50,KMAXX=8192) !same values as in odeint
+      INTEGER kmax,kount,nok,nbad,i,j,k
+      REAL y(NVAR),xp(KMAXX),yp(NMAX,KMAXX),dxsav,x1,x2,ej,om,v0,a,q
+      REAL step
+      REAL rw(KMAXX),prw(kmaxx),prwtot,rav
+      EXTERNAL derivs,rkqs
+      COMMON /path/ kmax,kount,dxsav,xp,yp
+      COMMON /params/ om,v0,a,q
+c
+c     parameters of the potential and the step
+c
+      om = 1.
+      v0 = 1.
+      a=0.03
+      q=0.8
+      step=2.e-2
+c
+c     initial conditions
+c
+      x1 = 0.
+      x2 = step
+      y(1) = 0.                 !x
+      y(2) =  -1.15              !vx
+      y(3) = .087                 !y
+      y(4) = 0.                 !vy
+c
+c     orbit calculation, including radius and radial velocity
+c
+      rav   = 0.
+c
+      do 5  i=1,kmaxx
+c
+         call odeint(y,nvar,x1,x2,1.e-7,1.e-8,0.0,nok,nbad,derivs,rkqs)
+c
+         ej = 0.5*(y(2)**2+y(4)**2)-0.5*om*(y(1)**2+y(3)**2)
+     #      + 0.5*v0**2*alog(a**2+y(1)**2+(y(3)/q)**2)
+         rw  (i) = sqrt(y(1)*y(1)+y(3)*y(3))
+         rav = rav + rw  (i)
+         x1=x2
+         x2=float(i+1)*step
+ 5    continue
+c
+      rav = rav/float(kmaxx) 
+      write(*,*) 'average radius: ', rav
+c
+      do 6 k=1,kmaxx
+         rw(k) = rw(k) - rav
+ 6    continue
+c     
+      write(*,*) 'orbit calculated'
+c
+c     FFT and PSD calculation
+c
+      call realft(rw,kmaxx,1)
+c
+      prwtot = 0.
+      do 10 k=1,kmaxx/2
+         prw(k) = 2.*(rw(2*k-1)*rw(2*k-1) + rw(2*k)*rw(2*k))
[…]

### 2016-08-23 Project3/example-codes_project3/Part1.m

@@ -1,0 +1,39 @@
+% problem1B.m
+% Time evolution of temperature distribution u: spherical system
+% Define A, B, C, unew(1) and unew(N)
+
+% When you write code, it is a good practice to add comments that 
+% describe the code. Comments allow others to understand your code, 
+% and can refresh your memory when you return to it later. 
+% Add comments to Octave code using the percent (%) symbol.
+% Comment lines can appear anywhere in a program file, and you can 
+% append comments to the end of a line of code. see section 26, Octave tutorial
+
+mu    = 0.2;       % mu=dt/dx^2
+N     = 50;        % grid points
+steps = 2400;      % total time-steps - tau=0.2
+u     = ones(N,1); % initial condition: u = 1 anywhere
+u(N)  = 0;         % boundary condition at the surface
+approx=[];         % empty array
+for n=1:steps      % time evolution     
+    approx(:,n)=u; % Distribution u at each step saved in matrix: approx
+                   % see section 5, Octave tutorial for the colon. 
+                   % Add a new column to approx matrix in each timestep.
+                   % k-th column of approx gives u at timestep k.
+    for j=2:N-1
+        A = mu;   % see Eq (15) and define A, B, C
+	B = 1-(2*mu*(1+(1/(j-1))));
+        C = mu*(1+(2/(j-1)));
+        unew(j)= A*u(j-1)+B*u(j)+C*u(j+1);
+    end      
+    unew(1) = 1; % boundary condition at the centre
+    unew(N) = 0; % boundary condition at the surface
+    u = unew;      % update the distribution
+end
+
+x=linspace(0,1,N);     % generate N points between 0 and 1. 
+plot(x,approx(:,[1:100:steps])) % plot temperture distributions
+                                % distributions at n=1, 101, 201, 301...
+                                % see section 5 and 6, Octave tutorial 
+xlabel('x','FontSize',20)
+ylabel('u','FontSize',20)

### 2016-08-23 Project3/example-codes_project3/Part2.m

@@ -1,0 +1,41 @@
+% problem1B.m
+% Time evolution of temperature distribution u: spherical system
+% Define A, B, C, unew(1) and unew(N)
+
+% When you write code, it is a good practice to add comments that 
+% describe the code. Comments allow others to understand your code, 
+% and can refresh your memory when you return to it later. 
+% Add comments to Octave code using the percent (%) symbol.
+% Comment lines can appear anywhere in a program file, and you can 
+% append comments to the end of a line of code. see section 26, Octave tutorial
+
+mu    = 0.2;       % mu=dt/dx^2
+N     = 50;        % grid points
+steps = 2400;      % total time-steps
+u     = ones(N,1); % initial condition: u = 1 anywhere
+u(N)  = 0;         % boundary condition at the surface
+approx=[];         % empty array
+for n=1:steps      % time evolution     
+    approx(:,n)=u; % Distribution u at each step saved in matrix: approx
+                   % see section 5, Octave tutorial for the colon. 
+                   % Add a new column to approx matrix in each timestep.
+                   % k-th column of approx gives u at timestep k.
+    for j=2:N-1
+        A = mu;   % see Eq (15) and define A, B, C
+	B = 1-(2*mu*(1+(1/(j-1))));
+        C = mu*(1+(2/(j-1)));
+        unew(j)= A*u(j-1)+B*u(j)+C*u(j+1);
+    end      
+    unew(1) = unew(2); % boundary condition at the centre
+    unew(N) = 0; % boundary condition at the surface
+    u = unew;      % update the distribution
+end
+u1 = approx(2,:);      % time evolution of temperature at the center j=1
+nc = min(find(u1<0.5)) % critical timestep: central temp becomes < 0.5. 
+
+x=linspace(0,1,N);     % generate N points between 0 and 1. 
+plot(x,approx(:,[1:100:steps])) % plot temperture distributions
+                                % distributions at n=1, 101, 201, 301...
+                                % see section 5 and 6, Octave tutorial 
+xlabel('x','FontSize',20)
+ylabel('u','FontSize',20)

### 2016-08-23 Project3/example-codes_project3/problem1Aa.m

@@ -1,0 +1,48 @@
+% example 1Aa: drawing a circle
+
+% When you write code, it is a good practice to add comments 
+% that describe the code. Comments allow others to understand 
+% your code, and can refresh your memory when you return to it 
+% later. Add comments to Octave code using the percent (%) symbol.
+% Anytime Octave encounters a percent sign (%), it treats the rest 
+% of the line as a comment.  section 26, Octave tutorial
+
+clear        % clear variables and functions from memory
+tic
+numsteps=2000;
+numparticles=100000;
+dx=0.05;
+binedges=0:dx:1;
+for j=1:numparticles
+    x(1)=0;
+    y(1)=0;
+    for i=2:numsteps  % Repeat statements a specific number of times,i.e. 100 times
+             % see section 27.1, Octave tutorial
+    theta = rand*2*pi; % theta= 2pi/100, 4pi/100, 6pi/100,...
+                        % pi: a built-in octave variable, 3.1415....
+                        % semicolon suppresses output from octave command
+                        % see section 17, Octave tutorial 
+    x(i)  = x(i-1)+0.01*cos(theta); % define i-th element of x vector. sec 5, tutorial 
+    y(i)  = y(i-1)+0.01*sin(theta); % define i-th element of y vector     
+    end
+    xfinal(j)=x(numsteps);
+    yfinal(j)=y(numsteps);
+    rfinal(j)=sqrt(x(numsteps)^2+y(numsteps)^2);
+end
+%subplot(2,1,1)
+%plot(xfinal,yfinal,"+")  % plots vector Y versus vector X. see section 17
+%axis equal % sets the aspect ratio so that equal tick mark
+           % increments on the x and y axis are equal in size.
+%xlabel('X','FontSize',20) % X-axis label
+%ylabel('y','FontSize',20) % Y-axis label
+counts=histc(rfinal,binedges);
+midpoints=binedges+dx/2;
+subplot(2,1,2)
+plot(midpoints,counts,'ro')
+xlim([0 1])                
+grid on                    
+xlabel('X','FontSize',20)
+ylabel('Counts','FontSize',20)
+set(gca, 'FontSize', 15)
+
+toc

### 2016-08-23 Project3/example-codes_project3/problem2/binary.m

@@ -1,0 +1,34 @@
+% this is the main code to estimate the binary evolution
+% two functions: energy.m and RK4.m are used in this code.
+% See section 24, Octave tutorial for Function m-files
+% Define mp, ms, x(1)-x(8) and tmax
+
+clear;clf          
+h      = 1.d-2;  % time-step size
+Ns     = 100  ;  % sampling
+%-----------------
+mp    =  .75 ;  % primary mass
+ms    =  .25 ;  % secondary mass
+x(1)  = ms*cos(pi);  % primary x
+x(2)  = ms*sin(pi);  % primary y
+x(3)  = -ms*sin(pi);  % primary vx
+x(4)  = ms*cos(pi);  % primary vy
+x(5)  =  mp*cos(0);  % secondary x
+x(6)  =  mp*sin(0);  % secondary y
+x(7)  =  -mp*sin(0);  % secondary vx
+x(8)  = mp*cos(0);  % secondary vy
+tmax  =  6*pi;  % final time
+%-----------------
+t      =  0  ;  % initial time
+tprint =  t;
+dtp    =  (tmax-t)/Ns;
+while t < tmax
+     if  t >= tprint
+         E = energy(x,mp,ms);       % Estimat binary energy
+         v = [t x(1) x(2) x(5) x(6) E];
+         save out v -ascii -append  % save the v vector in the file out           
+         tprint =  tprint + dtp;
+     end
+     x = RK4(h,t,x,mp,ms); % Runge-Kutta method 
+     t = t+h;
+end

### 2016-08-23 Project3/example-codes_project3/problem2/energyplot.m

@@ -1,0 +1,22 @@
+% plot the evolution of the binary energy (it should be a constant). 
+% Defin mp, ms, and tmax
+clear; clf;
+%----------------
+mp   = .75;
+ms   = .25;
+tmax = 6*pi;
+%----------------
+load out  % load the data file out
+t     = out(:,1); % time 
+E     = out(:,6); % energy
+exact = -mp*ms/2; % analytic estimate
+
+plot(t,E,'o') % numerical energy
+hold on 
+plot([0 tmax],[exact exact],'r') % exact line
+xlabel('t','FontSize',20)
+ylabel('E','FontSize',20)
+xlim([0 tmax])
+ylim([-0.2 0])
+set(gca, 'Fontsize', 20)
+

### 2016-08-23 Project3/example-codes_project3/problem2/orbitplot.m

@@ -1,0 +1,18 @@
+% plot the orbits of the two stars
+clear; clf;
+load out     % load the data file out
+xp=out(:,2); % the primary star
+yp=out(:,3);
+xs=out(:,4); % the secondary star
+ys=out(:,5);
+
+plot(xp,yp) % plot the orbit of the primary star
+hold on     % hold ON holds the current plot and all axis properties 
+            % so that subsequent graphing commands add to the existing 
+            % graph without resetting the color and linestyle.
+plot(xs,ys,'r:') % plot the orbit of the secondary star
+xlabel('x','FontSize',20)
+ylabel('y','FontSize',20)
+ylim([-0.8 0.8]) % plotting range from x=-0.8 to 0.8. 
+set(gca, 'Fontsize', 20)
+axis equal

### 2016-08-23 Project3/example-codes_project3/problem3/HVS.m

@@ -1,0 +1,25 @@
+% The disruption of a binary by a massive BH
+% the main code
+% define mb, mp, ms. These are dimensionless masses. 
+
+clear             % clear all variables and functions from memory
+D     = 0.1;        % penetration factor: Rp/Rtidal
+h     = 1.d-3;    % time-step size
+Ns    = 10000;    % sampling
+mb    = 1e6;     % BH mass in units of the binary mass m
+mp    = .75;     % primary star mass in units of m
+ms    = .25;     % secondary star mass in units of m
+[t, x] = initialc(D,mb,mp,ms); % Define the initial values of t and x. 
+tmax   = -t;
+tprint =  t;
+dtp    =  (tmax-t)/Ns;
+while t < tmax
+     if  t >= tprint
+        [Ep,Es] = energy(x,mb,mp,ms);
+         v = [t x(1) x(2) x(5) x(6) Ep Es];
+         save out v -ascii -append              
+         tprint =  tprint + dtp;
+     end
+     x = RK4(h,t,x,mb,mp,ms);
+     t = t+h;
+end

### 2016-08-23 Project3/example-codes_project3/problem3/energyplot.m

@@ -1,0 +1,25 @@
+clear; clf;
+out=load('out'); % load the data file out
+t =out(:,1); % time 
+xp=out(:,2); % the position of the primary star
+yp=out(:,3);
+xs=out(:,4); % the position of the secondary star
+ys=out(:,5); 
+Ep=out(:,6); % the energy of the primary star
+Es=out(:,7); % the energy of the secondary star
+
+%plot(xp,yp);
+%hold on
+%plot(xs-xp, ys-yp, 'r:')
+%xlabel('x','Fontsize',20)
+%ylabel('y','Fontsize',20)
+%set(gca, 'Fontsize', 20)
+%axis equal
+
+plot(t,Ep,'o')
+hold on
+plot(t,Es,'r')
+xlabel('t','Fontsize',20)
+ylabel('E','Fontsize',20)
+set(gca,'Fontsize',20)
+

### 2016-08-23 Project3/example-codes_project3/problem3/f.m

@@ -1,0 +1,15 @@
+function dxdt = f(t,x,mb,mp,ms)
+% a new input variabel mb has been added. 
+% include the BH gravity terms. see eq 38.
+  rp = sqrt(x(1)^2+x(2)^2); 
+rs = sqrt(x(5)^2+x(6)^2);
+  r = sqrt((x(5)-x(1))^2+(x(6)-x(2))^2);
+  dxdt(1) = x(3);
+  dxdt(2) = x(4);
+dxdt(3) = ((mb/rp^3)*(-x(1)))+((ms/r^3)*(x(5)-x(1)));
+dxdt(4) = ((mb/rp^3)*(-x(2)))+((ms/r^3)*(x(6)-x(2)));
+  dxdt(5) = x(7);
+  dxdt(6) = x(8);
+dxdt(7) = ((mb/rs^3)*(-x(5)))+((mp/r^3)*(x(1)-x(5)));
+dxdt(8) = ((mb/rs^3)*(-x(6)))+((mp/r^3)*(x(2)-x(6)));
+

### 2016-08-23 Project3/example-codes_project3/problem3/initialc.m

@@ -1,0 +1,23 @@
+function [t, x] =  initialc(D,mb,mp,ms)
+% input valiables: D,mb,mp,ms. D=Rp/Rtidal is the penetration factor. 
+% out variables: t and x. Note that x is a vector. 
+% Define the initial values of t and x as functions of the input valiables. 
+  phase=2*pi;
+  f=-acos(-1+D/5);
+t = (sqrt(2)/3)*(D^(3/2))*tan(f/2)*(3+(tan(f/2))^2); % initial time
+Rt=mb^(1/3);
+R=Rt*10;
+Xx=R*cos(f);
+Xy=R*sin(f);
+rdot=(mb^(1/3)/(sqrt(2*D)))*sin(f);
+fdot=(sqrt(2)/4)*(D^(-3/2))*(1+cos(f))^2;
+xdotx=(rdot*cos(f))-(R*fdot*sin(f));
+xdoty=(R*fdot*cos(f))+(sin(f)*rdot);
+x(1) = Xx+(ms*cos(phase+pi)); % x : primary star
+x(2) = Xy+(ms*sin(phase+pi)); % y
+x(3) = xdotx+(-ms*sin(phase+pi)); % dx/dt
+x(4) = xdoty+(ms*cos(phase+pi)); % dy/dt
+x(5) = Xx+(mp*cos(phase)); % x : secondary star
+x(6) = Xy+(mp*sin(phase)); % y
+x(7) = xdotx+(-mp*sin(phase)); % dx/dt
+x(8) = xdoty+(mp*cos(phase)); % dy/dt

### 2016-08-23 Project3/example-codes_project3/problem3/script4.m

@@ -1,0 +1,27 @@
+clear; clf;
+out=load('out'); % load the data file out
+t =out(:,1); % time 
+xp=out(:,2); % the position of the primary star
+yp=out(:,3);
+xs=out(:,4); % the position of the secondary star
+ys=out(:,5); 
+Ep=out(:,6); % the energy of the primary star
+Es=out(:,7); % the energy of the secondary star
+
+%plot(xp,yp);
+%hold on
+plot(xs-xp, ys-yp, 'r:')
+xlabel('x','Fontsize',20)
+ylabel('y','Fontsize',20)
+set(gca, 'Fontsize', 20)
+axis equal
+
+%plot(t,Ep,'o')
+%hold on
+%plot(t,Es,'r')
+%xlabel('t','Fontsize',20)
+%ylabel('E','Fontsize',20)
+%set(gca,'Fontsize',20)
+
+
+

### 2016-08-23 Python/python.py

@@ -1,0 +1,8 @@
+def countdown():
+    yield 3
+    yield 2
+    yield 1
+    yield "Blast Off!"
+
+g=countdown()
+next(g)

### 2016-08-24 workspace/DoubleBars/DoubleBars_v2.f90
Double Bars working!

@@ -19,132 +19,92 @@
  contains
 
    subroutine initialconditions
-     real :: A, B, C, omega, szebehely, E, U, T, E0
-     
-!m1=m3=M=1 - Outer Bar - constant
-!m2=m4=u=m - Inner Bar - variable
-!x1=-x3 - Outer
-!x2=-x4 - Inner
-!Vx1=-Vx3 - Outer
-!Vx2=-Vx4 - Inner
 
-     step = 1.e-1 !vary step size
-     m=.1 !vary mass of inner bar
+     step = 1.e-6 !vary step size
      start=0.
      end=20 !vary end time
-     x1=.4 !x coord of outer bar
-     x2=.2 !vary x coord of inner bar
      m1 = 1
-     m2 = m
-     m3 = 1
-     m4 = m
-     szebehely=.9 ! vary Szebehely constant - angular momentum
-     E0=-1.2 ! Total energy 
-     
-     omega=sqrt(szebehely/(-E0)) ! angular momentum
+     m2 = .05
+     m3 = .000005
+     m4 = .000005
 
-     ys(1) = x1     ! m1,x
-     ys(2) = 0   ! m1,vx
-     ys(3) = 0        ! m1,y
+     ys(1) = 0     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.05		!m1,vy
      
-     ys(5) = x2         ! m2,x
+     ys(5) = 1         ! m2,x
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
-    
-     ys(9) = -ys(1)       ! m3,x
-     ys(10) = -ys(2)       ! m3,vx
-     ys(11) = -ys(3)        ! m3,y
-     
-     ys(13) = -ys(5)     ! m4,x
-     ys(14) = -ys(6)      ! m4,vx
-     ys(15) = -ys(7)        ! m4,y
-     
-     U=(m2/(abs(ys(5)-ys(1))))+(m3/(abs(ys(9)-ys(1))))+(m4/(abs(ys(13)-ys(1))))+ &
[…]

### 2016-11-25 obs/ProblemSet1.tex
Started the write up

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{Observational Astrophysics Problem Set 1}
\begin{document}
\maketitle
\section{Question 1}
In order to derive Wiens's Displacement Law we need to find the maximum of Planck's Law, to do this we set the derivative to be zero: 
\begin{equation}
\frac{dI_\lambda(T)}{d\lambda} = \frac{d}{d\lambda}\left(\frac{2hc^2}{\lambda^5}\left(e^{\frac{hc}{\lambda kT}}-1\right)^{-1}\right) = 0
\end{equation}
Using the product rule $duv=vdu+udv$ yields the following:
\begin{equation}
\frac{-10hc^2}{\lambda^6}\left(e^{\frac{hc}{\lambda kT}}-1\right)^{-1}-\frac{2hc^2}{\lambda^5}\left(\frac{-hc}{\lambda^2kT}\right)\left(e^{\frac{hc}{\lambda kT}}\right)\left(e^{\frac{hc}{\lambda kT}}-1\right)^{-2} = 0
\end{equation}
As we are setting the equation to zero we can divide out the multiplicative terms giving:
\begin{equation}
\frac{-2hc^2}{\lambda^6}\left(e^{\frac{hc}{\lambda kT}}-1\right)^{-1}\left[5-\frac{hc}{\lambda kT}e^{\frac{hc}{\lambda kT}}\left(e^{\frac{hc}{\lambda kT}}-1\right)^{-1}\right] = 0
\end{equation}
If we now substitute $x=\frac{hc}{\lambda kT}$ we obtain:
\begin{equation}
5-xe^x\left(e^x-1\right)^{-1}
\end{equation}
which gives:
\begin{equation}
xe^x = 5(e^x-1)
\end{equation}
This equation has to be solved numerically however when solved for wavelength in nm and temperature in Kelvin we get:
\begin{equation}
\lambda_{max}=\frac{hc}{x}\frac{1}{kT}=\frac{2.89776829x10^6 nm . K}{T}
\end{equation}

\section{Question 2}
Specific intensity is distance independent because it contains the solid angle subtended by the source. If the source is unresolved i.e. at great distances like most stars
then this solid angle cannot be measured and instead of measuring specific intensity we would be measuring flux.

\section{Question 3}
The probability that a measurement is smaller than average by 2.5$\sigma$ is given by:
\begin{equation}
\int_{-\infty}^{\mu+2,5\sigma}\frac{1}{\sigma\sqrt{2\pi}}exp\left(-\frac{\left(x-\mu\right)^2}{2\sigma^2}\right)dt
\end{equation}
From the error table we get .99959 so the probability can be calculated as $1-(.5+.99959/2)\approx .0002$ or $.02\%$

\section{Question 4}
$B-V=8.35-7.81=0.54$ so from the table the spectral type is likely to be F8 with an effective temperature of 6250K.
It's absolute magnitude $M_v=4$ and so it's distance can be calculated using $D=10^{\left[\left(m-M+5\right)/5\right]} pc$ which 
equals $10^{\left[\left(7.81-4+5\right)/5\right]}=57.8 pc$.
The luminosity can be calculated from the bolometric magnitude so $\left(3\times10^{28}\right)\times10^{-0.4M_{bol}}=3.6\times10^{26} W$.

\section{Question 5}
\subsection{Part a}
Flux is given by $F=\frac{L}{4\pi r^2} W/m^2$ so solar luminosity is $3.846\times10^{26}W$ making Vega luminosity $1.538\times10^{28}$. 
Therefore Vega flux at Earth is $\frac{1.538\times10^{28}}{\left(4\pi 2.47\times10^{17}\right)^2}=2.52\times10^{-7} W/m^2$.
\subsection{Part b}
Flux density is also given by the inverse square rule so again $F_lambda=\frac{L}{4\pi r^2}$ but units are different so converting to cgs units we get:
$1W=10^7 erg s^{-1}$. Luminosity of Vega $=3.846\times10^{33} erg s^{-1}$. $8pc=2.469\times10^{19} cm$.

\subsection{Part c}

\end{document}

### 2016-11-25 workspace/DoubleBars/MScProjectDoubleBars.tex
Started the write up

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{MSc Project - Modelling a Double Barred Galaxy as a Double Binary}
\begin{document}
\maketitle
\section{Introduction}
The aim of this project was to create a simple model of a double barred galaxy using a double binary.

\section{Methods}
Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
\newline
The classical equations of motion for an n-body problem are:
\begin{equation}
m_i\mathbf{\ddot{r}}=\sum_{i\neq{j}}\frac{{m_i}{m_j}}{r^3_{ij}}\mathbf{r_{ij}}
\qquad
i=1,2,3...
\end{equation}
where $\mathbf{r_i}=\left(x_i,y_j\right)$ and $\mathbf{r_{ij}}=\mathbf{r_j}-\mathbf{r_i}$. So for a four body problem 
this equation results in the following:
\begin{equation}
\ddot{\mathbf{r_1}}=\frac{m_2\mathbf{r_{12}}}{r^3_{12}}+\frac{m_3\mathbf{r_{13}}}{r^3_{13}}+\frac{m_4\mathbf{r_{14}}}{r^3_{14}}
\end{equation}
\begin{equation}
\ddot{\mathbf{r_2}}=\frac{m_1\mathbf{r_{21}}}{r^3_{21}}+\frac{m_3\mathbf{r_{23}}}{r^3_{23}}+\frac{m_4\mathbf{r_{24}}}{r^3_{24}}
\end{equation}
\begin{equation}
\ddot{\mathbf{r_3}}=\frac{m_1\mathbf{r_{31}}}{r^3_{31}}+\frac{m_2\mathbf{r_{32}}}{r^3_{32}}+\frac{m_4\mathbf{r_{34}}}{r^3_{34}}
\end{equation}
\begin{equation}
\ddot{\mathbf{r_4}}=\frac{m_1\mathbf{r_{41}}}{r^3_{41}}+\frac{m_2\mathbf{r_{42}}}{r^3_{42}}+\frac{m_3\mathbf{r_{43}}}{r^3_{43}}
\end{equation}
The above equations were encoded into Fortran. A full code listing is supplied in a separate file.
\subsection{Initial Conditions}
As a starting point both the outer binary and inner binary were placed in circular orbits and tested separately and then combined to ensure a stable
starting point. The inner binary had initially a neglible mass.
The outer binary initial conditions were:
\begin{lstlisting}
m1=.5, m2=.5, x1=-.5, x2=.5, y1=0, y2=0,
vx1=0, vx2=0, vy1=-.5, vy2=.5
\end{lstlisting}
and for the inner binary:
\begin{lstlisting}
m3=.001, m3=.001, x3=-.001, x4=.001, y3=0, y4=0, 
vx3=0, vx4=0, vy3=-.5, vy4=.5
\end{lstlisting}
A plot of this initial configuration is shown in Fig. \ref{fig:stablebase}.
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/stablebase/Orbit.eps}
\caption{Initial stable configuration}
\label{fig:stablebase}
\end{figure}
The inner binary is only visible as a point at the centre of the outer binary.
I then proceeded to perturb the system by gradually increasing the mass of the inner binary. As the system became unstable I componensated for
the increasing mass of the inner binary by increasing the velocity of the outer binary and also decreasing the velocity and increasing the binary 
separation of the inner binary. In many of the configurations the systems were inherently unstable and produced collisions with some 
or all of the bodies being ejected from the system. These were rejected and only 'stable' configurations were considered.
Table \ref{tab:variables} shows the perturbations made to the system - only the variables in this table changed and all the other initial
conditions remained the same.
\begin{table}[ht!]
  \centering
  \caption{Configurations}
  \label{tab:variables}
  \begin{tabular}{ccccccccc}
   Config. & m3 & m4 & vy1 & vy2 & x3 & x4 & vy3 & vy4\\
    \hline
   1 & .002 & .002 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
   2 & .003 & .003 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
   3 & .004 & .004 & -.5 & .5 & -.004 & .004 & -.5 & .5\\
   4 & .004 & .004 & -.55 & .55 & -.004 & .004 & -.5 & .5\\
   5 & .004 & .004 & -.57 & .57 & -.004 & .004 & -.5 & .5\\
   6 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.3 & .3\\
   7 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.4 & .4\\
   8 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.35 & .35\\
   9 & .006 & .006 & -.6 & .6 & -.006 & .006 & -.3 & .3\\
   10 & .01 & .01 & -.7 & .7 & -.01 & .01 & -.2 & .2\\
   11 & .02 & .02 & -.75 & .75 & -.02 & .02 & -.1 & .1\\
   12 & .025 & .025 & -.65 & .65 & -.02 & .02 & -.5 & .5\\
   13 & .05 & .05 & -.9 & .9 & -.05 & .05 & -.1 & .1\\
   14 & .05 & .05 & -.9 & .9 & -.05 & .05 & -.15 & .15\\
   15 & .06 & .06 & -.95 & .95 & -.06 & .06 & -.15 & .15\\
   16 & .07 & .07 & -.1 & .1 & -.07 & .07 & -.15 & .15\\
   17 & .08 & .08 & -.105 & .105 & -.08 & .08 & -.12 & .12\\
  \end{tabular}
\end{table}

\section{Results}
As can be seen from Fig. \ref{fig:stablebase} even a neglible mass inner binary causes the outer binary to deviate from a circular orbit.
Figures Fig. \ref{fig:config1} to \ref{fig:config17} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
As the mass of the inner binary increased the system became increasingly unstable and the effect on the outer binary became more pronounced.
Even the 'stable' configurations listed in Table \ref{tab:variables} hold for only a few orbits.
I was unable to recreate a stable system for a mass $>.08$ for the inner bar.
The plots below reveal that increasing the mass of the inner binary causes the outer binary to separate and
form wider orbits.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/002-5-001/Orbit.eps}
\caption{Configuration 1}
\label{fig:config1}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/003-5-001/Orbit.eps}
\caption{Configuration 2}
\label{fig:config2}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/004-5-004/Orbit.eps}
\caption{Configuration 3}
\label{fig:config3}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/004-55-004/Orbit.eps}
\caption{Configuration 4}
\label{fig:config4}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/004-57-004/Orbit.eps}
\caption{Configuration 5}
\label{fig:config5}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/005-58-005-3/Orbit.eps}
\caption{Configuration 6}
\label{fig:config6}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/005-58-005-4/Orbit.eps}
\caption{Configuration 7}
\label{fig:config7}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/005-58-005-35/Orbit.eps}
\caption{Configuration 8}
\label{fig:config8}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/006-6-006-3/Orbit.eps}
\caption{Configuration 9}
\label{fig:config9}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/01-7-01-2/Orbit.eps}
\caption{Configuration 10}
\label{fig:config10}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/02-75-02-1/Orbit.eps}
\caption{Configuration 11}
\label{fig:config11}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/025-65-02/Orbit.eps}
\caption{Configuration 12}
\label{fig:config12}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/05-9-05-1/Orbit.eps}
\caption{Configuration 13}
\label{fig:config13}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/05-9-05-15/Orbit.eps}
\caption{Configuration 14}
\label{fig:config14}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/06-95-06-15/Orbit.eps}
\caption{Configuration 15}
\label{fig:config15}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/07-1-07-15/Orbit.eps}
\caption{Configuration 16}
\label{fig:config16}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./results/08-105-08-12/Orbit.eps}
\caption{Configuration 17}
\label{fig:config17}
\end{figure}
\end{document}

### 2016-11-25 workspace/DoubleBars/DoubleBars_v2.f90

@@ -21,7 +21,7 @@
 
    subroutine initialconditions
 
-     step = 1.e-5 !vary step size
+     step = 1.e-4 !vary step size
      start=0.
      end=650 !vary end time
 !     m1 = 1
@@ -48,12 +48,12 @@
      ys(9) = -.09    ! m3,x
      ys(10) = 0       ! lm3,vx
      ys(11) = 0        ! m3,y
-     ys(12) = -.07!-.01		!m3, vy
+     ys(12) = -.067!-.01		!m3, vy
 
      ys(13) = .09     ! m4,x
      ys(14) = 0     ! m4,vx
      ys(15) = 0        ! m4,y
-	ys(16) = .07!.01		!m4, vy
+	ys(16) = .067!.01		!m4, vy
 
 !     ys(1) = 0     ! m1,x
 !     ys(2) = 0  ! m1,vx

### 2016-11-29 workspace/DoubleBars/MScProjectDoubleBars.tex
Started the write up

@@
 The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
+Please see the Compilers subsection for a few notes on compilers - may or may not be relevant to the project.
 \newline
@@
 \end{equation}
-where $\mathbf{r_i}=\left(x_i,y_j\right)$ and $\mathbf{r_{ij}}=\mathbf{r_j}-\mathbf{r_i}$. So for a four body problem 
+where $\mathbf{r_i}=\left(x_i,y_i\right)$ and $\mathbf{r_{ij}}=\mathbf{r_j}-\mathbf{r_i}$. So for a four body problem 
 this equation results in the following:
 \begin{equation}
-\ddot{\mathbf{r_1}}=\frac{m_2\mathbf{r_{12}}}{r^3_{12}}+\frac{m_3\mathbf{r_{13}}}{r^3_{13}}+\frac{m_4\mathbf{r_{14}}}{r^3_{14}}
+\ddot{\mathbf{r_1}}=\frac{m_2\mathbf{r}_{12}}{r^3_{12}}+\frac{m_3\mathbf{r}_{13}}{r^3_{13}}+\frac{m_4\mathbf{r}_{14}}{r^3_{14}}
 \end{equation}
 \begin{equation}
-\ddot{\mathbf{r_2}}=\frac{m_1\mathbf{r_{21}}}{r^3_{21}}+\frac{m_3\mathbf{r_{23}}}{r^3_{23}}+\frac{m_4\mathbf{r_{24}}}{r^3_{24}}
+\ddot{\mathbf{r_2}}=\frac{m_1\mathbf{r}_{21}}{r^3_{21}}+\frac{m_3\mathbf{r}_{23}}{r^3_{23}}+\frac{m_4\mathbf{r}_{24}}{r^3_{24}}
 \end{equation}
 \begin{equation}
-\ddot{\mathbf{r_3}}=\frac{m_1\mathbf{r_{31}}}{r^3_{31}}+\frac{m_2\mathbf{r_{32}}}{r^3_{32}}+\frac{m_4\mathbf{r_{34}}}{r^3_{34}}
+\ddot{\mathbf{r_3}}=\frac{m_1\mathbf{r}_{31}}{r^3_{31}}+\frac{m_2\mathbf{r}_{32}}{r^3_{32}}+\frac{m_4\mathbf{r}_{34}}{r^3_{34}}
 \end{equation}
 \begin{equation}
-\ddot{\mathbf{r_4}}=\frac{m_1\mathbf{r_{41}}}{r^3_{41}}+\frac{m_2\mathbf{r_{42}}}{r^3_{42}}+\frac{m_3\mathbf{r_{43}}}{r^3_{43}}
+\ddot{\mathbf{r_4}}=\frac{m_1\mathbf{r}_{41}}{r^3_{41}}+\frac{m_2\mathbf{r}_{42}}{r^3_{42}}+\frac{m_3\mathbf{r}_{43}}{r^3_{43}}
 \end{equation}
@@
 \end{lstlisting}
-A plot of this initial configuration is shown in Fig. \ref{fig:stablebase}.
+Plots of the two initial binaries are shown in Fig. \ref{fig:outerbinary} and 
+\ref{fig:innerbinary}. The scale of the inner binary is so small that it is difficult
+to see in GnuPlot.
 \begin{figure}[H]
 \centering
-\includegraphics[width=0.75\textwidth]{./results/stablebase/Orbit.eps}
-\caption{Initial stable configuration}
-\label{fig:stablebase}
+\includegraphics[width=0.75\textwidth]{./results/outerbinary/Orbit.eps}
+\caption{Outer Binary}
+\label{fig:outerbinary}
 \end{figure}
-The inner binary is only visible as a point at the centre of the outer binary.
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.75\textwidth]{./results/innerbinary/Orbit.eps}
+\caption{Inner Binary}
+\label{fig:innerbinary}
+\end{figure}
 I then proceeded to perturb the system by gradually increasing the mass of the inner binary. As the system became unstable I componensated for
@@
     \hline
-   1 & .002 & .002 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
-   2 & .003 & .003 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
-   3 & .004 & .004 & -.5 & .5 & -.004 & .004 & -.5 & .5\\
-   4 & .004 & .004 & -.55 & .55 & -.004 & .004 & -.5 & .5\\
-   5 & .004 & .004 & -.57 & .57 & -.004 & .004 & -.5 & .5\\
-   6 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.3 & .3\\
-   7 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.4 & .4\\
-   8 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.35 & .35\\
-   9 & .006 & .006 & -.6 & .6 & -.006 & .006 & -.3 & .3\\
-   10 & .01 & .01 & -.7 & .7 & -.01 & .01 & -.2 & .2\\
-   11 & .02 & .02 & -.75 & .75 & -.02 & .02 & -.1 & .1\\
-   12 & .025 & .025 & -.65 & .65 & -.02 & .02 & -.5 & .5\\
-   13 & .05 & .05 & -.9 & .9 & -.05 & .05 & -.1 & .1\\
-   14 & .05 & .05 & -.9 & .9 & -.05 & .05 & -.15 & .15\\
-   15 & .06 & .06 & -.95 & .95 & -.06 & .06 & -.15 & .15\\
-   16 & .07 & .07 & -.1 & .1 & -.07 & .07 & -.15 & .15\\
-   17 & .08 & .08 & -.105 & .105 & -.08 & .08 & -.12 & .12\\
+   1 & .001 & .001 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
+   2 & .002 & .002 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
+   3 & .003 & .003 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
+   4 & .004 & .004 & -.5 & .5 & -.004 & .004 & -.5 & .5\\
+   5 & .004 & .004 & -.55 & .55 & -.004 & .004 & -.5 & .5\\
+   6 & .004 & .004 & -.57 & .57 & -.004 & .004 & -.5 & .5\\
+   7 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.3 & .3\\
+   8 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.4 & .4\\
+   9 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.35 & .35\\
+   10 & .006 & .006 & -.6 & .6 & -.006 & .006 & -.3 & .3\\
+   11 & .01 & .01 & -.7 & .7 & -.01 & .01 & -.2 & .2\\
+   12 & .02 & .02 & -.75 & .75 & -.02 & .02 & -.1 & .1\\
+   13 & .025 & .025 & -.65 & .65 & -.02 & .02 & -.5 & .5\\
+   14 & .05 & .05 & -.9 & .9 & -.05 & .05 & -.1 & .1\\
+   15 & .05 & .05 & -.9 & .9 & -.05 & .05 & -.15 & .15\\
+   16 & .06 & .06 & -.95 & .95 & -.06 & .06 & -.15 & .15\\
+   17 & .07 & .07 & -.1 & .1 & -.07 & .07 & -.15 & .15\\
+   18 & .08 & .08 & -.105 & .105 & -.08 & .08 & -.12 & .12\\
   \end{tabular}
 \end{table}
+\subsection{Compilers}
+As I had converted OdeInt into Fortran 90 I wanted to make sure I hadn't introduced any errors. I therefore compared the output of the Numerical Recipes
+version of OdeInt with mine. I was compiling with GFortran and the Fortran 77 version was compiled with g77. They were different! So naturally concerned 
+I then compiled the Fortran 77 version with GFortran. The results were identical to mine, which satisified me that I hadn't introduced any errors.
 
 \section{Results}
-As can be seen from Fig. \ref{fig:stablebase} even a neglible mass inner binary causes the outer binary to deviate from a circular orbit.
-Figures Fig. \ref{fig:config1} to \ref{fig:config17} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
+As can be seen from Fig. \ref{fig:config1} even a neglible mass inner binary causes the outer binary to begin to separate and become less tightly bound.
+Figures Fig. \ref{fig:config1} to \ref{fig:config18} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
+In many cases the inner binary is only visible as a point at the centre of the outer binary.
 As the mass of the inner binary increased the system became increasingly unstable and the effect on the outer binary became more pronounced.
@@
 form wider orbits.
-
 \begin{figure}[H]
 \centering
-\includegraphics[width=0.75\textwidth]{./results/002-5-001/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/stablebase/Orbit.eps}
 \caption{Configuration 1}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/003-5-001/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/002-5-001/Orbit.eps}
 \caption{Configuration 2}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/004-5-004/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/003-5-001/Orbit.eps}
 \caption{Configuration 3}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/004-55-004/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/004-5-004/Orbit.eps}
 \caption{Configuration 4}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/004-57-004/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/004-55-004/Orbit.eps}
 \caption{Configuration 5}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/005-58-005-3/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/004-57-004/Orbit.eps}
 \caption{Configuration 6}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/005-58-005-4/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/005-58-005-3/Orbit.eps}
 \caption{Configuration 7}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/005-58-005-35/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/005-58-005-4/Orbit.eps}
 \caption{Configuration 8}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/006-6-006-3/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/005-58-005-35/Orbit.eps}
 \caption{Configuration 9}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/01-7-01-2/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/006-6-006-3/Orbit.eps}
 \caption{Configuration 10}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/02-75-02-1/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/01-7-01-2/Orbit.eps}
 \caption{Configuration 11}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/025-65-02/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/02-75-02-1/Orbit.eps}
 \caption{Configuration 12}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/05-9-05-1/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/025-65-02/Orbit.eps}
 \caption{Configuration 13}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/05-9-05-15/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/05-9-05-1/Orbit.eps}
 \caption{Configuration 14}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/06-95-06-15/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/05-9-05-15/Orbit.eps}
 \caption{Configuration 15}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/07-1-07-15/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/06-95-06-15/Orbit.eps}
 \caption{Configuration 16}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/08-105-08-12/Orbit.eps}
+\includegraphics[width=0.75\textwidth]{./results/07-1-07-15/Orbit.eps}
 \caption{Configuration 17}
@@
 \end{figure}
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.75\textwidth]{./results/08-105-08-12/Orbit.eps}
+\caption{Configuration 18}
+\label{fig:config18}
+\end{figure}
 \end{document}

### 2016-11-29 workspace/DoubleBars/results/innerbinary/DoubleBars_v2.f90

@@ -21,7 +21,7 @@
 
    subroutine initialconditions
 
-     step = 1.e-5 !vary step size
+     step = 1.e-6 !vary step size
      start=0.
      end=50 !vary end time
 !     m1 = 1
@@ -30,10 +30,10 @@
 !     m4 = .000005
 !      m1=.5
 !      m2=.5
-		m1=.75   !blue
-		m2=.25  !red
-      m3=0!.001    !black
-      m4=0!.001   !orange
+		m1=0!.5   !blue
+		m2=0!.5  !red
+      m3=.001    !black
+      m4=.001   !orange
 
      ys(1) = -.5     ! m1,x
      ys(2) = 0  ! m1,vx

### 2016-11-29 workspace/DoubleBars/results/outerbinary/DoubleBars_v2.f90

@@ -21,17 +21,17 @@
 
    subroutine initialconditions
 
-     step = 1.e-5 !vary step size
+     step = 1.e-7 !vary step size
      start=0.
-     end=50 !vary end time
+     end=25 !vary end time
 !     m1 = 1
 !     m2 = .05
 !     m3 = .000005
 !     m4 = .000005
 !      m1=.5
 !      m2=.5
-		m1=.75   !blue
-		m2=.25  !red
+		m1=.5   !blue
+		m2=.5  !red
       m3=0!.001    !black
       m4=0!.001   !orange
 

### 2016-11-29 workspace/DoubleBars/results/test/DoubleBars_v2.f90

@@ -21,7 +21,7 @@
 
    subroutine initialconditions
 
-     step = 1.e-5 !vary step size
+     step = 1.e-7 !vary step size
      start=0.
      end=50 !vary end time
 !     m1 = 1
@@ -30,10 +30,10 @@
 !     m4 = .000005
 !      m1=.5
 !      m2=.5
-		m1=.75   !blue
-		m2=.25  !red
-      m3=0!.001    !black
-      m4=0!.001   !orange
+		m1=0!.5   !blue
+		m2=0!.5  !red
+      m3=.001    !black
+      m4=.001   !orange
 
      ys(1) = -.5     ! m1,x
      ys(2) = 0  ! m1,vx

### 2017-01-13 ObsProjects/ArxivReview.tex
OA Project

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{A review of 'Searching for Primordial Black Holes in the radio and X-ray sky'}
\begin{document}
\maketitle
'Searching for Primordial Black Holes in the radio and X-ray sky' is a short paper \citep{gaggero} published on Arxiv on 1st December 2016 - https://arxiv.org/pdf/1612.00457v1.pdf.
It describes a technique for modelling primordial black holes (PBHs) and then comparing the 
model with observational measurements in both the radio and X-ray of potential PBH candidates in the Milky Way.
The recent LIGO detection of gravitational waves \citep{abbott}  has inferred the existence of $\approx30M_{\odot}$ black holes and prompted 
the suggestion that these may be PBHs \citep{bird}. 
A PBH is a hypothetical black hole that supposedly formed from the very dense matter found during the universe's early expansion. 
The idea that PBHs can account for the dark matter in the universe is not new e.g. \citep{bertone1} and \citep{bertone2}.

The study focuses on the galactic centre of the Milky Way and, given current estimates of the mass of the inner bulge, in order 
to account for all the dark matter there should be approximately $10^9$ PBHs in that vicinity. As the galactic centre contains
much high energy gas it is probable that a large portion of these PBHs would form accretion disks which would 
emit radiation at a wide range of wavelengths. This paper is restricted to radio and X-ray.
This study modelled PBHs using a Monte Carlo simulation which is a computer simulation based on sampling many random 
events and is widely used in astrophysics. A model of a galaxy was then built populated by PBHs. The accretion rate, radio
and X-ray emissions were then calculated.
These results were then compared with observations. For the radio candidates a VLA (Very Large Array) survey was used
to identify possible PBH candidates. A total of 24 possible PBHs were identified. These were compared with X-Ray sources from Chandra.
However the paper suggests that these candidates have to be ruled out as their X-Ray and radio fluxes do not lie together on the fundamental plane - 
a empirical relation familiar to MSc students from the Galaxy module. 
X-Ray candidates were selected from a survey carried out by the NuSTAR space telescope - a total of 70 were found to
be potential PBHs. However when compared with the VLA survey again no matches were found and so all these candidates were ruled out.
The paper concludes by saying that their accretion model has added more constraints on the possibility for PBHs to account for dark matter and suggests
that PBHs may only account for less than 20\% of dark matter. The study suggests that this technique
could be utilised further with the new radio telescopes such as MeerKat (South Africa) and the planned Square Kilometer
Array (SKA - South Africa and Australia) which could detect fainter sources.

As a newcomer to this field I found the paper readable and the introduction explained the background and purpose of the 
paper well. Although the conclusion was ultimately a negative one, the authors went on to investigate how the 
study could be repeated with new planned telescopes. The references indicate quite a substantial literature review
for such a small paper. I found the paper to be concise and informative. The reader could easily consult the references
if any further information on this subject was needed. I couldn't find any references to whether it
has been accepted by any refereed journals. I just have one criticism and that for me there wasm't a clear explanation as how they
reached the conclusion that $\approx30M_{\odot}$ PBHs account for less than $20\%$ of dark matter in the Milky Way.

\begin{thebibliography}{1}
\bibitem[Gaggero et al.(2016)]{gaggero}D. Gaggero, G. Bertone, F. Calore, R. M. T. Connors, M. Lovell, S. Marko, E. Storm Searching for Primordial Black Holes in the radio and X-ray sky (2016), arXiv:1612.00457 
\bibitem[Abbott et al.(2016)]{abbott}B. P. Abbott et al. (Virgo, LIGO Scientific), Phys. Rev. Lett. 116, 061102 (2016), arXiv:1602.03837 
\bibitem[Bird et al.(2016)]{bird}S. Bird, I. Cholis, J. B. Munoz, Y. Ali-Haimoud, M. Kamionkowski, E. D. Kovetz, A. Raccanelli, and A. G. Riess, Phys. Rev. Lett. 116, 201301 (2016), ArXiv:1603.00464
\bibitem[Bertone et al.(2005)]{bertone1}G. Bertone, D. Hooper, and J. Silk, Phys. Rept. 405, 279 (2005), arXiv:hep-ph/0404175
\bibitem[Bertone(2010)]{bertone2}G. Bertone, Nature 468, 389 (2010), arXiv:1011.3532
\end{thebibliography}
\end{document}

### 2017-01-13 ObsProjects/SgrA*.tex
OA Project

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage{gensymb}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.png}
\author{Howard Kinsman}
\title{A Multiwavelength Report on Recent Findings on Sgr A*}
\begin{document}
\maketitle
\subsection{Introduction}
Einstein's General Theory of Relativity (GR) predicts the existence of blackholes and the closest and best known candidate is Sgr A* at the centre of the Milky Way. It is believed to be a supermassive 
blackhole (SMBH) and lies at a distance of approximately 8 kpc with an estimated mass of $4\times10^6 M_{\odot}$. It's proximity therefore makes it ideal for the study of blackholes and as a 
potential testbed for GR, as discussed later.
It was first identified as a radio source \citep{balick} back in 1974. It has been studied using the entire electromagnetic spectrum although interstellar extinction makes some wavelengths more 
problematic than others.
Even though, by definition, black-holes are black, they do emit electromagnetic radiation from near the event horizon - as gas and plasma are accreted by the blackhole this material is heated up
and large amounts of energy are emitted and radiated at all wavelengths of the electromagnetic spectrum. However since optical radiation is completely absorbed the best observing bands are radio 
(including sub-mm), the NIR (including far infrared) and X-rays.
The scale of a blackhole is set by the size of the Schwarzschild radius $R_s=2GM_{BH}/c^2\approx 3km (M_{bh}/M_{\odot})$. For Sgr A* this would lead to $\theta_{RS}=10\nu as$  however
for Sgr A* the angular size, affected by gravitational lensing of its own potential, is predicted to be $\approx 10 \mu as$ making it the largest subtended angle blackhole anywhere in the universe and so
is ideal for study. The distance to the Galactic Centre has been determined through high-resolution spectroscopy and the proper motion of stars giving a distance of $D=8.3(\pm 0.4) kpc$.
Apart from mass the other most important parameter of a blackhole is its accretion rate. This is determined by the distance from which it can accrete, the Bondi radius, given by
$R_B=2GM/v_w^2$ where v is the speed of surrounding gas. Estimates for Sgr A* indicate $R_b\approx 2.5\times10^5 R_s \approx 0.1pc$ determined by radio polarization measurements.
Sgr A* has extremely low levels of activity. Its current accretion rate is about four orders of magnitude lower than that required to for a SBMH to grow in a Hubble time \citep{falke1}, its radio emission 
is lower than expected and the amount of gas available as fuel is much lower that required to fuel an SMBH of that size. This has greatly constrained the models of Sgr A*.

This report is a short discussion on recent (2016) findings that have been published on Arxiv with an emphasis on multiwavelength comparisions and the potential for future studies. However the report
begins with a brief summary of the orbits of stars around Sgr A* surveyed over a number of years using the near infrared (NIR) as this data provides some of the most compelling evidence
that Sgr A* is actually a SMBH.

\subsection{Orbiting Stars and Mass of Sgr A* - IR}
The first evidence for a central dark mass was suggested in 1980 through observations from the mid-infrared and radio recombination line observations \citep{falke1}. Although it wasn't until the 
invention of adaptive optics for ground based telescopes that the resolution became sensitive enough to track stars orbiting this mass.
Due to the high levels of exinction when viewing the galactic centre in the optical the NIR has been used to view stars orbiting Sgr A* for decades. There are numerous papers on this subject but this
paragraph begins by citing from \citep{schodel} an early publication on this subject. This survey carried out ten years of high resolution NIR imaging and spectroscopy of the central region of the Milky Way in the
vicinity of Sgr A*. They used an 8m ground-based telescope to study the orbits of S1 and S2, the two closest stars to Sgr A*. S2, the star closest to Sgr A* at the time, was found to be in a
highly elliptical Keplerian orbit with a period of 15.2 years and a peri-centre distance of 17 light hours - at peri-centre S2 was found to have a velocity in excess of 5000 km/s.
From this data they were able to deduce that this orbit would require an enclosed point mass of $3.7\pm1.5\times10^6 M_{\odot}$ at the position of Sgr A*. At this time there were various models
that tried to explain Sgr A* e.g. a dark mass (low mass stars, neutron stars or stellar black holes), a ball of heavy fermions, or a  ball of bosons. This study effectively ruled out the first two of 
these models and put constraints on the possibility of the existence of a ball of bosons. The study was a major first step in testing the SMBH model.
By 2008, 16 years  of stellar orbits of 28 stars had been monitored \citep{gillessen1} using high resolution NIR. This survey combined a long time baseline with high astrometric accuracy using adaptive
optics. This further refined the mass of Sgr A* and showed that a single point mass was the best fit model for this system, pointing again to the SMBH model - see \ref{fig:orbits}

Other evidence that points to the association dark mass wit Sgr A* is its own proper motion. Through VLBI it has been possible to track its motion over several years and its motion has been found to be
entirely consistent with the motion of solar system around the Galactic Centre. This implies the radio source is actually the Galactic Centre.

Today, about 30 stellar orbits have been monitored in the centre of the Milky Way and, combined, this data has yielded the most precise measurement of the mass of Sgr A* as 
$4.23\pm0.14\times10^6 M_{\odot}$ \citep{goddi} with S2 achieving a velocity of 10000 km/s. In 2016 a new methodology of data reduction was piloted \citep{boehle}. They used the star S0-38, which is an
order of magnitude fainter than S2, and so much more difficult to track. The paper illustrates the use of a technique known as speckle holography which allows the speckle imaging data from 1995-2005 to be
re-analysed and fainter stars such as S0-38 to be tracked. The study has put further constraints on both the mass and distance of Sgr A* and reduced the uncertainties to a confidence level of 99.7\%.
The paper hopes that these improved constraints on the gravitational potential in the vicinity of Sgr A*, together with the addition of S0-38, will allow future tests of GR to be carried out when S2
goes through its time of closest approach in 2018. One of the first relatavistic observations for GRAVITY, a second generation instrument on the Very Large Telescope Interferometer (VLTI), an adaptive
optics assisted interferometer, will be to observe the peri-astron shift of S2 in 2018. With a precision of about $10\mu$as GRAVITY will push interferometry accuracy way beyond what is possible today.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{./orbits}
\caption{Inferred orbits of 6 stars around SMBH candidate SGR A* \citep{eisenhower}}
\label{fig:orbits}
\end{figure}

\subsection{Radio}
As mentioned previously blackholes do emit radiation from around the event horizon due to accretion. It has been calculated \citep{falke2} that the visual appearance of a blackhole would appear as a bright
photon ring with a dim "shadow" within its interior. Photons orbiting very close to the event horizon will disappear into the blackhole whilst those further away will escape to infinity therefore
a clear boundary between light and dark regions is predicted.
Gravitational lensing could increase the size of this shadow to bring it within limits VLBI measurements at mm wavelengths. 
Radio emission from Sgr A* is believed to be produced the accretion onto the SMBH. At millimetre wavelengths, Very Long Baseline Interferometry (VLBI) can resolve the accretion region of the SMBH.
It is believed that this radio emission is either synchotron radiation from a jet outflow, accretion flow onto the SMBH, or an isothermal jet coupled to an accretion flow \citep{ortiz}.
Wavelengths longer than a few centimetres are scattered by the interstellar medium but at smaller wavelengths this scattering is reduced. The scattering obeys a $\lambda^2$ dependence \citep{ortiz} but
at shorter wavelengths this dependence is seen to diminish indicating the structure of Sgr A* may be revealed.
In 2016 a study was carried out to attempt to identify the 2D structure of Sgr A* \citep{ortiz}. Up until this time existing VLBI baselines were limited in the north-south baselines however this study
combined the National Radio Astronomy Observatory Very Long Baseline Array (VLBA) with the Large Millimieter Telescope Alfonso Serrano (LMT) to act a single VLBI array. This provided good north-south as
well as east-west coverage. The study revealed that the image of Sgr A* resembles an elliptical Gaussian. 
A similar VLBI survery was carried out \citep{brinkerink} also using the VLBA, LMT and the Robert C. Byrd Green Bank Telescope (GBT) but this time at a 3mm wavelength using closure phase
interferometry. Closure amplitudes are formed by combining complex ampltitudes in the correlated data from four different telescopes so that the telescope-based gain errors cancel \citep{falke1}.
 They also managed to account for some structure in the Sgr A8 image. The paper describes an Eastern secondary source separate from the primary source, although the authors do
note that this asymmetry may potentially be due to interstellar scattering rather than being intrinisic to the source.
The Atacama Large Millimetre Array (ALMA) is the most sensitive (sub)mm-wave telescope ever builts and consists of 50 individual antennas of 12m diameter. Joint VLBI observations which include ALMA are planned
for 2017 and should provide unprecedented sensitivity due to the improved north-south baseline. Given the overwhelming evidence that Sgr A* lies in the close vicinity to a SMBH there are a number of ongoing observations which aim to test for the existence of blackholes, test GR and study 
spacetime around blackholes \citep{goddi}. Also submm telescopes are combining around the world, icluding ALMA, to form the Event Horizon Telescope (EHT) which should provide unprecedented
sensitivity for VLBI.

The radio spectrum for Sgr A* exhibits a slow rise with frequency until it reaches the submm band where it peaks and then cuts off - this has been termed the "submm bump". It is generally thought that
this is due to synchtron emission in transition from being optically thick to optically thin \citep{falke1}.

\subsection{G2 and Gamma}
A notable, and widely publicised, event occurred in the vicinity of Sgr A* during 2013-14. It was reported \citep{gillessen2} that a gas cloud, now known as G2, of the order three Earth masses, was on
a highly eccentric orbit towards the SMBH and it's predicted pericentre passage would occur late 2013 and would last up to a year. This was again observed using IR. 
This sparked interest in the scientific community as it was assumed that in-falling matter into the SMBH could accelerate
high-energy particles resulting emission at various wavelengths, including gamma rays. A four year observational campaign was therefore started by using the Major Atmospheric Gamma Imaging Cherenkov
(MAGIC) telescopes \citep{ahnen}. These two 17m diameter telescopes record flashes of Cherenkov light produced the Extensive Air Showers which occurr in the upper atmosphere and are caused by very
high energy $\gamma$ ray photons with energies $>50GeV$. The results however were disappointing and the close proximity of G2 to the SMBH had no measureable effect on $\gamma$ ray emissions, although
the authors hope this study will act as a baseline for monitoring any flaring activity in the future. These results, however, are not surpising as it had already been reported \cite{ghez} that G2
had passed by Sgr A* largely unaffected when monitored using NIR.

\subsection{X-ray}
Since the launch of Chandra in 1999 it was found that Sqr A* is a source of suprisingly weak X-rays \citep{falke1}.

\subsection{Multiwavelength}
The flaring activity associated with Sgr A* are also widely studied and observed phenomena. These can be observed using infrared, X-ray and submillimetre wavelengths. The origin of these flares is still
being debated but to understand this activity it is important to compare results from all of these wavelengths. A multiwavelength campaign \citep{mossoux} was therefore carried out in 2014, also timed
to be carried out at G2's closest approach.
This ambitious study aimed to make simultaneous observations using X-Ray Multi-Mirror Mission (XMM-Newton) and Wide Field Camera 3 (WFC3) onboard the Hubble Space Telescope (HST) giving observations in X-Ray and
NIR respectively. NIR observations were also carried out using the Spectrograph for Integral Field Observations in the Near Infrared (SINFONI) at the VLT in Chile. 
Radio observations were also carried using the Karl G. Jansky Very Large Array (VLA) in the X band (3.75-2.5cm) 
and also using the Combined Array for Research in Millimeter-wave Astronomy (CARMA) at 3.2mm. There was no increased flaring activity detected that could be attributed to the passage of G2 however a number of flares
were detected at the various wavelengths allowing the data to be analysed and compared using a multiwavelength approach. 
In total three flares were detected by the WCF3 on board HST plus two more using SINFONIA on the VLT. These were very luminous NIR flares but given that SINFONIA and WFC3 have very high detection limits this
is to be expected statistically. Two X-ray flares were detected by XMM-Newton and were two of the brightest flares ever detected by XMM-Newton. Two of the NIR flares had definite X-ray counterparts
and the one had the highest NIR-to-X-ray ratio ever observed. It seems that X-ray flares are often accompanied by NIR flares but not necessarily vice versa. Data from CARMA revealed a slight 
'bump' but this could not be correlated with any of the NIR/X-ray flares. 
The three main explainations for these flares are: a jet with clumps of ejected material, short-lived "hot spots" orbiting the blackhole or statistical fluctuations in the accretion flow \citep{goddi}.
With its $10\mu$as precision GRAVITY may also soon be able to determine the nature of these flares and settle this debate. GRAVITY should also have the precision to detect Lense-Thirring precession
for stars orbiting the SMBH and hopefully constrain the blackhole spin from stellar orbits.

In a separate study \citep{stone} the Herschel Space Observatory's (no longer operational) Spectral and Photometric Imaging Receiver (SPIRE) was used to carry out the longest ever continuous submillimetre 
survey of Sgr A*. These were observed in three FIR bands: 0.25, 0.35 and 0.5mm. These wavelength are difficult, if not impossible, to observe from the ground. The aim of the study
was to test for the variability of luminosity of the accretion flow into SMBH.
This was then compared with ground-based observations at the Caltech Submillimetre Observatory using 0.85mm and X-ray data from XMM-Newton. Small (<1\%) variations were seen at the three wavelengths
used by SPIRE and comparison with 0.5mm ground data support these small variations. However comparison XMM-Newton didn't reveal any corresponding variations.

\subsection{Pulsars}
Given the overwhelming evidence that Sgr A* lies in the close vicinity to a SMBH there are a number of ongoing observations which aim to test for the existence of blackholes, test GR and study 
spacetime around blackholes \citep{goddi}. Some very precise tests of GR have been carried out using pulsars: a pulsar in a binary can be used to probe the spacetime around that system by comparing the 
"pulsar clock" with the radio telescope maser. However it is believed that a pulsar orbiting a blackhole could provide an even more accurate test. No such system has yet been found however it is 
believed that there are many pulsars orbiting Sgr A*. The huge mass of SMBH would provide a very accurate test of GR. Unfortunately only five pulsars within 0.5\degree of Sgr A* have so far been found,
the lack of which
has been attributed to interstellar scattering. However in 2013 radio emission from a magnetar (PSR J1745-2900) was detected supporting the hypothesis that many more pulsars may yet still be found. It is
hoped that ALMA will help achieve this aim.
Through the use of modelling \citep{rajwade} state that it is not surpising current surveys have failed to find many pulsars in the Galactic Centre has they have only probed $<2\%$ of the population.
The authors state that upcoming telescopes such as the Square Kilometre Array would probe deepper and survey a more sizeable population of the pulsars in the Galactic Centre.

\subsection{Summary}
This report has so far mentioned three ongoing projects to test GR: the blackhole shadow using EHT, stars orbiting Sgr A* using GRAVITY and pulsars with ALMA. Each of these techniques uses utilises
different wavelengths and observational techniques: mm-VLBI, astrometry with NIR and pulsar timing with radio. It is hoped the combination of these results will test the validity of GR with
more accuracy than ever before. The increasing precision of VLBI observations has prompted some authors \citep{haggard} to suggest that we are approaching the observational limit to where quantum 
gravitational effects may be observed. Sgr A* provides us with the most direct evidence of the existence of blackholes and is by far the closest so ongoing observations at all possible wavelengths are
essential to provide us with more information on the physics of blackholes and ultimately a test of GR in the strong field limit.

\begin{thebibliography}{1}
\bibitem[Balick et al.(1974)]{balick}Balick, B., Brown R. L., (1974), ApJ, 194, 265
\bibitem[Schodel et al.(2002)]{schodel}Schodel, R., Ott, T., Genzel, R., Hofmann, R., Lehnert, M., Eckart, A., et al., 2002, Closest Star Seen Orbiting the Supermassive Black Hole at the Centre of the Milky Way, 
arXiv:astro-ph/0210426v1
\bibitem[Goddi et al.(2016)]{goddi}Goddi, C., Falcke, H., Rezzolla, L., Brinkerink, C. et al., BlackHoleCam: fundemental physics of the Galactic center, arXiv:1606.08879c1, WSPC Proceedings, 2016 
\bibitem[Boehle et al.(2016)]{boehle}Boehle, A., Ghez, A., M., Schodel, R., Meyer, L. Yelda, S. et al, 2016, An Improved Distance and Mass Estimate for Sgr A* from a Multistar Orbit Analysis, 
arXiv:1607.05726v1
\bibitem[Gillessen et al.(2009)]{gillessen1}Gillessen, S., Eisenhauer, F., Trippe S., et al., 2008, Monitoring Stellar Orbits Around the Massive Black Hole in the Galactic Center, The Astrophysical Journal,
692 (2): 1075-1109, arXiv:0910.3069v1
\bibitem[Ortiz-Leon et al.(2016)]{ortiz}Ortiz-Leon, G., N., Johnson, M., D., Doelman, S.,S., Blackburn, L., et al., 2016, The Intrinsic Shape of Sagittarius A* at 3.5mm Wavelength, arXiv:1601.06571v2
\bibitem[Falke, Markoff(2013)]{falke1}Falke, H., Markoff, S. B., 2013, Towards the event horizon - the supermassive black hole in the Galactic Center, arXiv:1311.184v1
\bibitem[Brinkerink et al.(2016)]{brinkerink}Brinkerink, C., D., Muller, C., Falcke, H., Bower, G., C., et al., 2016, Asymmetric structure in Sgr A* at 3mm from closure phase measurements with VLBA, GBT and LMT,
MNRAS, arXiv:1608.0651v1
\bibitem[Gillessen et al.(2012)]{gillessen2}Gillessen, S., Genzel, R., Fritz, T., et al., 2012, Nature, 481, 51
\bibitem[Ahnen et al.(2016)]{ahnen}Ahnen, M.,L., Ansoldi, S., Antonelli, L., A., Antoranz, P., Arcaro, C., et al., 2016, Observations of Sagittarius A* during the pericenter passage of the G2 object with MAGIC, A\&A, arXiv:1611.07095v1
\bibitem[Ghez et al.(2014)]{ghez}Ghez, A., M., Witzel, G., Sitarski, B., et al, 2014, The Astronomer's Telegram, 6110, 1
\bibitem[Mossoux et al.(2016)]{mossoux}Mossoux, E., Grosso, N., Bushouse, H., Eckart, A., Yusef-Zadeh, F., et al., Multiwavelngth study of the flaring activity of Sgr A* in 2014 February-April, 2016,
A\&A, ArXiv:1603.01048v1
\bibitem[Stone et al.(2016)]{stone}Stone, J., M., Marrone, D., P., Dowell, C., D., Schulz, B., et al., Far Infrared Variability of Sagittarius A*: 25.5 Hours of Monitoring with Herschel, 2016, arXiv:1605.05392v1
\bibitem[Haggard, Rovelli(2016)]{haggard}Haggard, H., M., Rovelli, C., Quantum Gravity Effects around Sagittarius A*, 2016, arXiv:1607.00364v2
\bibitem[Falke et al.(2000)]{falke2}Falke, J., Melia, F., Agol, E., 2000, ApJL, 528, L13
\bibitem[Rajwade et al.(2016)]{rajwade}Rajwade, K., M., Lorimer, D., R., Anderson, L., D., 2016, Detecting pulsars in the Galactic Centre, MNRAS, arXiv:1611.06977v1
\bibitem[Eisenhauer et al.(2005)]{eisenhower}Eisnenhower, F., et al, 2005, SINFONI in the Galactic Centre: Young Stars and Infrared Flares in the Central Light-Month, Apj, 628: 246-259
\end{thebibliography}
\end{document}

### 2017-01-16 ObsProjects/SgrA*.tex
Finished OA paper

@@
 \author{Howard Kinsman}
-\title{A Multiwavelength Report on Recent Findings on Sgr A*}
+\title{A Multiwavelength Report on Recent Findings and Ongoing Observations of Sgr A*}
 \begin{document}
 \maketitle
-\subsection{Introduction}
+\section{Introduction}
 Einstein's General Theory of Relativity (GR) predicts the existence of blackholes and the closest and best known candidate is Sgr A* at the centre of the Milky Way.
 It is believed to be a supermassive
-blackhole (SMBH) and lies at a distance of approximately 8 kpc with an estimated mass of $4\times10^6 M_{\odot}$. It's proximity therefore makes it ideal for the study of blackholes and as a 
-potential testbed for GR, as discussed later.
+blackhole (SMBH) and lies at a distance of approximately 8 kpc with an estimated mass of $4\times10^6 M_{\odot}$. Its proximity therefore makes it ideal for the study of blackholes and as a 
+potential testbed for GR, as discussed later in Section 6.
 It was first identified as a radio source \citep{balick} back in 1974. It has been studied using the entire electromagnetic spectrum although interstellar extinction makes some wavelengths more 
 problematic than others.
-Even though, by definition, black-holes are black, they do emit electromagnetic radiation from near the event horizon - as gas and plasma are accreted by the blackhole this material is heated up
+Even though, by definition, black-holes are black, they do emit electromagnetic radiation from near the event horizon - as gas and plasma are accreted by the blackhole, this material is heated up
 and large amounts of energy are emitted and radiated at all wavelengths of the electromagnetic spectrum. However since optical radiation is completely absorbed the best observing bands are radio 
-(including sub-mm), the NIR (including far infrared) and X-rays.
-The scale of a blackhole is set by the size of the Schwarzschild radius $R_s=2GM_{BH}/c^2\approx 3km (M_{bh}/M_{\odot})$. For Sgr A* this would lead to $\theta_{RS}=10\nu as$  however
-for Sgr A* the angular size, affected by gravitational lensing of its own potential, is predicted to be $\approx 10 \mu as$ making it the largest subtended angle blackhole anywhere in the universe and so
-is ideal for study. The distance to the Galactic Centre has been determined through high-resolution spectroscopy and the proper motion of stars giving a distance of $D=8.3(\pm 0.4) kpc$.
+(including sub-mm), the infrared (from near-infrared, NIR, to  far-infrared, FIR) and X-rays.
+
+The scale of a blackhole is set by the size of the Schwarzschild radius $R_s=2GM_{BH}/c^2\approx 3km (M_{bh}/M_{\odot})$. For Sgr A* this would lead to an angular size of $\theta_{RS}=10\mu as$ 
+ however the angular size is also affected by gravitational lensing due to its own gravitational potential, and is predicted to be $\approx 10 \mu as$ making it the largest 
+subtended angle blackhole anywhere in the universe and so is ideal for study. 
+The distance to the Galactic Centre has been determined through high-resolution spectroscopy and the proper motion of stars giving a distance of $D=8.3(\pm 0.4) kpc$.
+
 Apart from mass the other most important parameter of a blackhole is its accretion rate. This is determined by the distance from which it can accrete, the Bondi radius, given by
-$R_B=2GM/v_w^2$ where v is the speed of surrounding gas. Estimates for Sgr A* indicate $R_b\approx 2.5\times10^5 R_s \approx 0.1pc$ determined by radio polarization measurements.
-Sgr A* has extremely low levels of activity.
-Its current accretion rate is about four orders of magnitude lower than that required to for a SBMH to grow in a Hubble time \citep{falke1}, its radio emission
-is lower than expected and the amount of gas available as fuel is much lower that required to fuel an SMBH of that size. This has greatly constrained the models of Sgr A*.
+$R_B=2GM/v_w^2$ where v is the speed of surrounding gas. Estimates for Sgr A* indicate $R_b\approx 2.5\times10^5 R_s \approx 0.1pc$ which was determined by radio polarization measurements.
+Sgr A* shows suprisingly low levels of activity. Its current accretion rate is about four orders of magnitude lower than that required for a SBMH to grow in a Hubble time \citep{falke1}, 
+its radio emission is lower than expected and the amount of gas available as fuel is much lower that required to fuel an SMBH of that size. This has greatly constrained the models of Sgr A*.
+Conventional models of Sgr A* are based on radiatively inefficient accretion flow (RIAF) where the disk radiates inefficiently due to low particle density \citep{goddi}. However the latest models
+are based on general relativistic magneto-hydrodynamic (GRMHD) simulations \citep{goddi}, using a stationary rotating torus and the presence of a magnetic field. The BlackHoleCam collaboration
+\citep{goddi} has developed a Black Hole Accretion Code to further this area of study. As different models predict different results continued observations are necessary
+to understand the astrophysics and refine the models.
 
@@
 
-\subsection{Orbiting Stars and Mass of Sgr A* - IR}
+\section{Infrared - The Mass of Sgr A* and the Stars Orbiting Sgr A*}
 The first evidence for a central dark mass was suggested in 1980 through observations from the mid-infrared and radio recombination line observations \citep{falke1}. Although it wasn't until the 
 invention of adaptive optics for ground based telescopes that the resolution became sensitive enough to track stars orbiting this mass.
-Due to the high levels of exinction when viewing the galactic centre in the optical the NIR has been used to view stars orbiting Sgr A* for decades. There are numerous papers on this subject but this
-paragraph begins by citing from \citep{schodel} an early publication on this subject.
-This survey carried out ten years of high resolution NIR imaging and spectroscopy of the central region of the Milky Way in the
+Due to the high levels of exinction when viewing the galactic centre in the optical the NIR has been used to view stars orbiting Sgr A* now for decades.
+There are numerous papers on this subject but this
+section begins by citing from \citep{schodel}, an early publication on this subject.
+This survey carried out ten years of high resolution NIR imaging and spectroscopy of the central region of the Milky
+Way in the
 vicinity of Sgr A*. They used an 8m ground-based telescope to study the orbits of S1 and S2, the two closest stars to Sgr A*. S2, the star closest to Sgr A* at the time, was found to be in a
@@
 This survey combined a long time baseline with high astrometric accuracy using adaptive
-optics. This further refined the mass of Sgr A* and showed that a single point mass was the best fit model for this system, pointing again to the SMBH model - see \ref{fig:orbits}
+optics. This further refined the mass of Sgr A* and showed that a single point mass was the best fit model for this system, pointing again to the SMBH model - see Fig. \ref{fig:orbits} for a plot
+of the eight closest stars orbiting Sgr A*.
 
-Other evidence that points to the association dark mass wit Sgr A* is its own proper motion.
-Through VLBI it has been possible to track its motion over several years and its motion has been found to be
-entirely consistent with the motion of solar system around the Galactic Centre. This implies the radio source is actually the Galactic Centre.
+Other evidence that points to the association of a large dark mass with Sgr A* is its own proper motion. Through Very Long Baseline Interferometry (VLBI) it has been possible to track its motion over 
+several years and its motion has been found to be
+entirely consistent with the motion of the solar system around the Galactic Centre. This implies that the radio source is actually the Galactic Centre.
 
@@
 order of magnitude fainter than S2, and so much more difficult to track.
-The paper illustrates the use of a technique known as speckle holography which allows the speckle imaging data from 1995-2005 to be
+The paper illustrates the use of a technique known as speckle holography which allows the speckle imaging data from 1995-2005 to
+ be
 re-analysed and fainter stars such as S0-38 to be tracked. The study has put further constraints on both the mass and distance of Sgr A* and reduced the uncertainties to a confidence level of 99.7\%.
 The paper hopes that these improved constraints on the gravitational potential in the vicinity of Sgr A*, together with the addition of S0-38, will allow future tests of GR to be carried out when S2
-goes through its time of closest approach in 2018. One of the first relatavistic observations for GRAVITY, a second generation instrument on the Very Large Telescope Interferometer (VLTI), an adaptive
+goes through its time of closest approach in 2018.
+
+Observations of these stellar orbits is also ongoing in the radio wavelengths as well.
+One of the first relatavistic observations for GRAVITY, a second generation instrument on the Very Large Telescope
+ Interferometer (VLTI), an adaptive
 optics assisted interferometer, will be to observe the peri-astron shift of S2 in 2018. With a precision of about $10\mu$as GRAVITY will push interferometry accuracy way beyond what is possible today.
+GRAVITY combines the light from four telescopes and measures the interferograms from six baselines simultaneously. It operates in the K-band ($2.2 \mu m$). It is hoped that GRAVITY will finally
+prove that Sgr A* is a SMBH and resolve the debate as to whether NIR flares (discussed in Section 5) are individual hot spots close to the last stable orbit, from statistical fluctuations in the 
+accretion region, or from a jet. GRAVITY is also likely to be included in tests of GR in the strong field limit (Section 6).
 
@@
 
-\subsection{Radio}
-As mentioned previously blackholes do emit radiation from around the event horizon due to accretion.
-It has been calculated \citep{falke2} that the visual appearance of a blackhole would appear as a bright
+\section{Radio including Sub-mm - The Structure of Sgr A*}
+As mentioned previously blackholes do emit radiation from around the event horizon due to accretion. It has been calculated \citep{falke2} that the visual appearance of a blackhole would appear as a 
+bright
 photon ring with a dim "shadow" within its interior. Photons orbiting very close to the event horizon will disappear into the blackhole whilst those further away will escape to infinity therefore
 a clear boundary between light and dark regions is predicted.
-Gravitational lensing could increase the size of this shadow to bring it within limits VLBI measurements at mm wavelengths. 
-Radio emission from Sgr A* is believed to be produced the accretion onto the SMBH. At millimetre wavelengths, Very Long Baseline Interferometry (VLBI) can resolve the accretion region of the SMBH.
+Gravitational lensing could increase the size of this shadow to bring it to within the limits of Very Long Baseline Interferometry (VLBI) measurements at mm wavelengths. 
+Radio emission from Sgr A* is believed to be produced by the accretion of matter onto the SMBH. At millimetre wavelengths, VLBI can resolve the accretion region of the SMBH.
 It is believed that this radio emission is either synchotron radiation from a jet outflow, accretion flow onto the SMBH, or an isothermal jet coupled to an accretion flow \citep{ortiz}.
@@
 at shorter wavelengths this dependence is seen to diminish indicating the structure of Sgr A* may be revealed.
+The radio spectrum for Sgr A* exhibits a slow rise with frequency until it reaches the submm band where it peaks and then cuts off - this has been termed the "submm bump". It is generally thought that
+this is due to synchotron emission in transition from being optically thick to optically thin \citep{falke1}.
+
 In 2016 a study was carried out to attempt to identify the 2D structure of Sgr A* \citep{ortiz}. Up until this time existing VLBI baselines were limited in the north-south baselines however this study
-combined the National Radio Astronomy Observatory Very Long Baseline Array (VLBA) with the Large Millimieter Telescope Alfonso Serrano (LMT) to act a single VLBI array.
+combined the National Radio Astronomy Observatory Very Long Baseline Array (VLBA) with the Large Millimieter Telescope Alfonso Serrano (LMT) to act as a single VLBI array.
 This provided good north-south as
@@
 interferometry. Closure amplitudes are formed by combining complex ampltitudes in the correlated data from four different telescopes so that the telescope-based gain errors cancel \citep{falke1}.
- They also managed to account for some structure in the Sgr A8 image. The paper describes an Eastern secondary source separate from the primary source, although the authors do
+They also managed to account for some structure in the Sgr A* image. The paper describes an Eastern secondary source separate from the primary source, although the authors do
 note that this asymmetry may potentially be due to interstellar scattering rather than being intrinisic to the source.
-The Atacama Large Millimetre Array (ALMA) is the most sensitive (sub)mm-wave telescope ever builts and consists of 50 individual antennas of 12m diameter.
-Joint VLBI observations which include ALMA are planned
-for 2017 and should provide unprecedented sensitivity due to the improved north-south baseline.
-Given the overwhelming evidence that Sgr A* lies in the close vicinity to a SMBH there are a number of ongoing observations which aim to test for the existence of blackholes, test GR and study
-spacetime around blackholes \citep{goddi}. Also submm telescopes are combining around the world, icluding ALMA, to form the Event Horizon Telescope (EHT) which should provide unprecedented
-sensitivity for VLBI.
 
-The radio spectrum for Sgr A* exhibits a slow rise with frequency until it reaches the submm band where it peaks and then cuts off - this has been termed the "submm bump". It is generally thought that
-this is due to synchtron emission in transition from being optically thick to optically thin \citep{falke1}.
+The Atacama Large Millimetre Array (ALMA) is the most sensitive (sub)mm-wave telescope ever built and consists of 50 individual antennas of 12m diameter.
+Joint VLBI observations which include ALMA are
+planned
+for 2017 and should provide unprecedented sensitivity due to the improved north-south baseline. Another major development is that sub-mm telescopes are combining around the world, including ALMA, to 
+form the Event Horizon Telescope (EHT) which should 
+provide unrivaled precision for VLBI observations. It is hoped that EHT will have the necessary precision to directly observe the blackhole's "shadow". EHT will also form one of the experiments
+that tests GR in a strong field (Section 6). EHT aims to improve our understanding of accretion around a blackhole and the physics behind blackhole jets.
 
-\subsection{G2 and Gamma}
+\section{Gamma Rays and the G2 object}
 A notable, and widely publicised, event occurred in the vicinity of Sgr A* during 2013-14. It was reported \citep{gillessen2} that a gas cloud, now known as G2, of the order three Earth masses, was on
@@
 This sparked interest in the scientific community as it was assumed that in-falling matter into the SMBH could accelerate
-high-energy particles resulting emission at various wavelengths, including gamma rays. A four year observational campaign was therefore started by using the Major Atmospheric Gamma Imaging Cherenkov
-(MAGIC) telescopes \citep{ahnen}. These two 17m diameter telescopes record flashes of Cherenkov light produced the Extensive Air Showers which occurr in the upper atmosphere and are caused by very
+high-energy particles resulting in emission at various wavelengths, including gamma rays.
+A four year observational campaign was therefore started by using the Major Atmospheric Gamma Imaging Cherenkov
+(MAGIC) telescopes \citep{ahnen}. These two 17m diameter telescopes record flashes of Cherenkov light produced by Extensive Air Showers which occurr in the upper atmosphere and are caused by very
 high energy $\gamma$ ray photons with energies $>50GeV$. The results however were disappointing and the close proximity of G2 to the SMBH had no measureable effect on $\gamma$ ray emissions, although
-the authors hope this study will act as a baseline for monitoring any flaring activity in the future. These results, however, are not surpising as it had already been reported \cite{ghez} that G2
+the authors hope this study will act as a baseline for monitoring any flaring activity in the future. These results, however, are not surpising as it had already been reported \citep{ghez} that G2
 had passed by Sgr A* largely unaffected when monitored using NIR.
 
-\subsection{X-ray}
-Since the launch of Chandra in 1999 it was found that Sqr A* is a source of suprisingly weak X-rays \citep{falke1}.
-
-\subsection{Multiwavelength}
-The flaring activity associated with Sgr A* are also widely studied and observed phenomena.
-These can be observed using infrared, X-ray and submillimetre wavelengths.
-The origin of these flares is still
+\section{X-rays, Flares and Multiwavelength Surveys}
+Since the launch of Chandra in 1999 it was found that Sqr A* is a source of suprisingly weak X-rays \citep{falke1}, however Sgr A* does exhibit periods of higher X-ray emission during flaring periods.
+These occur on a much faster timescale than the steady emission of X-rays.
+The flaring activity associated with Sgr A* are widely studied and observed phenomena. These can be observed using infrared, X-ray and submillimetre wavelengths. The origin of these flares is still
 being debated but to understand this activity it is important to compare results from all of these wavelengths. A multiwavelength campaign \citep{mossoux} was therefore carried out in 2014, also timed
 to be carried out at G2's closest approach.
-This ambitious study aimed to make simultaneous observations using X-Ray Multi-Mirror Mission (XMM-Newton) and Wide Field Camera 3 (WFC3) onboard the Hubble Space Telescope (HST) giving observations in X-Ray and
+This ambitious study aimed to make simultaneous observations using X-Ray Multi-Mirror Mission (XMM-Newton) and Wide Field Camera 3 (WFC3) onboard the Hubble Space Telescope (HST) giving observations in
+X-Ray 
+and
 NIR respectively. NIR observations were also carried out using the Spectrograph for Integral Field Observations in the Near Infrared (SINFONI) at the VLT in Chile. 
 Radio observations were also carried using the Karl G. Jansky Very Large Array (VLA) in the X band (3.75-2.5cm) 
-and also using the Combined Array for Research in Millimeter-wave Astronomy (CARMA) at 3.2mm.
-There was no increased flaring activity detected that could be attributed to the passage of G2 however a number of flares
+and also using the Combined Array for Research in Millimeter-wave Astronomy (CARMA) at 3.2mm. There was no increased flaring activity detected that could be attributed to the passage of G2 however a 
+number of
+ flares
 were detected at the various wavelengths allowing the data to be analysed and compared using a multiwavelength approach. 
 In total three flares were detected by the WCF3 on board HST plus two more using SINFONIA on the VLT.
-These were very luminous NIR flares but given that SINFONIA and WFC3 have very high detection limits this
+These were very luminous NIR flares but given that SINFONIA and WFC3 have very high detection limits
+this
 is to be expected statistically. Two X-ray flares were detected by XMM-Newton and were two of the brightest flares ever detected by XMM-Newton. Two of the NIR flares had definite X-ray counterparts
@@
 'bump' but this could not be correlated with any of the NIR/X-ray flares. 
-The three main explainations for these flares are: a jet with clumps of ejected material, short-lived "hot spots" orbiting the blackhole or statistical fluctuations in the accretion flow \citep{goddi}.
+The three main hypothoses for these flares are: a jet with clumps of ejected material, short-lived "hot spots" orbiting the blackhole or statistical fluctuations in the accretion flow \citep{goddi}.
 With its $10\mu$as precision GRAVITY may also soon be able to determine the nature of these flares and settle this debate. GRAVITY should also have the precision to detect Lense-Thirring precession
-for stars orbiting the SMBH and hopefully constrain the blackhole spin from stellar orbits.
+for stars orbiting the SMBH and hopefully put constraints on the blackhole spin from these stellar orbits.
 
-In a separate study \citep{stone} the Herschel Space Observatory's (no longer operational) Spectral and Photometric Imaging Receiver (SPIRE) was used to carry out the longest ever continuous submillimetre
+In a separate study \citep{stone} the Herschel Space Observatory's (no longer operational) Spectral and Photometric Imaging Receiver (SPIRE) was used to carry out the longest ever continuous 
+submillimetre 
 survey of Sgr A*. These were observed in three FIR bands: 0.25, 0.35 and 0.5mm. These wavelength are difficult, if not impossible, to observe from the ground. The aim of the study
-was to test for the variability of luminosity of the accretion flow into SMBH.
-This was then compared with ground-based observations at the Caltech Submillimetre Observatory using 0.85mm and X-ray data from XMM-Newton. Small (<1\%) variations were seen at the three wavelengths
-used by SPIRE and comparison with 0.5mm ground data support these small variations. However comparison XMM-Newton didn't reveal any corresponding variations.
+was to test for the variability of luminosity of the accretion flow into the SMBH in order to constrain low-luminosity accretion models.
+This was then compared with ground-based observations at the Caltech Submillimetre Observatory using 0.85mm and X-ray data from XMM-Newton. 
+Small ($<1\%$) variations were seen at the three wavelengths
+used by SPIRE and comparison with 0.85mm ground data support these small variations. However comparison XMM-Newton didn't reveal any corresponding variations.
 
-\subsection{Pulsars}
+\section{GR Tests and Pulsars}
 Given the overwhelming evidence that Sgr A* lies in the close vicinity to a SMBH there are a number of ongoing observations which aim to test for the existence of blackholes, test GR and study 
 spacetime around blackholes \citep{goddi}.
-Some very precise tests of GR have been carried out using pulsars: a pulsar in a binary can be used to probe the spacetime around that system by comparing the
+Some very precise tests of GR have been carried out using pulsars: a pulsar in a binary system can be used to probe the spacetime around that system by comparing the
 "pulsar clock" with the radio telescope maser. However it is believed that a pulsar orbiting a blackhole could provide an even more accurate test. No such system has yet been found however it is 
@@
 The huge mass of SMBH would provide a very accurate test of GR.
-Unfortunately only five pulsars within 0.5\degree of Sgr A* have so far been found,
-the lack of which
-has been attributed to interstellar scattering.
-However in 2013 radio emission from a magnetar (PSR J1745-2900) was detected supporting the hypothesis that many more pulsars may yet still be found.
-It is
-hoped that ALMA will help achieve this aim.
-Through the use of modelling \citep{rajwade} state that it is not surpising current surveys have failed to find many pulsars in the Galactic Centre has they have only probed $<2\%$ of the population.
-The authors state that upcoming telescopes such as the Square Kilometre Array would probe deepper and survey a more sizeable population of the pulsars in the Galactic Centre.
+Unfortunately only five pulsars \citep{goddi} within 0.5\degree of Sgr A* have so far
+ been found,
+the lack of which has been attributed to interstellar scattering. 
+However in 2013 radio emission from a magnetar (PSR J1745-2900) was detected supporting the hypothesis that many more pulsars may
+ yet still be found. It is hoped that ALMA will help achieve this aim.
+\cite{rajwade} modelled both canonical pulsars (CPs) and millisecond pulars (MSPs) at the Galactic Centre and found that the optimum frequencies for future studies would be 9 GHz for CPs and 22 GHz
+for MSPs. They state that it is not surpising current surveys have failed to find many pulsars in the Galactic Centre has they have only probed $<2\%$ of the population.
+The authors hope that upcoming telescopes such as the Square Kilometre Array would probe deeper and survey a more sizeable population of the pulsars in the Galactic Centre.
+Next generation instruments, such as LMT, and ALMA should allow for the first systematic search for pulsars to be carried out at 90 GHz, or even higher \citep{goddi}.
 
-\subsection{Summary}
-This report has so far mentioned three ongoing projects to test GR: the blackhole shadow using EHT, stars orbiting Sgr A* using GRAVITY and pulsars with ALMA. Each of these techniques uses utilises
-different wavelengths and observational techniques: mm-VLBI, astrometry with NIR and pulsar timing with radio. It is hoped the combination of these results will test the validity of GR with
+This report has so far mentioned three ongoing projects to test GR: the blackhole shadow using EHT, stars orbiting Sgr A* using GRAVITY and pulsars with ALMA. Each of these techniques utilises
+different wavelengths and observational techniques: mm-VLBI, astrometry with NIR and pulsar timing with radio.
+GR has so far only been tested in the weak field limit and in the strong limit using pulsar binary systems.
+Sgr A* provides a suitable testbed for a comprehensive test of GR in the strong field limit.
+In order to do this the results from these
+different observations will need to be combined \citep{goddi}. Each of the above measurements uses a different observational techniques and are therefore subject to different kinds of errors. By 
+combining these results any uncertainties could be reduced. Also measurements of the blackhole mass and spin measured independently using different techniques could be compared to each other 
+increasing the precision of these measurements.
+
+\section{Summary}
+SMBHs provide us with the strongest gravitational fields with which to perform comprehensive GR tests in the strong field limit and Sgr A* is by far the closest SMBH. Three experiments are currently
+ongoing to test GR. The first is to identify the structure of Sgr A* and obtain an image of the "shadow" of the blackhole. EHT will be an almost Earth sized telescope operating using millimetre VLBI
+and should have the resolution to obtain an image of this "shadow". The second experiment is continued monitoring of stellar orbits and GRAVITY will have the precision to identify relativistic 
+precession of these orbits. Finally the third ongoing experiment is the continued search for pulsars in the Galactic Centre. It is hoped the combination of these results will test the validity of 
+GR with
 more accuracy than ever before. The increasing precision of VLBI observations has prompted some authors \citep{haggard} to suggest that we are approaching the observational limit to where quantum 
@@
 \begin{thebibliography}{1}
+\bibitem[Ahnen et al.(2016)]{ahnen}Ahnen, M.,L., Ansoldi, S., Antonelli, L., A., Antoranz, P., Arcaro, C., et al., 2016, Observations of Sagittarius A* during the pericenter passage of the G2 object with
+MAGIC, A\&A, arXiv:1611.07095v1
 \bibitem[Balick et al.(1974)]{balick}Balick, B., Brown R. L., (1974), ApJ, 194, 265
-\bibitem[Schodel et al.(2002)]{schodel}Schodel, R., Ott, T., Genzel, R., Hofmann, R., Lehnert, M., Eckart, A., et al., 2002, Closest Star Seen Orbiting the Supermassive Black Hole at the Centre of the Milky Way,
-arXiv:astro-ph/0210426v1
-\bibitem[Goddi et al.(2016)]{goddi}Goddi, C., Falcke, H., Rezzolla, L., Brinkerink, C. et al., BlackHoleCam: fundemental physics of the Galactic center, arXiv:1606.08879c1, WSPC Proceedings, 2016 
 \bibitem[Boehle et al.(2016)]{boehle}Boehle, A., Ghez, A., M., Schodel, R., Meyer, L. Yelda, S. et al, 2016, An Improved Distance and Mass Estimate for Sgr A* from a Multistar Orbit Analysis, 
 arXiv:1607.05726v1
-\bibitem[Gillessen et al.(2009)]{gillessen1}Gillessen, S., Eisenhauer, F., Trippe S., et al., 2008, Monitoring Stellar Orbits Around the Massive Black Hole in the Galactic Center, The Astrophysical Journal,
-692 (2): 1075-1109, arXiv:0910.3069v1
-\bibitem[Ortiz-Leon et al.(2016)]{ortiz}Ortiz-Leon, G., N., Johnson, M., D., Doelman, S.,S., Blackburn, L., et al., 2016, The Intrinsic Shape of Sagittarius A* at 3.5mm Wavelength, arXiv:1601.06571v2
+\bibitem[Brinkerink et al.(2016)]{brinkerink}Brinkerink, C., D., Muller, C., Falcke, H., Bower, G., C., et al., 2016, Asymmetric structure in Sgr A* at 3mm from closure phase measurements with VLBA,
+ GBT and LMT, MNRAS, arXiv:1608.0651v1
+\bibitem[Eisenhauer et al.(2005)]{eisenhower}Eisnenhower, F., et al, 2005, SINFONI in the Galactic Centre: Young Stars and Infrared Flares in the Central Light-Month, Apj, 628: 246-259
+\bibitem[Falke et al.(2000)]{falke2}Falke, J., Melia, F., Agol, E., 2000, ApJL, 528, L13
 \bibitem[Falke, Markoff(2013)]{falke1}Falke, H., Markoff, S. B., 2013, Towards the event horizon - the supermassive black hole in the Galactic Center, arXiv:1311.184v1
-\bibitem[Brinkerink et al.(2016)]{brinkerink}Brinkerink, C., D., Muller, C., Falcke, H., Bower, G., C., et al., 2016, Asymmetric structure in Sgr A* at 3mm from closure phase measurements with VLBA, GBT and LMT,
-MNRAS, arXiv:1608.0651v1
+\bibitem[Ghez et al.(2014)]{ghez}Ghez, A., M., Witzel, G., Sitarski, B., et al, 2014, The Astronomer's Telegram, 6110, 1
+\bibitem[Gillessen et al.(2009)]{gillessen1}Gillessen, S., Eisenhauer, F., Trippe S., et al., 2008, Monitoring Stellar Orbits Around the Massive Black Hole in the Galactic Center, The Astrophysical
+ Journal, 692 (2): 1075-1109, arXiv:0910.3069v1
 \bibitem[Gillessen et al.(2012)]{gillessen2}Gillessen, S., Genzel, R., Fritz, T., et al., 2012, Nature, 481, 51
-\bibitem[Ahnen et al.(2016)]{ahnen}Ahnen, M.,L., Ansoldi, S., Antonelli, L., A., Antoranz, P., Arcaro, C., et al., 2016, Observations of Sagittarius A* during the pericenter passage of the G2 object with MAGIC, A\&A, arXiv:1611.07095v1
-\bibitem[Ghez et al.(2014)]{ghez}Ghez, A., M., Witzel, G., Sitarski, B., et al, 2014, The Astronomer's Telegram, 6110, 1
+\bibitem[Goddi et al.(2016)]{goddi}Goddi, C., Falcke, H., Rezzolla, L., Brinkerink, C. et al., BlackHoleCam: fundemental physics of the Galactic center, arXiv:1606.08879c1, WSPC Proceedings, 2016 
+\bibitem[Haggard, Rovelli(2016)]{haggard}Haggard, H., M., Rovelli, C., Quantum Gravity Effects around Sagittarius A*, 2016, arXiv:1607.00364v2
 \bibitem[Mossoux et al.(2016)]{mossoux}Mossoux, E., Grosso, N., Bushouse, H., Eckart, A., Yusef-Zadeh, F., et al., Multiwavelngth study of the flaring activity of Sgr A* in 2014 February-April, 2016,
 A\&A, ArXiv:1603.01048v1
-\bibitem[Stone et al.(2016)]{stone}Stone, J., M., Marrone, D., P., Dowell, C., D., Schulz, B., et al., Far Infrared Variability of Sagittarius A*: 25.5 Hours of Monitoring with Herschel, 2016, arXiv:1605.05392v1
-\bibitem[Haggard, Rovelli(2016)]{haggard}Haggard, H., M., Rovelli, C., Quantum Gravity Effects around Sagittarius A*, 2016, arXiv:1607.00364v2
-\bibitem[Falke et al.(2000)]{falke2}Falke, J., Melia, F., Agol, E., 2000, ApJL, 528, L13
+\bibitem[Ortiz-Leon et al.(2016)]{ortiz}Ortiz-Leon, G., N., Johnson, M., D., Doelman, S.,S., Blackburn, L., et al., 2016, The Intrinsic Shape of Sagittarius A* at 3.5mm Wavelength, arXiv:1601.06571v2
 \bibitem[Rajwade et al.(2016)]{rajwade}Rajwade, K., M., Lorimer, D., R., Anderson, L., D., 2016, Detecting pulsars in the Galactic Centre, MNRAS, arXiv:1611.06977v1
-\bibitem[Eisenhauer et al.(2005)]{eisenhower}Eisnenhower, F., et al, 2005, SINFONI in the Galactic Centre: Young Stars and Infrared Flares in the Central Light-Month, Apj, 628: 246-259
+\bibitem[Schodel et al.(2002)]{schodel}Schodel, R., Ott, T., Genzel, R., Hofmann, R., Lehnert, M., Eckart, A., et al., 2002, Closest Star Seen Orbiting the Supermassive Black Hole at the Centre of the
+Milky Way, arXiv:astro-ph/0210426v1
+\bibitem[Stone et al.(2016)]{stone}Stone, J., M., Marrone, D., P., Dowell, C., D., Schulz, B., et al., Far Infrared Variability of Sagittarius A*: 25.5 Hours of Monitoring with Herschel, 2016,
+ arXiv:1605.05392v1
 \end{thebibliography}

### 2017-01-23 workspace/DoubleBars/MScProjectDoubleBars.tex
New plots

@@
 \usepackage{listings}
+\usepackage[skip=0pt]{caption}
 \graphicspath{{./}}
 \DeclareGraphicsExtensions{.eps}
+%\setlength\abovecaptionskip{-5pt}
 \author{Howard Kinsman}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/outerbinary/Orbit.eps}
+\includegraphics[width=.9\textwidth]{./results/outerbinary/Orbit.eps}
 \caption{Outer Binary}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/innerbinary/Orbit.eps}
+\includegraphics[width=.9\textwidth]{./results/innerbinary/Orbit.eps}
 \caption{Inner Binary}
@@
 As can be seen from Fig. \ref{fig:config1} even a neglible mass inner binary causes the outer binary to begin to separate and become less tightly bound.
-Figures Fig. \ref{fig:config1} to \ref{fig:config18} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
+Figures Fig. \ref{fig:config1} to \ref{fig:config18i} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
 In many cases the inner binary is only visible as a point at the centre of the outer binary.
-As the mass of the inner binary increased the system became increasingly unstable and the effect on the outer binary became more pronounced.
+As the mass of the inner binary increased the system became increasingly unstablle and the effect on the outer binary became more pronounced.
 Even the 'stable' configurations listed in Table \ref{tab:variables} hold for only a few orbits.
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/stablebase/Orbit.eps}
 \caption{Configuration 1}
+\includegraphics[width=.9\textwidth]{./results/stablebase/Orbit.eps}
 \label{fig:config1}
 \end{figure}
+
 \begin{figure}[H]
 \centering
-\includegraphics[width=0.75\textwidth]{./results/002-5-001/Orbit.eps}
+\caption{Configuration 1 - Inner Bar}
+\includegraphics[width=.9\textwidth]{./results/stablebase/Inner.eps}
+\label{fig:config1i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
 \caption{Configuration 2}
+\includegraphics[width=.9\textwidth]{./results/002-5-001/Orbit.eps}
 \label{fig:config2}
 \end{figure}
+
 \begin{figure}[H]
 \centering
-\includegraphics[width=0.75\textwidth]{./results/003-5-001/Orbit.eps}
+\caption{Configuration 2 - Inner Bar}
+\includegraphics[width=.9\textwidth]{./results/002-5-001/Inner.eps}
+\label{fig:config2i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=.9\textwidth]{./results/003-5-001/Orbit.eps}
 \caption{Configuration 3}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/004-5-004/Orbit.eps}
+\includegraphics[width=.9\textwidth]{./results/003-5-001/Inner.eps}
+\caption{Configuration 3 - Inner Bar}
+\label{fig:config3i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/004-5-004/Orbit.eps}
 \caption{Configuration 4}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/004-55-004/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/004-5-004/Inner.eps}
+\caption{Configuration 4 - Inner Bar}
+\label{fig:config4i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/004-55-004/Orbit.eps}
 \caption{Configuration 5}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/004-57-004/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/004-55-004/Inner.eps}
+\caption{Configuration 5 - Inner Bar}
+\label{fig:config5i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/004-57-004/Orbit.eps}
 \caption{Configuration 6}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/005-58-005-3/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/004-57-004/Inner.eps}
+\caption{Configuration 6 - Inner Bar}
+\label{fig:config6i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/005-58-005-3/Orbit.eps}
 \caption{Configuration 7}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/005-58-005-4/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/005-58-005-3/Inner.eps}
+\caption{Configuration 7 - Inner Bar}
+\label{fig:config7i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/005-58-005-4/Orbit.eps}
 \caption{Configuration 8}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/005-58-005-35/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/005-58-005-4/Inner.eps}
+\caption{Configuration 8 - Inner Bar}
+\label{fig:config8i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/005-58-005-35/Orbit.eps}
 \caption{Configuration 9}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/006-6-006-3/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/005-58-005-35/Inner.eps}
+\caption{Configuration 9 - Inner Bar}
+\label{fig:config9i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/006-6-006-3/Orbit.eps}
 \caption{Configuration 10}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/01-7-01-2/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/006-6-006-3/Inner.eps}
+\caption{Configuration 10 - Inner Bar}
+\label{fig:config10i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/01-7-01-2/Orbit.eps}
 \caption{Configuration 11}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/02-75-02-1/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/01-7-01-2/Inner.eps}
+\caption{Configuration 11 - Inner Bar}
+\label{fig:config11i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/02-75-02-1/Orbit.eps}
 \caption{Configuration 12}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/025-65-02/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/02-75-02-1/Inner.eps}
+\caption{Configuration 12 - Inner Bar}
+\label{fig:config12i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/025-65-02/Orbit.eps}
 \caption{Configuration 13}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/05-9-05-1/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/025-65-02/Inner.eps}
+\caption{Configuration 13 - Inner Bar}
+\label{fig:config13i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/05-9-05-1/Orbit.eps}
 \caption{Configuration 14}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/05-9-05-15/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/05-9-05-1/Inner.eps}
+\caption{Configuration 14 - Inner Bar}
+\label{fig:config14i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/05-9-05-15/Orbit.eps}
 \caption{Configuration 15}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/06-95-06-15/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/05-9-05-15/Inner.eps}
+\caption{Configuration 15 - Inner Bar}
+\label{fig:config15i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/06-95-06-15/Orbit.eps}
 \caption{Configuration 16}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/07-1-07-15/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/06-95-06-15/Inner.eps}
+\caption{Configuration 16 - Inner Bar}
+\label{fig:config16i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/07-1-07-15/Orbit.eps}
 \caption{Configuration 17}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./results/08-105-08-12/Orbit.eps}
+\includegraphics[width=0.9\textwidth]{./results/07-1-07-15/Inner.eps}
+\caption{Configuration 17 - Inner Bar}
+\label{fig:config17i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/08-105-08-12/Orbit.eps}
 \caption{Configuration 18}
@@
 \end{figure}
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./results/08-105-08-12/Inner.eps}
+\caption{Configuration 18 - Inner Bar}
+\label{fig:config18i}
+\end{figure}
 \end{document}

### 2017-01-23 workspace/DoubleBars/DoubleBars_v2.f90

@@ -21,9 +21,9 @@
 
    subroutine initialconditions
 
-     step = 1.e-5 !vary step size
+     step = 1.e-4 !vary step size
      start=0.
-     end=50 !vary end time
+     end=20 !vary end time
 !     m1 = 1
 !     m2 = .05
 !     m3 = .000005
@@ -32,18 +32,18 @@
 !      m2=.5
 		m1=.75   !blue
 		m2=.25  !red
-      m3=0!.001    !black
-      m4=0!.001   !orange
+      m3=.001    !black
+      m4=.001   !orange
 
-     ys(1) = -.5     ! m1,x
+     ys(1) = -.25     ! m1,x
      ys(2) = 0  ! m1,vx
      ys(3) = 0       ! m1,y
-     ys(4) = -.5		!m1,vy
+     ys(4) = -.25		!m1,vy
      
-     ys(5) = .5         ! m2,x
+     ys(5) = .75         ! m2,x
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
-     ys(8) = .5		!m2,vy
+     ys(8) = .75		!m2,vy
 
      ys(9) = -.001     ! m3,x
      ys(10) = 0       ! m3,vx

### 2017-04-25 cosmology/Tutorial1/Tutorial1.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{Cosmology Tutorial 1}
\begin{document}
\maketitle

\section{}
\begin{equation}
f=\frac{L}{4\pi r^2}
\end{equation}
and differentiating
\begin{equation}
\frac{df}{dr}=-\frac{L}{2\pi r^3}
\end{equation}
Number density = $\frac{1}{r}$ so
\begin{equation}
N=\frac{4}{3}\pi r^3 \times \frac{1}{r}
\end{equation}
\begin{equation}
N=\frac{4\pi r^2}{3}
\end{equation}
and differentiating
\begin{equation}
\frac{dN}{dr}=\frac{8\pi r}{L}
\end{equation}
dividing equation 5 by equation 2 gives
\begin{equation}
\frac{dN}{df}=\frac{-48\pi^2 r^4}{L}
\end{equation}
so that
\begin{equation}
\frac{dN}{df}\propto r^4
\end{equation}
$r^2\propto f^{-1}$ so that
\begin{equation}
\frac{dN}{df}\propto f^{-2}
\end{equation}
Therefore $\alpha = -2$. 
$dln f=\frac{1}{f}df$ so that
\begin{equation}
\mathcal{N}=\frac{dN}{dln f}=f\frac{dN}{df}
\end{equation}
which gives $\mathcal{N}\propto f^{-1}$ and so
\begin{equation}
log\mathcal{N}=-log f + const
\end{equation}
This is linear so measuring stellar fluxes would allow an observer to determine if the distribution was homogeneous.

\section{}
The NRAO VLA Sky Survey (NVSS) is a radio survey covering the sky north of -40° declination at 1.4GHz. It contains over a million 
rows so I extracted a subset of sources where the flux density $>100mJy$ giving appoximately 62,000 sources.

The Two Micron All Sky Survey (2MASS) Redshift Survey (2MRS) was a ten-year project to map the full three-dimensional distribution of galaxies in
the nearby universe. It provides a good census of baryonic matter in the nearby universe (to about 300 Mpc). It contains
about 400,000 rows so I only extracted sources with a magnitude $<13$ in the K band giving about 44,000 rows.

The 2MASS Extended Source Catalog (2MASX) contains over 1.6 millions rows and suveys the entire sky in the J, H and K bands.
I restricted the output to sources with a magnitude of $<13$ in the J band giving about 77,000 rows.

\section{}
Figure 1 shows a log of number counts per log flux for the NVSS. The slope was calculated:
\begin{equation}
\frac{log(2000/1000)}{log(100-200)}=-1
\end{equation}
This figure matches quite well with the figure of -1.5 calculated for a homogeneous universe. The slope is constant apart
from very small flux densities (probably due to sampling errors) and large flux densities (the universe is not
homogeneous on smaller scales).
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./NVSS.eps}
\caption{NVSS}
\label{fig:1}
\end{figure}
Figure 2 shows a log of number counts per magnitude in the K band for the 2MRS survey.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2MRS.eps}
\caption{2MRS}
\label{fig:2}
\end{figure}
The slope was calculated as:
\begin{equation}
\frac{log(1000/100)}{10/7}=.8
\end{equation}
This also reasonably matches the expected slope of .6 for a homogeneous universe.
Finally figure 3 shows a log of number counts per magnitude in the J band for the 2MASX survey.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2MASX.eps}
\caption{2MASX}
\label{fig:3}
\end{figure}
The slope was calculated as:
\begin{equation}
\frac{log(10000/1000)}{12/11}=.9
\end{equation}
This is a similar figure to the 2MRS survey.

\section{}
Figure 4 shows the 2MRS data in galactic coordinates restricted to the the closest half of the galaxies i.e. $cz<8000$ km/s.
It clearly shows evidence of structure indicating that the universe is not homogeneous on smaller scales.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./NearestGalaxies.eps}
\caption{2MRS}
\label{fig:4}
\end{figure}

%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-04-25 cosmology/Tutorial10/Tutorial10.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{gensymb}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Tutorial 10}
\begin{document}
\maketitle
\section{}
The Sloan Digital Sky Survey (SDSS) saw first light in 1998 and its scope has changed over the years. Current surveys (2014-2020) include:
\newline
eBOSS - Extended Baryon Oscillation Spectroscopic Survey - which is a spectroscopic survey of galaxies and quasars up to redshift z=3.
\newline
APOGEE-2 - APO Galaxy Evolution Experiment - infrared spectroscopic suvery of Milky Way.
\newline
MaNGA - Obtaining spectra of entire face of targetted local galaxies.

The general goals of SDSS are to use highly efficient instruments and software to create high quality surveys and public datasets and to make use
of international collaboration.
\section{} 
The ALFALFA survey is a HI survey and so maps extraglactic HI sources and then tries to map these to optical sources. It differs from other surveys in that
it detects HI line sources.

Figure 1 shows a histogram of the HI flux density, and shows the total survey data together with flux limited and volume limited samples. I chose 100 Jy  km/s
as the flux limit.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./SiH1.jpg}
\caption{}
\label{fig:1}
\end{figure}
Figure 2 shows mass plotted with distance and also shows the volume limited sample from 80 to 100 Mpc. I have to admit I was unsure of how to select a
volume limited sample and this step seems very arbitrary - so probably my lack of understanding.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./MassDistance.jpg}
\caption{}
\label{fig:2}
\end{figure}
Figure 3 is a histogram of the mass from the volume limited sample and shows the mass function.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./MassFunction.jpg}
\caption{}
\label{fig:3}
\end{figure}
The survey area of ALFALFA was 6300 square degrees which is 1.92 steradians so,
\begin{flalign*}
& V=\frac{1}{3}\Omega\left[D(z_{max})^3-D(z_{min})^3\right] &\\
& = \frac{1}{3}\times1.92\times\left(100^3-80^3\right) &\\
& =312,320 Mpc^{-3}
\end{flalign*}
I think I was supposed to use comoving distance for the above calculation but don't know how to get from data given.
\section{}
I couldn't do question 3.
\section{}
\begin{flalign*}
& PSFR(z)=.015\frac{\left(1+z\right)^{2.7}}{1+\left[\left(1+z\right)/2.9\right]^{5.6}} &\\
& =.015\frac{4^{2.7}}{1+\left(4/2.9\right)^{5.6}} &\\
& =.0898M_{\odot}yr^{-1}Mpc^{-3}
\end{flalign*}
Comvoing volume is used in order to compare the SFR density of galaxies at different redshifts i.e. the expansion
of the universe needs to be accounted for.
\section{}
Passive galaxies tend to occupy virialised (high-density) clusters and so have higher velocity dispersions.
From Madgwick et al., 2003, The 2dF Galaxy Redshift Survey: galaxy clustering per spectral type.
\section{}
Solid line is CDM as it dominates.
\newline
Dashed line is baryonic matter as it shows the BAO peak at around 100 Mpc.
\newline
I don't know what the dotted line is - Neutrinos?
\section{}
An additional neutrino species would increase the horizon scale at the time of photon-baryobn decoupling.
This would lead to values of $H_0$ that are in agreement with other measurements i.e. the distance ladder
and inverse distance ladder would agree.
%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-04-25 cosmology/Tutorial11/Tutorial11.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{gensymb}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Tutorial 11}
\begin{document}
\maketitle
\section{}
 \begin{flalign*}
& M=V\rho &\\
& V=\frac{4}{3}\pi r^3 &\\
& r=\left(\frac{M}{\frac{4}{3}\pi\rho}\right)^{\frac{1}{3}} &\\
& r=\left(\frac{2\times10^{45}}{\frac{4}{3}\pi 10^{-26}}\right)^{\frac{1}{3}} &\\
& r=6.2\times10^{22}m &\\
& = 11.7 Mpc
\end{flalign*}
\section{}
\begin{flalign*}
& R_{200,m}=\left(\frac{2\times10^{45}}{\frac{4}{3}\pi 2\times10^{-24}}\right)^{\frac{1}{3}} &\\
& =6.2\times10^{22} m &\\
& =2Mpc
\end{flalign*}
\begin{flalign*}
& R_{200,m}=\left(\frac{2\times10^{44}}{\frac{4}{3}\pi 2\times10^{-24}}\right)^{\frac{1}{3}} &\\
& =2.88\times10^{22} m &\\
& =0.9Mpc
\end{flalign*}
\section{}
The four main ways to detect galaxy clusters are:
\newline
X-ray emission from hot gas - space X-Ray telescopes e.g. Chandra. Disadvantage of this method is that it is distance dependent, so better for closer galaxies.
\newline
Sunyaev-Zel'dovich effect - galaxy clusters cause inverse Compton scattering of CMB photons. Advantage of this method is that it is independent of redshift and so
best for more distant galaxies.
\newline
Optical - galaxy clusters can be observed directly in the optical. Disadvantage of this is that there may be galaxies within the line of sight which are not part of the cluster and
have to be removed. Also distance dependent.
\newline
Gravitational lensing - galaxy clusters bend light from galaxies behind the cluster - both strong and weak lensing effects can be used. Also distance dependent i.e. harder to observe the more distant
a galaxy cluster is.
%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-04-25 cosmology/Tutorial2/Tutorial2.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{Cosmology Tutorial 2}
\begin{document}
\maketitle

\section{}
Gravitational mass is the mass in Newton's inverse square law i.e. $F=\frac{GMM}{r^2}$ (either the active or passive mass
under the influence of a gravitational field) whilst inertial mass is the mass in
Newton's 2nd law of motion ie. $F=Ma$ (the mass which resists motion). In Newtonian mechanics there is no obvious reason
or explanation as to why these are the same thing.
Two other problems with Newtonian mechanics are the difference in the predicted and observed precession of the 
perihelion of Mercury according to classical mechanics; and, as Newtonian mechanics rely on inertial frames, there is no
simple definition as to what an inertial frame is within the context of Newton's equations.

\section{}
Starting with energy $E=mc^2$ an object falling in a gravitational field will lose potential energy $E_p=mgh$ but gain
the opposite in kinetic energy so will now have energy $E=mc^2+mgh$. Converting this into photon energy and letting photon
rise back in gravitational field it would obviously have gained energy - not possible according to conservation of energy.
So redshift is given by
\begin{equation}
z=\frac{\delta\lambda}{\lambda}
\end{equation}
and can also be written as 
\begin{equation}
1+z=\frac{\lambda'}{\lambda}
\end{equation}
reversing the fraction now (as the photon is rising) the above is equivalent to
\begin{equation}
1+z=\frac{\nu'}{\nu}
\end{equation}
and as $E=h\nu$ we get
\begin{equation}
1+z=\frac{E'}{E}
\end{equation}
or
\begin{equation}
1+z=\frac{mc^2+mgh}{mc^2}
\end{equation} 
giving 
\begin{equation}
1+z=1+\frac{gh}{c^2}
\end{equation}
and finally 
\begin{equation}
z=\frac{gh}{c^2}
\end{equation}
It is the same result as the Equivalence Principle.

\section{}
Curved space is a universe where space (and time) deviates from a Euclidean metric. There are two main types of curvature for
a homogeneous and isotropic universe: positive curvature or spherical geometry e.g. a sphere where $C<2\pi a$ and $k>0$;
and negative curvature or hyperbolic geometry e.g. saddle shaped where $C>2\pi a$ and $k<0$.

\section{}
\begin{equation}
k=\frac{3}{\pi}\lim_{a\to 0}\left(\frac{2\pi a - C}{a^3}\right)
\end{equation}
and
\begin{equation}
C=2\pi x = 2\pi R sin\frac{a}{R}
\end{equation}
however
\begin{equation}
sin x \approx x-\frac{x^3}{3!}
\end{equation}
so that
\begin{equation}
C\approx 2\pi a\left(1-\frac{a^2}{6R^2}\right)
\end{equation}
giving
\begin{equation}
k=\frac{3}{\pi}\lim_{a\to 0}\left(\frac{2\pi a - 2\pi a\left(1-\frac{a^2}{6R^2}\right)}{a^3}\right)
\end{equation}
which simplifies to
\begin{equation}
k=\frac{3}{\pi}\lim_{a\to 0}\left(\frac{\frac{2\pi a^3}{6R^2}}{a^3}\right)
\end{equation}
The $a^3$ cancel giving
\begin{equation}
k=\frac{3}{\pi}\frac{2\pi}{6R^2}
\end{equation}
and finally $6\pi$ cancels leaving
\begin{equation}
k=\frac{1}{R^2}
\end{equation}

\section{}
As the coordinates $\theta$ and $\phi$ represent angles and we know that the universe is isotropic and looks the same from all
directions and so the distance cannot depend on direction and therefore the metric can be simplified into 
$ds^2=f(r)dr^2+r^2d\theta^2+r^2sin^2\theta d\phi^2$ where $f(r,\theta,\phi)$ only
depends on r or distance $f(r)$ and also that $f'(r)=f''(r)$.
As the universe is homogeneous then the curvature k will be the same everywhere in space and time so we can use Gauss'
formula to calculate k which ultimately gives
\begin{equation}
\frac{1}{f^2dr}\frac{df}{dr}=2kr
\end{equation}
Integrating the above gives
\begin{equation}
\int\frac{1}{f^2}df=k\int 2r dr
\end{equation}
\begin{equation}
-\frac{1}{f}=kr^2-C
\end{equation}
which finally gives
\begin{equation}
f(r)=\frac{1}{C-kr^2}
\end{equation}
and setting $C=1$ because $f=1$ and $k=0$ for flat space we arrive at
\begin{equation}
f(r)=\frac{1}{1-kr^2}
\end{equation}
k determines the curvature of the universe and hence whether the universe is flat ($k=0$), open ($k<0$) or closed ($k>0$).
%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-04-25 cosmology/Tutorial3/Tutorial3.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage{amsmath}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{Cosmology Tutorial 3}
\begin{document}
\maketitle

\section{}
Here goes:
\begin{flalign*}
&s^2=\left(ct\right)^2-x^2 \\
&=c^2\gamma^2\left(t'-\frac{v'x'}{c^2}\right)-\gamma^2\left(x'-v't'\right)^2 &\\
&=\gamma^2\left(ct'-\frac{v'x'}{c}\right)^2-\gamma^2\left(t'^2v'^2-2t'v'x'+x'^2\right) &\\
&=\gamma^2\left(c^2t'^2+\frac{v'^2x'^2}{c^2}-2t'v'x'\right)-\gamma^2\left(t'^2v'^2-2t'v'x'+x'^2\right) &\\
&=\gamma^2\left(c^2t'^2+\frac{v'^2x'^2}{c^2}\right)-\gamma^2\left(t'^2v'^2+x'^2\right) &\\
&=\gamma^2\left[\left(c^2t'^2-t'^2v'^2\right)-x'^2\left(1-\frac{v'^2}{c^2}\right)\right] &\\
&=\gamma^2\left[\left(1-\frac{v'^2}{c^2}\right)c^2t'^2-x^2\left(1-\frac{v'^2}{c^2}\right)\right] &\\
&=\gamma^2\left[\left(\frac{1}{\gamma^2}c^2t'^2-x'^2\frac{1}{\gamma^2}\right)\right] &\\
&=c^2t'^2-x'^2
\end{flalign*}
\section{}
In special relativity time doesn't have a preferred frame but in cosmology, with an homogeneous and isotropic universe, then all observers can agree on a preferred
frame e.g. the temperature of the cosmic microwave background which decreases with the age of the universe.
\section{}
\begin{flalign*}
&dA=\frac{R_0r}{1+z} &\\
&dL=R_0r(1+z)
\end{flalign*}
so combining these gives
\begin{flalign*}
&dA/dL=\frac{\frac{R_0r}{1+z}}{R_0r(1+z)} &\\
&=(1+z)^2
\end{flalign*}
Both angular diameter distance and luminosity distance both assume a static Euclidean universe. Angular diameter distance is the distance
an object would have to be for its measured angular size. Luminosity distance is the distance an object would have to be to measure its
known luminosity. Proper distance is the actual distance an object would be according to a moment in 
time and varies with expansion of the universe. Light travel distance is the time it took light to reach the observer multiplied by the speed of light.
For low values of z then these distances are similar but for higher z they are different. As z increases the proper distance increases but tends towards a maximum because
the observable universe is finite,
the luminosity distances increases indefinitely because the 1+z term can increase indefinitely, whilst the angular diameter distance reaches a maximum and then decreases,
because at great distances the universe was smaller and so the angular diameter occupied a larger fraction of the universe.
\section{}
Proper distance is the actual distance an object would be according to a moment in 
time and varies with expansion of the universe. It would be the actual distance measured if one were to actually pace it out!
\begin{flalign*}
&d_{pr}(t_0)=R_0(f) &\\
&d_pr(t)=Rf(r) &\\
&d_pr(t_0)=\frac{R_0}{R}d_{pr}(t) &\\
&V_r=\frac{d}{dt}d_{pr}=\dot{R}f(r) &\\
&V_r=\frac{\dot{R}}{R}d_pr
\end{flalign*}
At small scales the Hubble law breaks down due to perculiar velocities (movements not associated with expansion of universe) e.g. the Andromeda galaxy is moving towards the
Milky Way. At large scales the Hubble parameter is different as to what it is now, plus the curvature of the universe affects the results.
%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-04-25 cosmology/Tutorial5/Tutorial5.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage{amsmath}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{Cosmology Tutorial 5}
\begin{document}
\maketitle
\section{}
\subsection{}
For EDS universe $\Omega=1$ so
\begin{flalign*}
& rR_0=\frac{2c}{H_0}\left(\frac{z-(1+z)^{-1/2}-1}{1+z}\right)
\end{flalign*}
\subsection{}
Making the following substitutions: $\Theta=\frac{2c}{H_0}$ and $a=1+z$:
\begin{flalign*}
& d_A=\Theta\frac{1}{a}\left(1-a^{1/2}\right)
\end{flalign*}
The derivative is:
\begin{flalign*}
& \frac{dd_A}{da}=\frac{3\Theta}{2a^{5/2}}-\frac{\Theta}{a^2}
\end{flalign*}
and then setting to zero to get the maximum:
\begin{flalign*}
& \frac{a^2 3\Theta}{2a^{5/2}}=\Theta &\\
& \frac{3\Theta}{2\sqrt{a}}=0 &\\
& 2\sqrt{a}=3 &\\
& a=9/4 &\\
& z=5/4
\end{flalign*}
\subsection{}
\begin{flalign*}
& \frac{2 \times 2.99\times10^8}{70 \times 1001}\left(1-1001^{-1/2}\right) &\\
& \frac{2 \times 2.99\times10^8 \times .968}{70070}=8264 &\\
& \tan\theta=\frac{D}{d_A} &\\
& \tan\theta=\frac{2.99\times10^8 \times 3.8\times10^5}{8264} &\\
& \theta=63 radians &\\
& \theta=3609^\circ
\end{flalign*}
\section{}
I got a maximum angular diameter distance of 1744.31Mpc with a $z=1.75$.
From the plots the radiation era ($\Omega_R>0.9$) lasted up until approximately $10^{-6} Gyr$ which is 1000 years.
\section{}
Einstein seemed to like the idea of a static universe i.e. one that is neither expanding or contracting, therefore he added the cosmological constant to counteract gravity.
\newline
A. The expansion rate in the early universe would be similar to models with $\Lambda>0$.
\newline
B. The EDS model of the universe results in a universe which is too young to have the observed old globular clusters. It predicts the age of the universe to be about
$8.1\times10^9$ years however globular clusters have been observed that are older than this.
\newline
C. The EDS model assumes the universe is very close to the critical density and so this model would continue to expand indefinitely because there is no cosmological constant term to prevent it.
\section{}
\begin{flalign*}
& \frac{\Omega_m}{2}-\Omega_\Lambda<0 &\\
& \frac{\Omega_m}{2}<\Omega_\Lambda &\\
& \Omega_\Lambda>\frac{1}{3}
\end{flalign*}
\section{}
Parallax measurements are a geometrical method of determining distance. As the Earth orbits the Sun it does so at a distance of 1AU so the apparent position of a distant star is altered slightly.
This can be measured and used to determine distance to that star. The distance of 1pc is equal to the distance a star would have to be to subtend 1 arcsec per 1AU.
An example of a standard candle method of determing distance is that of supernovae. As a supernova erupts it becomes very luminous and this luminosity can be used to determine its absolute
magnitude (and therefore distance) by using models of core collapse. They are however rare events.
\section{}
To be within 10\% of 5000pc the distance measured would need to be at least 4500pc:
\begin{flalign*}
& d=1/p &\\
& 1/4500=.2222"
\end{flalign*}
The International Celestial Reference Frame (ICRF) has been setup as a standard reference frame of astrometry and is based on 212 compact extragalactic radio sources (mainly quasars).
\section{}
The Malmquist bias means that as we look at greater distances we are biased towards the most luminous objects and so we miss greater numbers of fainter objects with increasing distances.
The Scott effect is similar but this time picking up more luminous objects due to the larger volume of space.
%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-04-25 cosmology/Tutorial7/Tutorial7.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage{amsmath}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Tutorial 7}
\begin{document}
\maketitle
\section{}
This equation is till relevant for particles in the early universe because the particles are highly relatvistic and kinetic energy dominiated, as are photons.
\begin{flalign*}
& t=\left(\frac{T}{1.5\times 10^{10} K}\right)^{-2} &\\
& T=1.5\times 10^{10}t^{-1/2} &\\
& T=1.5\times 10^10{10}\times 1000 &\\
& T=1.5\times 10^{13} K &\\
& kT=mc^2 &\\
& m=\frac{kT}{c^2} &\\
& m=\frac{1.38\times 10^-23 \times 10^{13}}{2.99\times 10^8} &\\
& m=1.54\times 10^{-27} kg
\end{flalign*}
The above is very close to mass of proton $1.67\times 10^{-27} kg$ so prior to this protons would have been in thermal equilibrium. The Freeze Out occurs shortly after this time where the protons and neutrons 
no longer created by weak interactions.
The limit of the above equation appears where quantum gravity starts to take effect i.e. Planck Scale.
\section{}
Neutrinos are also expected to contribute to the kinetic energy because this has an effect on the helium abundance. The time it takes for the temperature to drop depends on the contribution
of kinetic energy from relativistic particles.
\section{}
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./BBN.jpg}
\caption{Borrowed straight from Ned Wright's tutorial.}
\label{fig:1}
\end{figure}
\section{}
At temperatures above $3\times 10^9 K$ nucleosynthesis cannot occur because all the particles and anti-particles are in thermal equilibrium. The quarks are still forming nucleons so the formation of elements is
impossible.
\newline
The deuterium bottleneck is the fact that deuterium is needed for the formation of helium however due to its low binding energy it is rapidly destroyed by photons at temperatures above $10^9 K$.
\newline
Prior to nucleosynthesis protons and neutrons are in equilibrium with their anti-particles and photons. During nucleosynthesis the neutron/proton ratio was governed by the fact that the neutron has a slightly 
higher rest mass than the proton, and also through weak interactions.
\newline
Helium-4 does not occur through $2n+2p\rightarrow ^{4}He$ because it would involve a highly unlikely 4-body interaction.
\newline
One chain reaction for formation of helium is as follows:
\begin{flalign*}
& p+n\rightarrow^{2}H+\gamma &\\
& ^{2}H+^{2}H\rightarrow ^{3}H+p &\\
& ^{3}H+^{2}H\rightarrow ^{4}He+n
\end{flalign*}
The relative fraction helium is about $25\%$ because it is twice the fraction of neutrons to all nucleons - $2\times0.13=0.26$.
\section{}
I'm guessing here but if neutron decay was 10 times faster then there would be more protons created, and less neutrons, so this would slow helium creation. So, possibly, we would have 10 times less helium that at present.
\section{}
$\Omega_{baryons}$ is the fractional amount of protons and neutrons in the universe. Big Bang nucleosynthesis calculations provide a value of $\Omega_{baryons}$ by first calculating a value $^4He$, and then using this
to calculate values for $D$, $^3He$ and $^7Li$ and finally, using these values, to determine $\Omega_{baryons}$ in the early universe. Measurements from stars and other astronomical objects provide details as to where
the baryons are in the present day.
\section{}
It is hard to directly measure all present day baryons because most baryons seem to be in the form of very hot ionised gas. This is very difficult to detect directly.
%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-04-25 cosmology/Tutorial8/Tutorial8.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage{amsmath}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Tutorial 8}
\begin{document}
\maketitle
\section{}
\subsection{}
\begin{flalign*}
& \rho=\left(\frac{R}{R_0}\right)^{-3} 5\times10^{-28} &\\
& \frac{R}{R_0}=(1+z)^{-1}=8\times10^{-4} &\\
& \rho=1.7\times10^9\times5\times10^{-28}\approx8.6\times10^{-19}kg m^{-3} &\\
& T=2.7(1+z)\approx 3000K
\end{flalign*}
\subsection{}
\begin{flalign*}
& 3.6\times10^9\times3000^{\frac{3}{2}}\left(8.6\times10^{-19}\right)^{\frac{-1}{2}} &\\
& M_J\approx6\times10^5 M_\odot 
\end{flalign*}
Globular clusters and possibly giant molecular clouds have around this mass.
\newline
Fractional fluctuations from statistics are very small $(3\times10^{-32})$. Fluctuations grow proportional to the scale factor so from $z=100$ to $z=10$ they would only grow by
a factor of 10. This therefore implies that there should have been a 'seed' perturbation as part of initial conditions of big bang e.g. dark matter (WIMPS).
\section{}
\subsection{}
In radiation era fluctuations don't grow because of damping due to photons random walking out of high pressure regions.
\newline
The problem is alleviated if WIMPS are involved because dark matter decouples earlier than baryonic matter and so can start to collapse earlier as it feels no pressure.
\subsection{}
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./pic.jpg}
\caption{}
\label{fig:1}
\end{figure}
\subsection{}
With CDM these particles decouple early and are non-relativistic earlier. Fluctuations are damped at all scales and so as small things collapse first
this leads to bottom up structure formation.
With HDM these particles are relativistic until recombination. They become damped due to free-streaming. Fluctuations are wiped out up to a large mass
which leads to top down structure formation.
\section{}
\subsection{}
\begin{flalign*}
& \dot{R}^2=\frac{8\pi G}{3}\rho_{r,0}\frac{R_0^4}{R^2}-kc^2 &\\
& kc^2=R_0^2 H_0^2(\Omega_0-1) &\\
& \dot{R}^2=\frac{8\pi G}{3}\rho_{r,0}\frac{R_0^4}{R^2}-R_0^2 H_0^2(\Omega_{0,r}-1) &\\
& \left(\frac{\dot{R}}{R}\right)^2=H^2=\frac{8\pi G}{3}\rho_{r,0}\frac{R_0^4}{R^4}-\frac{R_0^2}{R^2} H_0^2(\Omega_{0,r}-1) &\\
& =\frac{8\pi G}{3}\rho_{r,0}(1+z)^4-(1-z)^2 H_0^2(\Omega_{0,r}-1) &\\
& =\Omega_{0,r}H_0^2(1+z)^4-(1+z)^2H_0^2(\Omega_{0,r}-1) &\\
& =H_0^2(1+z)^2\left[\Omega_{0,r}z^2+2\Omega_{0,r}z+\Omega_{0,r}-\Omega_{0,r}+1\right] &\\
& =H_0^2(1+z)^2\left[\Omega_{0,r}z^2+2\Omega_{0,r}z+1\right] &\\
& =H_0^2(1+z)^2\left[1+\Omega_{0,r}(2z+z^2)\right] 
\end{flalign*}
\subsection{}
\begin{flalign*}
& \Omega=\frac{8\pi G}{3H^2} &\\
& \rho=\rho_0(1+z)^4 &\\
& \Omega(z)=\frac{8\pi G\rho_0(1+z)^4}{3H_0^2(1+z)^2\left[1+\Omega_{0,r}(2z+z^2)\right]} &\\
& =\frac{8\pi G\rho+0(1+z)^2}{3H_0^2\left[1+\Omega_{0,r}(2z+z^2)\right]} &\\
& =\frac{\Omega_{0,r}(1+z)^2}{1+\Omega_{0,r}(2z+z^2)}
\end{flalign*}
That is as far as I got.
%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-04-25 cosmology/Tutorial9/Tutorial9.tex
cosmollogy

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{gensymb}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Tutorial 9}
\begin{document}
\maketitle
\section{}
\subsection{}
In the radiation era fluctuations are frozen into a high density radiation field.
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
Radiation & Plasma & Neutral \\
\hline
= & = & + \\
= & = & + \\
= & + & + \\
\hline
\end{tabular}
\caption{\label{tab:task1}Baryonic Isothermal}
\end{table}
\subsection{}
Silk damping occurs in the radiation era but Silk mass is reduced after decoupling.
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
Radiation & Plasma & Neutral \\
\hline
- & - & + \\
+ & + & + \\
+ & + & + \\
\hline
\end{tabular}
\caption{\label{tab:task2}Baryonic Adiabatic}
\end{table}
\subsection{}
Fluctuations on small scales remain undamped.
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
Radiation & Plasma & Neutral \\
\hline
+ & + & + \\
- & + & + \\
- & + & + \\
\hline
\end{tabular}
\caption{\label{tab:task3}CDM}
\end{table}
\subsection{}
Damping occurs due to free-streaming when particles are relativistic before $t_{eq}$.
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
Radiation & Plasma & Neutral \\
\hline
- & + & + \\
- & + & + \\
- & + & + \\
\hline
\end{tabular}
\caption{\label{tab:task4}HDM}
\end{table}

\section{}
\begin{flalign*}
& k_x=n_x\frac{2\pi}{100} &\\
& 2\frac{2\pi}{100}=.126 &\\
& 3\frac{2\pi}{100}=.188 &\\
\end{flalign*}
so $n_x$ must be either 0,2 or 3. The wavelengths are:
\begin{flalign*}
& \frac{2\pi}{2}=\pi &\\
& \frac{2\pi}{3}=2.094 &\\
\end{flalign*}
There are $3^3=27$ permutations of $n_x$, $n_y$ and $n_z$.
For $1<k<2$ the interval is 10 times larger so presumably there are 10 times as many permutations i.e. 270?
.
\section{}
\subsection{}
The Horizon scale at matter-energy equality is approximately 100 Mpc. The amplitude of fluctuations is $\approx10^{-4}$.
\subsection{}
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
Redshift & Amplitude & Log(Amplitude) & Log(a) \\
\hline
$4\times10^{7}$ & $10^{-8}$ & -8 & -7.6 \\
$6.6\times10^{6}$ & $10^{-7}$ & -7 & -6.8 \\
$2.2\times10^{6}$ & $10^{-6}$ & -6 & -6.3 \\
$1.4\times10^{6}$ & $10^{-5}$ & -5 & -6.1 \\
$6.3\times10^{5}$ & $10^{-4}$ & -4 & -5.7 \\
39000 & $10^{-3}$ & -3 & -4.6 \\
490 & $.01$ & -2 & -2.7 \\
29 & $.1$ & -1 & -1.5 \\
$2.2$ & $1$ & 0 & -0.5 \\
\hline
\end{tabular}
\caption{\label{tab:task6}Question 3 data}
\end{table}
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./tutorial9.jpg}
\caption{}
\label{fig:1}
\end{figure}
\section{}
Horizon problem - regions of the sky more than $1\degree (\approx 200Kpc)$ were never in causal contact before decoupling so no information could have
been exchanged. So why it is so isotropic?

Flatness problem - Why is $\Omega=1$ in universe today? The early universe must have had $\Omega=1$ as well (proof is in the notes).

Inflation solves the horizon problem because the causal horizon could have been larger in the past. Inflation also predicts that $\Omega=1$ because
size of universe was so large after inflation that any curvature would have been negligible.
\section{}
\begin{flalign*}
& K_0=1+.227N_{\nu} &\\
& N_{\nu}=3.046 &\\
& K_0=1.691 &\\
& 10^{-35}=1.691^{-\frac{1}{2}}\left(\frac{1.52\times10^{10}}{T}\right)^2 &\\
& \left(\frac{1.52\times10^{10}}{T}\right)^2=\frac{10^{-35}}{.769} &\\
& \frac{1.52\times10^{10}}{T}=\sqrt{1.3\times10^{-35}} &\\
& T=\frac{1.52\times10^{10}}{3.6\times10^{-18}} &\\
& T=4.22\times10^{27}K &\\
& kT=8.62\times10^{-5} eVK^{-1}\times 4.22\times10^{27}K &\\
& =3.64\times10^{23} eV &\\
& =3.64\times10^{14} GeV
\end{flalign*}
I couldn't answer the last bit.
%\begin{thebibliography}{1}
%\end{thebibliography}
\end{document}

### 2017-05-05 cosmology/Essay/CosmoEssayHK.tex
cosmology

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Essay: A review of Dark Matter and Dark Energy}
\begin{document}
\maketitle
\section{Introduction}
Many observations, and subequent cosmological theories, have found that approximately 95\% of the mass and energy in the universe is dark and invisible, the true nature of which is
largely unknown. The matter component is known as dark matter (DM) and the energy component dark energy (DE) This is a brief review of the history and current status of 
DM and DE.
\section{Dark Matter}
DM has a surprisingly long and interesting history. In 1933, the renowned astrophysicist, Fritz Zwicky published a paper on the velocity dispersion of galaxies within the Coma cluster. He found
that the velocity dispersion of galaxies within the cluster was so high that in order for the system to remain stable there must be some ``missing mass'' i.e. mass which
is not detectable optically. Further research in this area continued but it wasn't until the 1970s when astronomers began accurately measuring the rotation curves of spiral galaxies,
with the discovery of HI 21cm radio emission, that interest in this area became of great importance. HI 21cm radio emission is produced by neutral hydrogen during a change of energy
state. It has a very precise wavelength of 21cm.
Rotation curves are a measure of the orbital velocity of the stars and gas in a galaxy plotted as a function of distance from the galactic centre. According to
Keplerian laws rotation curves would be expected to decline with increasing radius however it was found that they remain flat i.e. constant, all the way out to the largest radii observed, 
both optically and with 21 cm. According to Newtonian gravity this implies that there must be some ``hidden mass'' within the galaxies exerting a gravitational effect on the system resulting
in these flat rotation curves. This ``hidden mass'' was termed dark matter.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./rotationcurves.jpg}
\caption{Examples of flat galaxy rotation curves \citep{ruben}}
\label{fig:1}
\end{figure}

Apart from high velocity dispersions in galaxy clusters and flat rotation curves other evidence for DM includes gravitational lensing, hot gas in clusters, the Cosmic Microwave Background (CMB)
 and through computer modelling of the growth of 
large scale structure in the early universe (discussed later). Gravitational lensing occurs when light from distant galaxies behind a galaxy cluster is distorted due to the 
large gravitational pull of the cluster. This is known as strong lensing and is quite rare, however weak lensing also occurs where there is just a small distortion of the galaxies size
and shape. 
The mass of the galaxy cluster can be measured and compared to how much the light is expected to be bent by gravity. The results show that there must be far more mass in clusters 
than can be detected. Galaxy clusters when viewed in X-ray reveal huge amounts of hot gas which can only be explained by large amounts of DM which provide a potential gravitational well to hold the gas.
The CMB is the remnants of radiation from the early universe. Anisotropies in the CMB also reveal evidence for DM. Anisotropies are irregularities seen in the CMB which, after having been
thoroughly studied, reveal what woud be expected if thermal variations in a small space had grown to form the observable universe today.
Before decoupling from baryonic matter (i.e. ordinary matter comprised of 
mainly protons and neutrons), photons underwent oscillations that froze in at redshift $z\approx1100$. Decoupling refers to events in the early universe when different
particles fall out of thermal equilibrium with each other.
By studying the height of the peaks in these oscillations it has been 
found that the universe is comprised of approximately 5\% baryonic matter, 26\% dark matter
and the remainder dark energy (discussed later). Some authors \citep{freese} regard the evidence of DM from the CMB as ``irrefutable''.

Further evidence of DM, which has been described as ``direct empirical proof'' \citep{clowe}, has been found in observations of the Bullet cluster of galaxies. The Chandra X-ray observatory 
has detected the baryonic matter in the merger of two smaller clusters, whilst DM has been deduced from gravitational lensing, and they clearly show that are behaving differently.
At the collision point the baryonic matter has slowed down due to friction whilst the DM has passed through this point.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./bullet.jpg}
\caption{The bullet cluster: a) Optical view from Magellan telescope with plotted contours of mass distribution obtained from gravitational lensing.
b) Same contours but with Chandra X-ray data showing hot plasma. Most of the matter is lying in a different location to the hot plasma which has been slowed down by friction 
during the merger. \citep{clowe}}
\label{fig:2}
\end{figure}

With the growing evidence for the existence of DM, obvious questions to ask are ``What is it made of?'' and ``Is it baryonic?''. About twenty years ago it was suggested that dark matter is
comprised of objects such as faint stars, sub-stellar objects or stellar remnants. Collectively these came to be known as massive compact halo objects (MACHOs). However recent studies 
\citep{freese} have found that these objects couldn't account for all the DM in the universe. This then suggests that if DM exists it must be non-baryonic.

Particle physicists have postulated many possible candidates for non-baryonic DM particles. These include neutrinos, primordial black holes (PBHs), magnetic monopoles, neutalinos and photinos.
However, many of these have been effectively ruled out by research e.g. \cite{gaggero} who concluded that PBHs couldn't account for more than 20\% of DM in the universe.
Two of the most popular non-baryonic DM candidates currently are: Axions and Weakly Interacting Massive Particles (WIMPs). These are both hypothetical particles invented by particle 
physicists for other reasons than DM, which relieves cosmologists of having to invent new particles. Axions arise from a problem in quantum chromodynamics but have interesting 
consequences for cosmology as they are predicted to be stable over cosmological timescales \citep{bertone1} and could therefore theoretically constitute DM. WIMPs have been the source
of many theoretical studies of DM and are now considered to be the leading class of DM particles, partly because they are predicted to interact via gravity and have a large mass
compared to other particles.

There are currently four main approaches to discovering WIMPs. There are ongoing experiments, with ever increasing sensitivity, at CERN, using
the Large Hadron Collider. One of the goals of the two detectors, ATLAS and CMS, was specifically to try and discover the nature of DM. It has so far been unsuccessful in that respect.
There are also attempts to directly detect WIMPs in underground DM laboratories worldwide. So far there has been no confirmed detection of WIMPs. Indirect detection of WIMPs is also being
carried out in objects such as the galactic centre, galaxy clusters and dwarf galaxies i.e. objects with a predicted over-density of WIMPs. Again, there has been no confirmed detection.

An interesting fourth approach to the hunt for DM is to discover dark stars. It is predicted that the first stars to form in the early universe, at redshift $z\approx 10-50$, may have been 
comprised of mainly hydrogen and helium and yet were powered by DM heating rather than nuclear fusion. These are believed to have formed when the universe was much denser than now and 
consisted of high density DM halos. It is hoped that the James Webb Space Telescope will be able to detect dark stars and so enable us to study WIMPs in more detail.

With the advent of numerical simulations, models that involve DM have been extensively explored. Currently a very popular model of the evolution of structure in the universe is
the Lambda Cold Dark Matter ($\Lambda CDM$) model. $\Lambda$ here refers to the cosmological contant discussed in the DE section. 
Computer simulations have shown that if the DM particles are relativistic (``hot'') then, on small scales, density fluctuations are Silk damped, or
washed out, the random thermal motion of DM particles. This results in small scale structure fluctuations to be suppressed. However, if the DM particles are ``cold'' i.e. cold dark matter (CDM), then
they undergo a very different structure formation because they have a much smaller free-streaming length and can form low mass halos which can gradually build up into larger DM
structures. These DM structures provide a gravitational well that baryonic particles can fall into to form baryonic structure. This results in a bottom-up process of structure 
formation which is very different to the top-down sequence predicted for hot DM. These simulations have further convinced cosmologists that CDM comprised of WIMPs are the best
explanation for the growth of large scale structure in the universe. In fact, the $\Lambda CDM$ model agrees so well with observations that it is often regarded as the ``standard model
of cosmology''.

Even though the evidence for DM is growing, there has been no confirmed detection of DM or WIMPs and the theory of DM leads to a number of problems \citep{sellwood} which I don't
have space to detail here. This has led some cosmologists to believe that the Newtonian theory of gravity may need to be amended. One of these theories is called Modified Newtonian
Dynamics (MOND). It is a conceptually simple idea but it has some far-reaching consequences. The basic idea is that Newton's second law be amended from $F=ma$ to $F=ma^2/a_0$ in the
limit of very low accelerations ($a<<a_0\approx 1.2\times10^{-10}m/s^2$). This could then account for the observed motions of stars within galaxies without having to introduce
DM. MOND has been quite successful with explaining flat rotation curves but less successful with galaxy clusters. MOND, however, is not without its own problems particularly when it
comes to integrating MOND with general relativity (GR), explaining the growth of large scale structure, and explaining the first CMB acoustic peak.
It would appear that either the LHC, or any detection methods previously described, will eventually detect WIMPs and confirm the existence of DM and MOND will die out as a theory or, if
no detection of DM is forthcoming then the theory of MOND may well grow.

\section{Dark Energy}
Cosmic acceleration, and associated DE, is arguably one of the greatest unsolved problems in contemporary physics. DE actually has a longer history than DM. As Einstein was forming his 
field equations for general relativity he realised that this would result in a universe that would
gravitationally attract and therefore contract. He therefore introduced the cosmological constant to balance gravity and ensure a static universe. When observations by Hubble in 1929
revealed that the universe is expanding, Einstein referred to his failure to predict a dynamic universe has his biggest blunder. The cosmological constant therefore fell into disuse.
However, in 1980 Alan Guth and Alexei Starobinsky proposed the idea of cosmic inflation which involved an exponential expansion of the universe immediately after the Big Bang. The theory
of inflation involves a negative pressure field which creates a repulsive force. This idea is similar to DE, however even when inflation became widely accepted the 
cosmological constant was considered irrelevant. This changed however in 1998 when observations of supernovae revealed that the universal expansion is accelerating. This was the first
direct evidence of DE.

Type Ia supernovae occur when a white dwarf in a binary pair accretes enough mass so that it goes over the Chandrasekhar mass and triggers a supernova. The supernova therefore doesn't
depend on the mass of the white dwarf on the other star in the binary. Type Ia supernovae can be considered as standard candles and enable astronomers to obtain a luminosity distance.
To measure cosmic expansion the apparent magnitude of distant supernovae are compared to those of more local supernovae which are not affected by DE. The Supernova Cosmology Project
which carried out one of the surveys in 1998 found that, with a flat cosmology ($\Omega_{\Lambda}+\Omega_m=1$), that $\Omega_m=.28$ and so that $\Omega_{\Lambda}=.72$. $\Omega$ is the
density parameter and is defined as the ratio of the observed density to the critical density. $\Omega_m$ here is 
the proportion of matter in universe (DM and baryonic) whilst $\Omega_{\Lambda}$ represents the proportion of DE in universe. The deceleration parameter is calculated as follows:
\begin{equation}
\frac{1}{2}(\Omega_m-2\Omega_{\Lambda}+2\Omega_R) 
\end{equation}
This then results in a negative deceleration parameter indicating that the expansion of the universe is accelerating.
Since 1998 there have been many more supernova surveys and it is has become an active area of observational cosmology. All this supernova data indicates that the apparent luminosity
of Type Ia supernovae declines more rapidly with increasing redshift than would be expected if $\Omega_M=1$ and $\Omega_{\Lambda}=0$ i.e. an Einstein-de Sitter universe, or the 
luminosity distance is greater than expected. Of course, this data could be interpreted as being due to other effects such as dust or gas in the line of sight. More evidence is required.

Although it doesn't directly constrain DE, data obtained from CMB anisotropies does provide accurate figures for parameters that are needed in order to study DE. By measuring anisotropies 
within the CMB data, from WMAP and Planck satellites, accurate figures for $\Omega_B$ and $\Omega_{DM}$ have been obtained. 
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./planck2.jpg}
\caption{The Planck satellite helped provide accurate figures for cosmological parameters \citep{planck}}
\label{fig:3}
\end{figure}

Similarly, Baryonic Acoustic Oscillations (BAO) also
help the study of DE by constraining cosmological parameters. BAOs are regular and periodic fluctuations in the density of baryonic matter. They are measured by looking at large
scale structures in the universe through galaxy surveys. In a similar way that Type Ia supernovae
are standard candles, BAOs are standard rulers. By comparing the clustering of galaxies today (using redshift surveys like the Sloan Digital Sky Survey - SDSS) and comparing it with
data from CMB cosmologists have a way of measuring DE that is independent of the supernova method.

Unlike DM, where particle physicists and cosmologists have established theories as to what it may be comprised of, the nature of DE is completely unknown. DE is often referred to as
vacuum energy. According to quantum field theory a vacuum or empty space actually consists of transient fluctuations known as virtual particles. 
They do, however, have some properties in common with ``ordinary'' particles.
It has been suggested that these virtual particles may give
rise to DE. However, the density of virtual particles is about 120 times larger than the DE density. There is currently no explanation for this. The DE density is believed
to be $\approx10^{-27}kg/m^3$ which is low but it has such a profound effect on the universe because it fills all of empty space.

There are many theories for DE. However, the simplest one is that it is constant i.e. a cosmological constant. The cosmological constant has a negative pressure equal to
its energy density and so exerts a repulsive force causing the expansion of the universe to accelerate i.e. $P=-\rho$. The previously mentioned $\Lambda CDM$ model uses
this form of DE and is the currently preferred cosmological model. An important parameter in the study of DE is the equation of state parameter $\omega$. This comes from: 
\begin{equation}
P=\omega\rho c^2
\end{equation}
where P is pressue, $\rho$ is density, c is speed of light and $\omega$ is the equation of state parameter.
In the $\Lambda CDM$ model $\omega$ is assumed to be -1.

Another theory of DE is quintessence. In this theory, DE is assumed to be a scalar field and unlike the 
cosmological constant can vary in time and space. However, there is currently no evidence to support this theory.
It has been suggested that instead of vacuum energy, cosmic acceleration can be explained by a modified version of GR. Tests of GR within the solar system have proven the theory
to be accurate but it may be the case that GR breaks down on cosmological scales. However there currently is no viable theory of GR that can fully explain cosmic acceleration
without introducing some form of DE. However, this doesn't preclude a theory arising in the future as, for example, it wasn't until the advent of GR in 1915 that the precession
of Mercury's orbit could be fully explained.

The Dark Energy Survey (DES) was specifically created to probe DE. Using a wide-field camera on the 4m Blanco Telescope in Chile the project aims to catalogue 300 million
galaxies. The main goals of the survey are to characterise DE and DM and also to test alternative theories of gravity. The survey takes photometric redshifts of galaxies and
also carries out galaxy-galaxy weak lensing measurements i.e. where both the lenses and sources are galaxies. It is also carrying out a time domain survey of Type Ia supernovae.
Last year the project reported that they had further constrained $\Omega_m$ to be $=0.31\pm0.09$. The SDSS is also carrying out DE probes with the Extended Baryon Oscillation
Spectroscopic Survey (eBOSS). The main goals of eBOSS are to measure clustering on large scales and to measure BAOs. Other projects include Panoramic Survey Telescope and Rapid 
Response System (Pan-STARRS) and the Hobby-Eberly Telescope Dark Energy Experiment (HETDEX). 
One of the aims of Pan-STARRS was to constrain the equation of state parameter for DE ($\omega$) which was found to be $\omega=-1.14$ \citep{zheng}. The Hobby-Eberly Telescope 
has recently been upgraded for HETDEX and will probe DE through spectrographs of galaxies over the redshift $0<z<4$ and also analysing BAOs. The aim of HETDEX is to provide a
direct detection of DE at $z\approx3$ \citep{hill}.

A new and exciting development by the European Space Agency, which will further probe DE, will be the launch of Euclid which is expected to be in 2020. Euclid aims to find out
the nature the acceleration of the expansion of the universe and DE. It will measure shapes and redshifts of galaxies and also explore the large scale structure of the universe
by measuring the distribution of galaxy clusters. Euclid is a satellite equipped with a 1.2m telescope and three imaging and spectroscopic instruments working in optical and
near-infrared. It will be placed at the L2 Langrange point, a stable point in space, and is planned to operate for six years. The plan is cover 15000 square degrees of sky and
image one billion galaxies and take redshifts measurements of 100,000 galaxies \citep{amendola}. With this data Euclid will able to detail clustering of galaxies out to redshift
z=2 and out to redshift z=3 utilising weak lensing. Redshift measurements and weak lensing will also accompanied by measuring correlations with the CMB, obtaining galaxy cluster 
information, strong lensing and possibly even obtaining luminosity distance through supernovae Type Ia measurements.

DE may well have a profound effect on the final fate of the universe. Observations suggest that the expansion of the universe will continue to accelerate forever and so the 
universe will cool as it expands. Redshift will eventually stretch photons to undetectable wavelengths and galaxies will therefore appear to disappear. Star formation will cease 
as gas supply runs out. Stars will
eventually die out, one by one, leaving stellar remnants. Some theories even predict that stellar remnants will disappear due to proton decay leaving only black holes,
which themselves will eventually disappear due to Hawking radiation. Proton decay is a hypothetical form of radiactive decay in which protons decay into ligher particles. 
Hawking radiation is blackbody radiation believed to be emitted by black holes because of quantum
effects near the event horizon. This is often referred to as the heat death of the universe. It would seem that the ultimate fate of our universe could potentially be a very 
cold, dark void.

\begin{thebibliography}{1}
\bibitem[Abbott et al(2015)]{abbott}Abott, T., et al (Dark Energy Survey Collaboration), 2015, The Dark Energy Survey: more than dark energy - an overview, arXiv:1601.00329v3
\bibitem[Amendola et al(2016)]{amendola}Amendola, L., et al (Euclid Theory Working Group), 2016, Cosmology and Fundamental Physics with the Euclid Satellite, arXiv:1006.00180v1
\bibitem[Arun et al(2017)]{arun}Arun, K., Gudennavar, S., B., Sivaram, C., 2017, Dark Matter, Dark Energy, and Alternate Models: A Review, arXiv:1704.06155
\bibitem[Baldry(2016)]{baldry}Baldry, I., Cosmology lecture notes - MSc Astrophysics, 2016, Not in public domain
\bibitem[Bergstrom(2012)]{bergstrom}Bergstrom L., 2012,  Dark Matter Evidence, Particle Physics Candidates and Detection Methods, arXiv:1205.4882
\bibitem[Bertone et al(2004)]{bertone2}Bertone, G., Hooper, D., Silk, J., 2004, Particle Dark Matter: Evidence, Candidates and Constraints, arXiv:hep-ph/0404175v2
\bibitem[Bertone and Hooper(2016)]{bertone1}Bertone, G., Hooper, D., 2016, A History of Dark Matter, arXiv:1605.04909v2
\bibitem[Calmet and Kuntz(2017)]{calmet}Calmet, X., Kuntz, I., 2017, What is modified gravity and how to differentiate it from particle dark matter?, arXiv:1702.03832v2
\bibitem[Clowe et al(2006)]{clowe}Clowe, D., Bradac, M., Gonzalez, A., H., Markevitch, M., Randall, S., W., Jones, C., Zaritsky, D., 2006, A Direct Empirical Proof of the Existence of Dark Matter, arXiv:astro-ph/0608407v1
\bibitem[Cohen-Tannoudji(2015)]{cohen}Cohen-Tannoudji, G., 2015, The dark universe and quantum vacuum, arXiv:1507.00460
\bibitem[Fornengo(2017)]{forengo}Fornengo, N., 2017, Dark matter overview, arXiv:1701.00119v3
\bibitem[Freese(2017)]{freese}Freese, K., 2017, Status of dark matter in the universe, arXiv:1701.01840v1
\bibitem[Gaggero et al(2016)]{gaggero}Gaggero, D., Bertone, G., Calore, F., Connors, R., M., T., Lovell, M., Markoff, S., Storm E., 2016, Searching for Primodial Black Holes in the radio and X-ray sky, arXiv:1612.00457v1
\bibitem[He Zhang(2017)]{he}He, H., Zhang, Z., 2017, Direct Probe of Dark Energy through Gravitational Lensing Effect, arXiv:1701.03418v2
\bibitem[Hill et al(2008)]{hill}Hill,G., J., et al, 2008, The Hobby-Eberly Telescope Dark Energy Experiment (HETDEX): Description and Early Pilot Survey Results, arXiv:0806.0183v1
\bibitem[Li et al(2012)]{li}Li, M., Li., X.,Wang, S., Wang, Y., 2012, Dark Energy: a Brief Review, arXiv:1209.0922v1
\bibitem[Lisanti(2016)]{lisanti}Lisante M, 2016, Lectures on Dark Matter Physics, arXiv:1603.03797v2
\bibitem[Mortonson et al(2013)]{mortonson}Mortonson, M., J., Weinberg, D., H., White, M., 2013, Dark Energy: A Short Review, arXiv:1401.0046v1
\bibitem[Novosyadlyj et al(2013)]{novos}Novosyadlyj, B., Pelykh, V., Zhuk, A., 2013, Dark Energy: Observational Evidence and Theoretical Models, arXiv:1502.04177
\bibitem[Ruben(1979)]{ruben}Ruben, V., C., Rotation curves of high-luminosity spiral galaxies and the rotation curve of our Galaxy, 1979, The large-scale characteristics of the galaxy; Proceedings of the Symposium, College Park, Md., June 12-17, 1978
\bibitem[Sellwood and Kosowsky(2000)]{sellwood}Sellwood, J., A., Kosowsky, A., 2000, Does Dark Matter Exist?, arXiv:astro-ph/0009074v1
\bibitem[de Swart et al(2017)]{swart}de Swart, J., Bertone, G., van Dongen, J., 2017, How Dark Matter Came to Matter, arXiv:1703.00013v1
\bibitem[Weinberg(2008)]{weinberg1}Weinberg, S., 2008, Cosmology, Oxford, 1st Edition
\bibitem[Weinberg et al(2013)]{weinberg2}Weinberg, D., H., Mortonson, M., J., Eisenstein, D., J.,Hirata, C., Riess, A., G., Rozo, E., 2013, Observational Probes of Cosmic Acceleration, arXiv:1201.2434v2
\bibitem[Zheng et al(2014)]{zheng}Zheng, W., Li, S., Xia, J., Li, M., Lu, T., 2014, Constraints on Dark Energy new New Observations including Pan-STARRS, arXiv:1405.2724v2
\bibitem[ESA (2013)]{planck}Planck, ESA, Planck reveals an almost perfect universe, Retrieved 5-5-2017
\end{thebibliography}
\end{document}

### 2017-05-05 cosmology/Essay/CosmoEssayHK.tex
cosmology

@@
 \section{Introduction}
-Many observations, and subequent cosmological theories, have found that approximately 95\% of the mass and energy in the universe is dark and invisible, the true nature of which is
+Many observations, and subsequent cosmological theories, have found that approximately 95\% of the mass and energy in the universe is dark and invisible, the true nature of which is
 largely unknown. The matter component is known as dark matter (DM) and the energy component dark energy (DE) This is a brief review of the history and current status of 
@@
 The CMB is the remnants of radiation from the early universe. Anisotropies in the CMB also reveal evidence for DM. Anisotropies are irregularities seen in the CMB which, after having been
-thoroughly studied, reveal what woud be expected if thermal variations in a small space had grown to form the observable universe today.
+thoroughly studied, reveal what would be expected if thermal variations in a small space had grown to form the observable universe today.
 Before decoupling from baryonic matter (i.e. ordinary matter comprised of 
@@
 With the advent of numerical simulations, models that involve DM have been extensively explored. Currently a very popular model of the evolution of structure in the universe is
-the Lambda Cold Dark Matter ($\Lambda CDM$) model. $\Lambda$ here refers to the cosmological contant discussed in the DE section. 
+the Lambda Cold Dark Matter ($\Lambda CDM$) model. $\Lambda$ here refers to the cosmological constant discussed in the DE section. 
 Computer simulations have shown that if the DM particles are relativistic (``hot'') then, on small scales, density fluctuations are Silk damped, or
@@
 of Type Ia supernovae declines more rapidly with increasing redshift than would be expected if $\Omega_M=1$ and $\Omega_{\Lambda}=0$ i.e. an Einstein-de Sitter universe, or the 
-luminosity distance is greater than expected. Of course, this data could be interpreted as being due to other effects such as dust or gas in the line of sight. More evidence is required.
+luminosity distance is greater than expected. Of course, these data could be interpreted as being due to other effects such as dust or gas in the line of sight. More evidence is required.
 
@@
 \end{equation}
-where P is pressue, $\rho$ is density, c is speed of light and $\omega$ is the equation of state parameter.
+where P is pressure, $\rho$ is density, c is speed of light and $\omega$ is the equation of state parameter.
 In the $\Lambda CDM$ model $\omega$ is assumed to be -1.
@@
 A new and exciting development by the European Space Agency, which will further probe DE, will be the launch of Euclid which is expected to be in 2020. Euclid aims to find out
-the nature the acceleration of the expansion of the universe and DE. It will measure shapes and redshifts of galaxies and also explore the large scale structure of the universe
+the nature the acceleration of the expansion of the universe and DE. It will measure shapes and redshifts of galaxies and also explore the large-scale structure of the universe
 by measuring the distribution of galaxy clusters. Euclid is a satellite equipped with a 1.2m telescope and three imaging and spectroscopic instruments working in optical and
 near-infrared. It will be placed at the L2 Langrange point, a stable point in space, and is planned to operate for six years. The plan is cover 15000 square degrees of sky and
-image one billion galaxies and take redshifts measurements of 100,000 galaxies \citep{amendola}. With this data Euclid will able to detail clustering of galaxies out to redshift
-z=2 and out to redshift z=3 utilising weak lensing. Redshift measurements and weak lensing will also accompanied by measuring correlations with the CMB, obtaining galaxy cluster 
+image one billion galaxies and take redshifts measurements of 100,000 galaxies \citep{amendola}. With these data Euclid will able to detail clustering of galaxies out to redshift
+z=2 and out to redshift z=3 utilising weak lensing. Redshift measurements and weak lensing will also be accompanied by measuring correlations with the CMB, obtaining galaxy cluster 
 information, strong lensing and possibly even obtaining luminosity distance through supernovae Type Ia measurements.
@@
 eventually die out, one by one, leaving stellar remnants. Some theories even predict that stellar remnants will disappear due to proton decay leaving only black holes,
-which themselves will eventually disappear due to Hawking radiation. Proton decay is a hypothetical form of radiactive decay in which protons decay into ligher particles. 
+which themselves will eventually disappear due to Hawking radiation. Proton decay is a hypothetical form of radioactive decay in which protons decay into lighter particles. 
 Hawking radiation is blackbody radiation believed to be emitted by black holes because of quantum

### 2017-05-06 cosmology/Essay/CosmoEssayHK.tex
cosmology

@@
 \bibitem[Cohen-Tannoudji(2015)]{cohen}Cohen-Tannoudji, G., 2015, The dark universe and quantum vacuum, arXiv:1507.00460
+\bibitem[ESA (2013)]{planck}Planck, ESA, http://www.esa.int/Our_Activities/Space_Science/Planck/Planck_reveals_an_almost_perfect_Universe, Retrieved 5-5-2017
 \bibitem[Fornengo(2017)]{forengo}Fornengo, N., 2017, Dark matter overview, arXiv:1701.00119v3
@@
 \bibitem[Weinberg et al(2013)]{weinberg2}Weinberg, D., H., Mortonson, M., J., Eisenstein, D., J.,Hirata, C., Riess, A., G., Rozo, E., 2013, Observational Probes of Cosmic Acceleration, arXiv:1201.2434v2
-\bibitem[Zheng et al(2014)]{zheng}Zheng, W., Li, S., Xia, J., Li, M., Lu, T., 2014, Constraints on Dark Energy new New Observations including Pan-STARRS, arXiv:1405.2724v2
-\bibitem[ESA (2013)]{planck}Planck, ESA, Planck reveals an almost perfect universe, Retrieved 5-5-2017
+\bibitem[Zheng et al(2014)]{zheng}Zheng, W., Li, S., Xia, J., Li, M., Lu, T., 2014, Constraints on Dark Energy: New Observations including Pan-STARRS, arXiv:1405.2724v2
 \end{thebibliography}

### 2017-06-21 cosmology/Essay/CosmoEssayHK.tex
2017Results

@@
 \usepackage{natbib}
-\usepackage{listings}
+\usepackage{listing}
+\usepackage{url}
 \graphicspath{{./}}
@@
 Further evidence of DM, which has been described as ``direct empirical proof'' \citep{clowe}, has been found in observations of the Bullet cluster of galaxies. The Chandra X-ray observatory 
-has detected the baryonic matter in the merger of two smaller clusters, whilst DM has been deduced from gravitational lensing, and they clearly show that are behaving differently.
+has detected the baryonic matter in the merger of two smaller clusters, whilst DM has been deduced from gravitational lensing, and they clearly show that they are behaving differently.
 At the collision point the baryonic matter has slowed down due to friction whilst the DM has passed through this point.
@@
 by measuring the distribution of galaxy clusters. Euclid is a satellite equipped with a 1.2m telescope and three imaging and spectroscopic instruments working in optical and
-near-infrared. It will be placed at the L2 Langrange point, a stable point in space, and is planned to operate for six years. The plan is cover 15000 square degrees of sky and
+near-infrared. It will be placed at the L2 Langrange point, a stable point in space, and is planned to operate for six years. The plan is to cover 15000 square degrees of sky and
 image one billion galaxies and take redshifts measurements of 100,000 galaxies \citep{amendola}. With these data Euclid will able to detail clustering of galaxies out to redshift
@@
 \bibitem[Clowe et al(2006)]{clowe}Clowe, D., Bradac, M., Gonzalez, A., H., Markevitch, M., Randall, S., W., Jones, C., Zaritsky, D., 2006, A Direct Empirical Proof of the Existence of Dark Matter, arXiv:astro-ph/0608407v1
-\bibitem[Cohen-Tannoudji(2015)]{cohen}Cohen-Tannoudji, G., 2015, The dark universe and quantum vacuum, arXiv:1507.00460
-\bibitem[ESA (2013)]{planck}Planck, ESA, http://www.esa.int/Our_Activities/Space_Science/Planck/Planck_reveals_an_almost_perfect_Universe, Retrieved 5-5-2017
+\bibitem[Cohen-Tannoudji(2015)]{cohen}Cohen-Tannoudji, G., 2015, The dark universe and quantuxm vacuum, arXiv:1507.00460
+\bibitem[ESA (2013)]{planck}Planck, ESA, \url{www.esa.int/Our_Activities/Space_Science/Planck/Planck_reveals_an_almost_perfect_Universe}, Retrieved 5-5-2017
 \bibitem[Fornengo(2017)]{forengo}Fornengo, N., 2017, Dark matter overview, arXiv:1701.00119v3

### 2017-06-21 workspace/DoubleBars/2017results/025-7-025-5/DoubleBars_v2.f90

@@ -1,0 +1,302 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-6 !vary step size
+     start=0.
+     end=15 !vary end time
+!     m1 = 1
+!     m2 = .05
+!     m3 = .000005
+!     m4 = .000005
+!      m1=.5
+!      m2=.5
+		m1=.5   !blue
+		m2=.5  !red
+      m3=0.025    !black
+      m4=0.025   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.7		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .7		!m2,vy
+
+     ys(9) = -.025     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.5		!m3, vy
+
+     ys(13) = .025     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+	ys(16) = .5		!m4, vy
+
+!     ys(1) = 0     ! m1,x
+!     ys(2) = 0  ! m1,vx
+!     ys(3) = 0       ! m1,y
+!     ys(4) = -.05		!m1,vy
+!
+!     ys(5) = 1         ! m2,x
+!     ys(6) = 0          ! m2,vx
+!     ys(7) = 0         ! m2,y
+!     ys(8) = 1		!m2,vy
+!
[…]

### 2017-06-21 workspace/DoubleBars/2017results/03-7-03-5/DoubleBars_v2.f90

@@ -1,0 +1,302 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-6 !vary step size
+     start=0.
+     end=15 !vary end time
+!     m1 = 1
+!     m2 = .05
+!     m3 = .000005
+!     m4 = .000005
+!      m1=.5
+!      m2=.5
+		m1=.5   !blue
+		m2=.5  !red
+      m3=0.03    !black
+      m4=0.03   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.7		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .7		!m2,vy
+
+     ys(9) = -.03     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.5		!m3, vy
+
+     ys(13) = .03     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+	ys(16) = .5		!m4, vy
+
+!     ys(1) = 0     ! m1,x
+!     ys(2) = 0  ! m1,vx
+!     ys(3) = 0       ! m1,y
+!     ys(4) = -.05		!m1,vy
+!
+!     ys(5) = 1         ! m2,x
+!     ys(6) = 0          ! m2,vx
+!     ys(7) = 0         ! m2,y
+!     ys(8) = 1		!m2,vy
+!
[…]

### 2017-06-21 workspace/DoubleBars/2017results/035-75-04-4/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-6 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.035    !black
+      m4=0.035   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.75		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .75		!m2,vy
+
+     ys(9) = -.04     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.4		!m3, vy
+
+     ys(13) = .04     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .4		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/04-75-045-4/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-6 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.04    !black
+      m4=0.04   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.75		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .75		!m2,vy
+
+     ys(9) = -.045     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.4		!m3, vy
+
+     ys(13) = .045     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .4		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/05-75-045-4/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-7 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.05    !black
+      m4=0.05   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.75		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .75		!m2,vy
+
+     ys(9) = -.045     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.4		!m3, vy
+
+     ys(13) = .045     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .4		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/05-75-05-3/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-7 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.05    !black
+      m4=0.05   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.75		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .75		!m2,vy
+
+     ys(9) = -.05     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.3		!m3, vy
+
+     ys(13) = .05     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .3		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/05-75-05-35/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-7 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.05    !black
+      m4=0.05   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.75		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .75		!m2,vy
+
+     ys(9) = -.05     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.35		!m3, vy
+
+     ys(13) = .05     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .35		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/06-8-06-25/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-7 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.06    !black
+      m4=0.06   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.8		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .8		!m2,vy
+
+     ys(9) = -.06     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.25		!m3, vy
+
+     ys(13) = .06     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .25		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/08-9-08-15/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-7 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.08    !black
+      m4=0.08   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.9		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .9		!m2,vy
+
+     ys(9) = -.08     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.15		!m3, vy
+
+     ys(13) = .08     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .15		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/09-95-09-1/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-7 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.09    !black
+      m4=0.09   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -.95		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = .95		!m2,vy
+
+     ys(9) = -.09     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.1		!m3, vy
+
+     ys(13) = .09     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .1		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/1-1-1-04/DoubleBars_v2.f90

@@ -1,0 +1,277 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-7 !vary step size
+     start=0.
+     end=15 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.1    !black
+      m4=0.1   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -1		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = 1		!m2,vy
+
+     ys(9) = -.1     ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.04		!m3, vy
+
+     ys(13) = .1     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .04		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
+    dist3 = y(9)-y(1) !x3-x1
[…]

### 2017-06-21 workspace/DoubleBars/2017results/1-1-1-02/DoubleBars_v2.f90
2017Results2

@@ -23,32 +23,32 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=25 !vary end time
+     end=15 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
-      m3=0.12    !black
-      m4=0.12   !yellow
+      m3=0.1    !black
+      m4=0.1   !yellow
 
      ys(1) = -.5     ! m1,x
      ys(2) = 0  ! m1,vx
      ys(3) = 0       ! m1,y
-     ys(4) = -1.1		!m1,vy
+     ys(4) = -1		!m1,vy
      
      ys(5) = .5         ! m2,x
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
-     ys(8) = 1.1	!m2,vy
+     ys(8) = 1		!m2,vy
 
-     ys(9) = -.12     ! m3,x
+     ys(9) = -.1     ! m3,x
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
-     ys(12) = -.015		!m3, vy
+     ys(12) = -.02		!m3, vy
 
-     ys(13) = .12     ! m4,x
+     ys(13) = .1     ! m4,x
      ys(14) = 0      ! m4,vx
      ys(15) = 0        ! m4,y
-     ys(16) = .015		!m4, vy
+     ys(16) = .02		!m4, vy
 
    end subroutine initialconditions
 

### 2017-06-21 workspace/DoubleBars/2017results/1-1-1-03/DoubleBars_v2.f90

@@ -23,32 +23,32 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=25 !vary end time
+     end=15 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
-      m3=0.12    !black
-      m4=0.12   !yellow
+      m3=0.1    !black
+      m4=0.1   !yellow
 
      ys(1) = -.5     ! m1,x
      ys(2) = 0  ! m1,vx
      ys(3) = 0       ! m1,y
-     ys(4) = -1.1		!m1,vy
+     ys(4) = -1		!m1,vy
      
      ys(5) = .5         ! m2,x
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
-     ys(8) = 1.1	!m2,vy
+     ys(8) = 1		!m2,vy
 
-     ys(9) = -.12     ! m3,x
+     ys(9) = -.1     ! m3,x
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
-     ys(12) = -.015		!m3, vy
+     ys(12) = -.03		!m3, vy
 
-     ys(13) = .12     ! m4,x
+     ys(13) = .1     ! m4,x
      ys(14) = 0      ! m4,vx
      ys(15) = 0        ! m4,y
-     ys(16) = .015		!m4, vy
+     ys(16) = .03		!m4, vy
 
    end subroutine initialconditions
 

### 2017-06-21 workspace/DoubleBars/2017results/12-105-11-015/DoubleBars_v2.f90

@@ -23,7 +23,7 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=25 !vary end time
+     end=15 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
@@ -33,19 +33,19 @@
      ys(1) = -.5     ! m1,x
      ys(2) = 0  ! m1,vx
      ys(3) = 0       ! m1,y
-     ys(4) = -1.1		!m1,vy
+     ys(4) = -1.05		!m1,vy
      
      ys(5) = .5         ! m2,x
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
-     ys(8) = 1.1	!m2,vy
+     ys(8) = 1.05		!m2,vy
 
-     ys(9) = -.12     ! m3,x
+     ys(9) = -.11     ! m3,x
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
      ys(12) = -.015		!m3, vy
 
-     ys(13) = .12     ! m4,x
+     ys(13) = .11     ! m4,x
      ys(14) = 0      ! m4,vx
      ys(15) = 0        ! m4,y
      ys(16) = .015		!m4, vy

### 2017-06-21 workspace/DoubleBars/2017results/12-11-11-015/DoubleBars_v2.f90

@@ -23,7 +23,7 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=25 !vary end time
+     end=15 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
@@ -38,14 +38,14 @@
      ys(5) = .5         ! m2,x
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
-     ys(8) = 1.1	!m2,vy
+     ys(8) = 1.1		!m2,vy
 
-     ys(9) = -.12     ! m3,x
+     ys(9) = -.11     ! m3,x
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
      ys(12) = -.015		!m3, vy
 
-     ys(13) = .12     ! m4,x
+     ys(13) = .11     ! m4,x
      ys(14) = 0      ! m4,vx
      ys(15) = 0        ! m4,y
      ys(16) = .015		!m4, vy

### 2017-06-21 workspace/DoubleBars/2017results/12-11-115-015/DoubleBars_v2.f90

@@ -23,7 +23,7 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=25 !vary end time
+     end=15 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
@@ -40,12 +40,12 @@
      ys(7) = 0         ! m2,y
      ys(8) = 1.1	!m2,vy
 
-     ys(9) = -.12     ! m3,x
+     ys(9) = -.115     ! m3,x
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
      ys(12) = -.015		!m3, vy
 
-     ys(13) = .12     ! m4,x
+     ys(13) = .115     ! m4,x
      ys(14) = 0      ! m4,vx
      ys(15) = 0        ! m4,y
      ys(16) = .015		!m4, vy

### 2017-07-15 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
References

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage[skip=0pt]{caption}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
%\setlength\abovecaptionskip{-5pt}
\author{Howard Kinsman}
\title{MSc Project - Modelling a Double Barred Galaxy as a Double Binary}
\begin{document}
\maketitle
\begin{abstract}
TODO
\end{abstract}
{\bf Keywords:} TODO
\section{Introduction}
The aim of this project was to create a simple model of a double barred galaxy using a double binary.

\section{Methods}
Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
Please see the Compilers subsection for a few notes on compilers - may or may not be relevant to the project.
\newline
The classical equations of motion for an n-body problem are:
\begin{equation}
m_i\mathbf{\ddot{r}}=\sum_{i\neq{j}}\frac{{m_i}{m_j}}{r^3_{ij}}\mathbf{r_{ij}}
\qquad
i=1,2,3...
\end{equation}
where $\mathbf{r_i}=\left(x_i,y_i\right)$ and $\mathbf{r_{ij}}=\mathbf{r_j}-\mathbf{r_i}$. So for a four body problem 
this equation results in the following:
\begin{equation}
\ddot{\mathbf{r_1}}=\frac{m_2\mathbf{r}_{12}}{r^3_{12}}+\frac{m_3\mathbf{r}_{13}}{r^3_{13}}+\frac{m_4\mathbf{r}_{14}}{r^3_{14}}
\end{equation}
\begin{equation}
\ddot{\mathbf{r_2}}=\frac{m_1\mathbf{r}_{21}}{r^3_{21}}+\frac{m_3\mathbf{r}_{23}}{r^3_{23}}+\frac{m_4\mathbf{r}_{24}}{r^3_{24}}
\end{equation}
\begin{equation}
\ddot{\mathbf{r_3}}=\frac{m_1\mathbf{r}_{31}}{r^3_{31}}+\frac{m_2\mathbf{r}_{32}}{r^3_{32}}+\frac{m_4\mathbf{r}_{34}}{r^3_{34}}
\end{equation}
\begin{equation}
\ddot{\mathbf{r_4}}=\frac{m_1\mathbf{r}_{41}}{r^3_{41}}+\frac{m_2\mathbf{r}_{42}}{r^3_{42}}+\frac{m_3\mathbf{r}_{43}}{r^3_{43}}
\end{equation}
The above equations were encoded into Fortran. A full code listing is supplied in a separate file.
\subsection{Initial Conditions}
As a starting point both the outer binary and inner binary were placed in circular orbits and tested separately and then combined to ensure a stable
starting point. The inner binary had initially a neglible mass.
The outer binary initial conditions were:
\begin{lstlisting}
m1=.5, m2=.5, x1=-.5, x2=.5, y1=0, y2=0,
vx1=0, vx2=0, vy1=-.5, vy2=.5
\end{lstlisting}
and for the inner binary:
\begin{lstlisting}
m3=.001, m4=.001, x3=-.001, x4=.001, y3=0, y4=0, 
vx3=0, vx4=0, vy3=-.5, vy4=.5
\end{lstlisting}
Plots of the two initial binaries are shown in Fig. \ref{fig:outerbinary} and 
\ref{fig:innerbinary}. The scale of the inner binary is so small that it is difficult
to see in GnuPlot.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2016results/outerbinary/Orbit.eps}
\caption{Outer Binary}
\label{fig:outerbinary}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2016results/innerbinary/Orbit.eps}
\caption{Inner Binary}
\label{fig:innerbinary}
\end{figure}
I then proceeded to perturb the system by gradually increasing the mass of the inner binary. As the system became unstable I componensated for
the increasing mass of the inner binary by increasing the velocity of the outer binary and also decreasing the velocity and increasing the binary 
separation of the inner binary. In many of the configurations the systems were inherently unstable and produced collisions with some 
or all of the bodies being ejected from the system. These were rejected and only 'stable' configurations were considered.
Table \ref{tab:variables} shows the perturbations made to the system - only the variables in this table changed and all the other initial
conditions remained the same.
\begin{table}[ht!]
  \centering
  \caption{Configurations}
  \label{tab:variables}
  \begin{tabular}{ccccccccc}
   Config. & m3 & m4 & vy1 & vy2 & x3 & x4 & vy3 & vy4\\
    \hline
   1 & .001 & .001 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
   2 & .002 & .002 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
   3 & .003 & .003 & -.5 & .5 & -.001 & .001 & -.5 & .5\\
   4 & .004 & .004 & -.5 & .5 & -.004 & .004 & -.5 & .5\\
   5 & .004 & .004 & -.55 & .55 & -.004 & .004 & -.5 & .5\\
   6 & .004 & .004 & -.57 & .57 & -.004 & .004 & -.5 & .5\\
   7 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.3 & .3\\
   8 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.4 & .4\\
   9 & .005 & .005 & -.58 & .58 & -.005 & .005 & -.35 & .35\\
   10 & .006 & .006 & -.6 & .6 & -.006 & .006 & -.3 & .3\\
   11 & .025 & .025 & -.65 & .65 & -.02 & .02 & -.5 & .5\\
   12 & .025 & .025 & -.7 & .7 & -.025 & .025 & -.5 & .5\\
   13 & .03 & .03 & -.7 & .7 & -.03 & .03 &-.5 & .5\\
   14 & .035 & .035 & -.75 & .75 & -.04 & .04 &-.4 & .4\\
   15 & .04 & .04 & -.75 & .75 & -.045 & .045 &-.4 & .4\\
   16 & .05 & .05 & -.75 & .75 & -.045 & .045 &-.4 & .4\\
   17 & .05 & .05 & -.75 & .75 & -.05 & .05 &-.35 & .35\\
   18 & .05 & .05 & -.75 & .75 & -.05 & .05 &-.3 & .3\\
   19 & .06 & .06 & -.8 & .8 & -.06 & .06 &-.25 & .25\\
   20 & .08 & .08 & -.9 & .9 & -.08 & .08 &-.15 & .15\\
   21 & .09 & .09 & -.95 & .95 & -.09 & .09 &-.1 & .1\\
   22 & .1 & .1 & -1 & 1 & -.1 & .1 & -.05 & .05\\
   23 & .1 & .1 & -1 & 1 & -.1 & .1 & -.04 & .04\\
   24 & .1 & .1 & -1 & 1 & -.1 & .1 & -.03 & .03\\
   25 & .1 & .1 & -1 & 1 & -.1 & .1 & -.02 & .02\\
   26 & .12 & .12 & -1.05 & 1.05 & -.11 & .11 & -.015 & .015\\
   27 & .12 & .12 & -1.1 & 1.1 & -.11 & .11 & -.015 & .015\\
   28 & .12 & .12 & -1.1 & 1.1 & -.115 & .115 & -.015 & .015\\
   29 & .12 & .12 & -1.1 & 1.1 & -.12 & .12 & -.015 & .015\\
   30 & .1 & .1 & -1 & 1 &-1. & .1 & -.1 & .1\\
  \end{tabular}
\end{table}

\subsection{Compilers}
As I had converted OdeInt into Fortran 90 I wanted to make sure I hadn't introduced any errors. I therefore compared the output of the Numerical Recipes
version of OdeInt with mine. I was compiling with GFortran and the Fortran 77 version was compiled with g77. They were different! So naturally concerned 
I then compiled the Fortran 77 version with GFortran. The results were identical to mine, which satisified me that I hadn't introduced any errors.

\section{Results}
As can be seen from Fig. \ref{fig:config1} even a neglible mass inner binary causes the outer binary to begin to separate and become less tightly bound.
Figures Fig. \ref{fig:config1} to \ref{fig:config18i} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
In many cases the inner binary is only visible as a point at the centre of the outer binary.
As the mass of the inner binary increased the system became increasingly unstablle and the effect on the outer binary became more pronounced.
Even the 'stable' configurations listed in Table \ref{tab:variables} hold for only a few orbits.
I was unable to recreate a stable system for a mass $>.08$ for the inner bar.
The plots below reveal that increasing the mass of the inner binary causes the outer binary to separate and
form wider orbits.

\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2016results/stablebase/Orbit.eps}
\caption{Configuration 1}
\label{fig:config1}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2016results/stablebase/Inner.eps}
\caption{Configuration 1 - Inner Bar}
\label{fig:config1i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2016results/002-5-001/Orbit.eps}
\caption{Configuration 2}
\label{fig:config2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2016results/002-5-001/Inner.eps}
\caption{Configuration 2 - Inner Bar}
\label{fig:config2i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2016results/003-5-001/Orbit.eps}
\caption{Configuration 3}
\label{fig:config3}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./2016results/003-5-001/Inner.eps}
\caption{Configuration 3 - Inner Bar}
\label{fig:config3i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-5-004/Orbit.eps}
\caption{Configuration 4}
\label{fig:config4}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-5-004/Inner.eps}
\caption{Configuration 4 - Inner Bar}
\label{fig:config4i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-55-004/Orbit.eps}
\caption{Configuration 5}
\label{fig:config5}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-55-004/Inner.eps}
\caption{Configuration 5 - Inner Bar}
\label{fig:config5i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-57-004/Orbit.eps}
\caption{Configuration 6}
\label{fig:config6}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-57-004/Inner.eps}
\caption{Configuration 6 - Inner Bar}
\label{fig:config6i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-3/Orbit.eps}
\caption{Configuration 7}
\label{fig:config7}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-3/Inner.eps}
\caption{Configuration 7 - Inner Bar}
\label{fig:config7i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-4/Orbit.eps}
\caption{Configuration 8}
\label{fig:config8}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-4/Inner.eps}
\caption{Configuration 8 - Inner Bar}
\label{fig:config8i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-35/Orbit.eps}
\caption{Configuration 9}
\label{fig:config9}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-35/Inner.eps}
\caption{Configuration 9 - Inner Bar}
\label{fig:config9i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/006-6-006-3/Orbit.eps}
\caption{Configuration 10}
\label{fig:config10}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/006-6-006-3/Inner.eps}
\caption{Configuration 10 - Inner Bar}
\label{fig:config10i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/025-65-02/Orbit.eps}
\caption{Configuration 11}
\label{fig:config11}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/025-65-02/Inner.eps}
\caption{Configuration 11 - Inner Bar}
\label{fig:config11i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/025-7-025-5/Orbit.eps}
\caption{Configuration 12}
\label{fig:config12}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/025-7-025-5/Inner.eps}
\caption{Configuration 12 - Inner Bar}
\label{fig:config12i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/03-7-03-5/Orbit.eps}
\caption{Configuration 13}
\label{fig:config13}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/03-7-03-5/Inner.eps}
\caption{Configuration 13 - Inner Bar}
\label{fig:config13i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/035-75-04-4/Orbit.eps}
\caption{Configuration 14}
\label{fig:config14}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/035-75-04-4/Inner.eps}
\caption{Configuration 14 - Inner Bar}
\label{fig:config14i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/04-75-045-4/Orbit.eps}
\caption{Configuration 15}
\label{fig:config15}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/04-75-045-4/Inner.eps}
\caption{Configuration 15 - Inner Bar}
\label{fig:config15i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/05-75-045-4/Orbit.eps}
\caption{Configuration 16}
\label{fig:config16}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/05-75-045-4/Inner.eps}
\caption{Configuration 16 - Inner Bar}
\label{fig:config16i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/05-75-05-35/Orbit.eps}
\caption{Configuration 17}
\label{fig:config17}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/05-75-05-35/Inner.eps}
\caption{Configuration 17 - Inner Bar}
\label{fig:config17i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/05-75-05-3/Orbit.eps}
\caption{Configuration 18}
\label{fig:config18}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/05-75-05-3/Inner.eps}
\caption{Configuration 18 - Inner Bar}
\label{fig:config18i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/06-8-06-25/Orbit.eps}
\caption{Configuration 19}
\label{fig:config19}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/06-8-06-25/Inner.eps}
\caption{Configuration 19 - Inner Bar}
\label{fig:config19i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/08-9-08-15/Orbit.eps}
\caption{Configuration 20}
\label{fig:config20}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/08-9-08-15/Inner.eps}
\caption{Configuration 20 - Inner Bar}
\label{fig:config20i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/09-95-09-1/Orbit.eps}
\caption{Configuration 21}
\label{fig:config21}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/09-95-09-1/Inner.eps}
\caption{Configuration 21 - Inner Bar}
\label{fig:config21i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-05/Orbit.eps}
\caption{Configuration 22}
\label{fig:config22}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-05/Inner.eps}
\caption{Configuration 22 - Inner Bar}
\label{fig:config22i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-04/Orbit.eps}
\caption{Configuration 23}
\label{fig:config23}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-04/Inner.eps}
\caption{Configuration 23 - Inner Bar}
\label{fig:config23i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-03/Orbit.eps}
\caption{Configuration 24}
\label{fig:config24}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-03/Inner.eps}
\caption{Configuration 24 - Inner Bar}
\label{fig:config24i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-02/Orbit.eps}
\caption{Configuration 25}
\label{fig:config25}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-02/Inner.eps}
\caption{Configuration 25 - Inner Bar}
\label{fig:config25i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/12-105-11-015/Orbit.eps}
\caption{Configuration 26}
\label{fig:config26}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/12-105-11-015/Inner.eps}
\caption{Configuration 26 - Inner Bar}
\label{fig:config26i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/12-11-11-015/Orbit.eps}
\caption{Configuration 27}
\label{fig:config27}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/12-11-11-015/Inner.eps}
\caption{Configuration 27 - Inner Bar}
\label{fig:config27i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/12-11-115-015/Orbit.eps}
\caption{Configuration 28}
\label{fig:config28}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/12-11-115-015/Inner.eps}
\caption{Configuration 28 - Inner Bar}
\label{fig:config28i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/12-11-12-015/Orbit.eps}
\caption{Configuration 29}
\label{fig:config29}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/12-11-12-015/Inner.eps}
\caption{Configuration 29 - Inner Bar}
\label{fig:config29i}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-1/Orbit.eps}
\caption{Configuration 30}
\label{fig:config30}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/1-1-1-1/Inner.eps}
\caption{Configuration 30 - Inner Bar}
\label{fig:config30i}
\end{figure}

\begin{thebibliography}{1}
\bibitem[Erwin(2008)]{erwin1}Erwin, P., Double-Barred Galaxies, 2008, Mem. S.A.It. Vol. 75, 282
\bibitem[Du et. al.(2015)]{du}Du, M., Shen, J., Debattista, V., Forming Double-Barred Galaxies from Dynamically Cool Inner Disks, 2015, The Astrophysical Journal, 804:139 (10pp), 2015 May 10
\bibitem[Wozniak(2015)]{wozniak}Wozniak, H., How can double-barred galaxies be long-lived?, 2015, A and A 575, A7 (2015)
\bibitem[Debattista and Shen(2007)]{debattista}Debattista, V., Shen, J., Long-Lived Double-Barred Galaxies from Psuedobulges, 2007, The Astrophysical Journal, 654: L127–L130, 2007 January 10
\bibitem[Garzon(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1
\bibitem[Alvarez–Ramirez and Medina(2014)]{alvarez}Alvarez–Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014, TODO
\bibitem[Szell et. al.(2004)]{szell}Szell, A., Erdi, B., Sandor, Zs., Steves, B., Chaotic and stable behaviour in the Caledonian Symmetric Four-Body Problem, 2004, Mon. Not. R. Astron. Soc. 347, 380Ð388 (2004)
\bibitem[Steves and Roy(1998)]{steves}Steves, B., Roy, A., Some special restricted four-body problems-I. Modelling the Caledonian problem, 1998, Phrt. Spucx, Si., Vol. 46. No. I I /) 12, pp. 1465-1474, 1998
\bibitem[Shen and Debattista(2009)]{shen}Shen, J., Debattista, V., Observable Properties of Double-Barred Galaxies in N-Body Simulations, 2009, The Astrophysical Journal, 690:758–772, 2009 January 1
\bibitem[Erwin and Sparke(2002)]{erwin2}Erwin, P., Sparke, L., Double Bars, Inner Disks, and Nuclear Rings in Early-Type Disk Galaxies, 2002, The Astronomical Journal, 124:65–77, 2002 July
\bibitem[Binney and Tremaine(2008)]{binney}Binney, J., Tremaine, S., Galactic Dynamics, 2008, Princeton University Press
\bibitem[Aarseth(2003)]{aarseth}Aarseth, S., Gravitational N-Body Simulations, 2003, Cambridge University Press
\bibitem[Heggie and Hut(2002)]{heggie}Heggie D., Hut, P., The Gravitational Million-Body Problem, 2002, TODO
\bibitem[Erwin(2011)]{erwin3}Erwin, P., Double-Barred Galaxies, 2011, Mem. S.A.It. Suppl. Vol. 18, 145
\bibitem[Maciejewski(2003)]{macie1}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1
\bibitem[Maciejewski and Athanassoula(2008)]{macie1}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
\bibitem[Maciejewski and Small(2010)]{macie2}Maciejewski, W., Small, E., Orbital Support of Fast and Slow Inner Bars in Double Barred Galaxies, 2010, arXiv:1006.4574v1
\bibitem[Maciejewski and Sparke(1998)]{macie3}Maciejewski, W., Sparke, L., Bars within Bars in Galaxies, 1998, arXiv:astro-ph/9812228v1
\bibitem[Maciejewski and Sparke(1999)]{macie4}Maciejewski, W., Sparke, L., Orbits Supporting Bars within Bars, 1999, arXiv:astro-ph/9911281v1
\bibitem[Moulton(1914)]{moulton}Moulton, F., An Introduction to Celestial Mechanics, 1914, Palmyrin Library TODO
\bibitem[Trenti and Hut(2008)]{trenti}Trenti, M., Hut, P., Gravitational N-body Simulations, 2008, arXiv:0806.3950v1
\bibitem[Malhotra(1998)]{malhotra}Malhotra, R., Orbital Resonances and Chaos in the Solar System, 1998, Solar system Formation and Evolution ASP Conference Series, Vol. 149, 1998
\bibitem[Manos and Athanassoula(2011)]{manos}Manos, T., Athanassoula, E., Regular and chaotic orbits in barred galaxies - I. Applying the SALI/GALI method to explore their distribution in several models, 2011, arXiv:1102.1157v2
\bibitem[Saha and Maciejewski(2013)]{saha}Saha, K., Maciejewski, W., Spontaneous formation of double bars in dark matter dominated galaxies, 2013, arXiv:1304.7108v1
\bibitem[Sellwood and Wilkinson(2006)]{sellwood}Sellwood, J., Wilkinson, A., Dynamics of Barred Galaxies, 2006, arXiv:astro-ph/0608665v1
\bibitem[Collins(2004)]{collins}Collins, G., The Foundations Of Celestial Mechanics, 2004, Pachart Publishing House TODO
\end{thebibliography}
\end{document}

### 2017-07-15 workspace/DoubleBars/2017results/09-1-106-001/DoubleBars_v2.f90

@@ -1,9 +1,9 @@
  program doublebars
    implicit none
    integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
-   integer :: kmax=0,i,j,k
-   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
-  real :: period(4, KMAXX)
+   integer :: kmax=0,i,j, period1, period2, period3, period4
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX), initialx1,initialx2
+    real :: initialx3,initialx4
    real, parameter :: PI=4*ATAN(1.)
 
    call initialconditions
@@ -11,55 +11,51 @@
 
    do i=1,kmax
       call calcenergy
-      call calcperiod
-      write(13,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i), (period(k,i), k=1,4)
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
       if(xp(i)==end) then
           exit
       end if
    end do
- 33 format(17f10.4,i5,f10.4, 4f10.4)
+ 33 format(17f10.4,i5,f10.4)
 
  contains
 
    subroutine initialconditions
 
-     step = 1.e-7 !vary step size
+     step = 1.e-8 !vary step size
      start=0.
-     end=2.22 !vary end time
+     end=1 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
-      m3=0.1    !black
-      m4=0.1   !yellow
+      m3=0.09    !black
+      m4=0.09   !yellow
 
      ys(1) = -.5     ! m1,x
+     initialx1=ys(1)
      ys(2) = 0  ! m1,vx
      ys(3) = 0       ! m1,y
      ys(4) = -1		!m1,vy
      
      ys(5) = .5         ! m2,x
+    initialx2=ys(5)
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
      ys(8) = 1	!m2,vy
 
-     ys(9) = -.1     ! m3,x
+     ys(9) = -.106     ! m3,x
+    initialx3=ys(9)
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
[…]

### 2017-07-15 workspace/DoubleBars/2017results/1-1-1-1/DoubleBars_v2.f90

@@ -1,9 +1,8 @@
  program doublebars
    implicit none
    integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
-   integer :: kmax=0,i,j,k
+   integer :: kmax=0,i,j
    real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
-  real :: period(4, KMAXX)
    real, parameter :: PI=4*ATAN(1.)
 
    call initialconditions
@@ -11,13 +10,12 @@
 
    do i=1,kmax
       call calcenergy
-      call calcperiod
-      write(13,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i), (period(k,i), k=1,4)
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
       if(xp(i)==end) then
           exit
       end if
    end do
- 33 format(17f10.4,i5,f10.4, 4f10.4)
+ 33 format(17f10.4,i5,f10.4)
 
  contains
 
@@ -25,7 +23,7 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=2.22 !vary end time
+     end=25 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
@@ -53,13 +51,6 @@
      ys(16) = .1		!m4, vy
 
    end subroutine initialconditions
-
-  subroutine calcperiod()
-    period(1,i) = ATAN(yp(3,i)/yp(1,i))
-    period(2,i) = ATAN(yp(7,i)/yp(5,i))
-    period(3,i) = ATAN(yp(11,i)/yp(9,i))
-    period(4,i) = ATAN(yp(15,i)/yp(13,i))
-  end subroutine calcperiod
 
    subroutine derivs(x,y,dydx)
      real, intent(in) :: x,y(NVAR)

### 2017-07-15 workspace/DoubleBars/2017results/1-1-11-001/DoubleBars_v2.f90

@@ -1,9 +1,9 @@
  program doublebars
    implicit none
    integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
-   integer :: kmax=0,i,j,k
-   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
-  real :: period(4, KMAXX)
+   integer :: kmax=0,i,j, period1, period2, period3, period4
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX), initialx1,initialx2
+    real :: initialx3,initialx4
    real, parameter :: PI=4*ATAN(1.)
 
    call initialconditions
@@ -11,13 +11,12 @@
 
    do i=1,kmax
       call calcenergy
-      call calcperiod
-      write(13,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i), (period(k,i), k=1,4)
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
       if(xp(i)==end) then
           exit
       end if
    end do
- 33 format(17f10.4,i5,f10.4, 4f10.4)
+ 33 format(17f10.4,i5,f10.4)
 
  contains
 
@@ -25,7 +24,7 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=2.22 !vary end time
+     end=.5 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
@@ -33,33 +32,30 @@
       m4=0.1   !yellow
 
      ys(1) = -.5     ! m1,x
+     initialx1=ys(1)
      ys(2) = 0  ! m1,vx
      ys(3) = 0       ! m1,y
      ys(4) = -1		!m1,vy
      
      ys(5) = .5         ! m2,x
+    initialx2=ys(5)
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
      ys(8) = 1	!m2,vy
 
-     ys(9) = -.1     ! m3,x
+     ys(9) = -.11     ! m3,x
+    initialx3=ys(9)
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
-     ys(12) = -.1		!m3, vy
+     ys(12) = -.001		!m3, vy
 
[…]

### 2017-07-15 workspace/DoubleBars/2017results/keep1/DoubleBars_v2.f90

@@ -1,9 +1,9 @@
  program doublebars
    implicit none
    integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
-   integer :: kmax=0,i,j,k
-   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
-  real :: period(4, KMAXX)
+   integer :: kmax=0,i,j, period1, period2, period3, period4
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX), initialx1,initialx2
+    real :: initialx3,initialx4
    real, parameter :: PI=4*ATAN(1.)
 
    call initialconditions
@@ -11,13 +11,12 @@
 
    do i=1,kmax
       call calcenergy
-      call calcperiod
-      write(13,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i), (period(k,i), k=1,4)
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
       if(xp(i)==end) then
           exit
       end if
    end do
- 33 format(17f10.4,i5,f10.4, 4f10.4)
+ 33 format(17f10.4,i5,f10.4)
 
  contains
 
@@ -25,41 +24,38 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=2.22 !vary end time
+     end=20 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
-      m3=0.1    !black
-      m4=0.1   !yellow
+      m3=0.09    !black
+      m4=0.09   !yellow
 
      ys(1) = -.5     ! m1,x
+     initialx1=ys(1)
      ys(2) = 0  ! m1,vx
      ys(3) = 0       ! m1,y
      ys(4) = -1		!m1,vy
      
      ys(5) = .5         ! m2,x
+    initialx2=ys(5)
      ys(6) = 0          ! m2,vx
      ys(7) = 0         ! m2,y
      ys(8) = 1	!m2,vy
 
-     ys(9) = -.1     ! m3,x
+     ys(9) = -.106     ! m3,x
+    initialx3=ys(9)
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
-     ys(12) = -.1		!m3, vy
[…]

### 2017-07-19 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Resonance!

@@
 \author{Howard Kinsman}
-\title{MSc Project - Modelling a Double Barred Galaxy as a Double Binary}
+\title{MSc Project - Modelling a Double-Barred Galaxy as a Double Binary}
 \begin{document}
@@
 \section{Introduction}
-The aim of this project was to create a simple model of a double barred galaxy using a double binary.
+\subsection{Double-Barred Galaxies}
+Double-barred galaxies, where a smaller inner bar is nested inside a larger outer bar, were first observed in 1975 by de Vaucouleurs, however it was only in the 1990s that double-barred galaxies
+were recognsied as being a distinct category of galaxies.
+The current estimates of double-barred galaxy frequencies are $\approx30\%$ of barred galaxies and $\approx20\%$ of all galaxies \citep{erwin3}.
+\cite{erwin2} surveyed 38 barred galaxies and found a total of 10 double-barred galaxies with no preference for Hubble type. They found that inner bars were about 12\% the size of the outer bars with
+semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \cite{erwin2}.
+
+\subsection{N-Body Simulations}
+
+\subsection{Orbits and Resonances}
 
@@
 \label{fig:config30i}
-\end{figure}
+\end{figure}l
 
+\newpage
+\section{Disussion}
+TODO
+
+\newpage
 \begin{thebibliography}{1}
-\bibitem[Erwin(2008)]{erwin1}Erwin, P., Double-Barred Galaxies, 2008, Mem. S.A.It. Vol. 75, 282
+\bibitem[Aarseth(2003)]{aarseth}Aarseth, S., Gravitational N-Body Simulations, 2003, Cambridge University Press
+\bibitem[Alvarez–Ramirez and Medina(2014)]{alvarez}Alvarez–Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014, TODO
+\bibitem[Binney and Tremaine(2008)]{binney}Binney, J., Tremaine, S., Galactic Dynamics, 2008, Princeton University Press
+\bibitem[Collins(2004)]{collins}Collins, G., The Foundations Of Celestial Mechanics, 2004, Pachart Publishing House TODO
+\bibitem[Debattista and Shen(2007)]{debattista}Debattista, V., Shen, J., Long-Lived Double-Barred Galaxies from Psuedobulges, 2007, The Astrophysical Journal, 654: L127–L130, 2007 January 10
+
+Kimematic PROPERTIES OF DOUBLE-BARRED GALAXIES: SIMULATIONS VS. INTEGRAL-FIELD
+OBSERVATIONS
 \bibitem[Du et. al.(2015)]{du}Du, M., Shen, J., Debattista, V., Forming Double-Barred Galaxies from Dynamically Cool Inner Disks, 2015, The Astrophysical Journal, 804:139 (10pp), 2015 May 10 
-\bibitem[Wozniak(2015)]{wozniak}Wozniak, H., How can double-barred galaxies be long-lived?, 2015, A and A 575, A7 (2015)
-\bibitem[Debattista and Shen(2007)]{debattista}Debattista, V., Shen, J., Long-Lived Double-Barred Galaxies from Psuedobulges, 2007, The Astrophysical Journal, 654: L127–L130, 2007 January 10
+\bibitem[Erwin(2011)]{erwin3}Erwin, P., Double-Barred Galaxies, 2011, Mem. S.A.It. Suppl. Vol. 18, 145
+\bibitem[Erwin and Sparke(2002)]{erwin2}Erwin, P., Sparke, L., Double Bars, Inner Disks, and Nuclear Rings in Early-Type Disk Galaxies, 2002, The Astronomical Journal, 124:65–77, 2002 July
 \bibitem[Garzon(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1
-\bibitem[Alvarez–Ramirez and Medina(2014)]{alvarez}Alvarez–Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014, TODO
+\bibitem[Heggie and Hut(2002)]{heggie}Heggie D., Hut, P., The Gravitational Million-Body Problem, 2002, TODO
+\bibitem[Maciejewski(2003)]{macie1}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1
+\bibitem[Maciejewski and Athanassoula(2008)]{macie1}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
+\bibitem[Maciejewski and Small(2010)]{macie2}Maciejewski, W., Small, E., Orbital Support of Fast and Slow Inner Bars in Double Barred Galaxies, 2010, arXiv:1006.4574v1
+\bibitem[Maciejewski and Sparke(1998)]{macie3}Maciejewski, W., Sparke, L., Bars within Bars in Galaxies, 1998, arXiv:astro-ph/9812228v1
+\bibitem[Maciejewski and Sparke(1999)]{macie4}Maciejewski, W., Sparke, L., Orbits Supporting Bars within Bars, 1999, arXiv:astro-ph/9911281v1
+\bibitem[Malhotra(1998)]{malhotra}Malhotra, R., Orbital Resonances and Chaos in the Solar System, 1998, Solar system Formation and Evolution ASP Conference Series, Vol. 149, 1998
+\bibitem[Manos and Athanassoula(2011)]{manos}Manos, T., Athanassoula, E., Regular and chaotic orbits in barred galaxies - I.
+Applying the SALI/GALI method to explore their distribution in several models, 2011, arXiv:1102.1157v2
+\bibitem[Moulton(1914)]{moulton}Moulton, F., An Introduction to Celestial Mechanics, 1914, Palmyrin Library TODO
+\bibitem[Saha and Maciejewski(2013)]{saha}Saha, K., Maciejewski, W., Spontaneous formation of double bars in dark matter dominated galaxies, 2013, arXiv:1304.7108v1
+\bibitem[Sellwood and Wilkinson(2006)]{sellwood}Sellwood, J., Wilkinson, A., Dynamics of Barred Galaxies, 2006, arXiv:astro-ph/0608665v1
+\bibitem[Shen and Debattista(2008)]{shen1}Shen, J., Debattista, V., Long-lived double-barred galaxies in N-body simulations, 2008, Mem. S.A.It. Vol. 00, 169
+\bibitem[Shen and Debattista(2009)]{shen2}Shen, J., Debattista, V., Observable Properties of Double-Barred Galaxies in N-Body Simulations, 2009, The Astrophysical Journal, 690:758–772, 2009 January 1
+
+\bibitem[Steves and Roy(1998)]{steves}Steves, B., Roy, A., Some special restricted four-body problems-I.
+Modelling the Caledonian problem, 1998, Phrt.
+Spucx, Si., Vol.
+46.
+No.
+I I /) 12, pp.
+1465-1474, 1998
 \bibitem[Szell et.
@@
 347, 380Ð388 (2004)
-\bibitem[Steves and Roy(1998)]{steves}Steves, B., Roy, A., Some special restricted four-body problems-I.
-Modelling the Caledonian problem, 1998, Phrt.
-Spucx, Si., Vol.
-46.
-No.
-I I /) 12, pp.
-1465-1474, 1998
-\bibitem[Shen and Debattista(2009)]{shen}Shen, J., Debattista, V., Observable Properties of Double-Barred Galaxies in N-Body Simulations, 2009, The Astrophysical Journal, 690:758–772, 2009 January 1
-\bibitem[Erwin and Sparke(2002)]{erwin2}Erwin, P., Sparke, L., Double Bars, Inner Disks, and Nuclear Rings in Early-Type Disk Galaxies, 2002, The Astronomical Journal, 124:65–77, 2002 July
-\bibitem[Binney and Tremaine(2008)]{binney}Binney, J., Tremaine, S., Galactic Dynamics, 2008, Princeton University Press
-\bibitem[Aarseth(2003)]{aarseth}Aarseth, S., Gravitational N-Body Simulations, 2003, Cambridge University Press
-\bibitem[Heggie and Hut(2002)]{heggie}Heggie D., Hut, P., The Gravitational Million-Body Problem, 2002, TODO
-\bibitem[Erwin(2011)]{erwin3}Erwin, P., Double-Barred Galaxies, 2011, Mem. S.A.It. Suppl. Vol. 18, 145
-\bibitem[Maciejewski(2003)]{macie1}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1
-\bibitem[Maciejewski and Athanassoula(2008)]{macie1}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
-\bibitem[Maciejewski and Small(2010)]{macie2}Maciejewski, W., Small, E., Orbital Support of Fast and Slow Inner Bars in Double Barred Galaxies, 2010, arXiv:1006.4574v1
-\bibitem[Maciejewski and Sparke(1998)]{macie3}Maciejewski, W., Sparke, L., Bars within Bars in Galaxies, 1998, arXiv:astro-ph/9812228v1
-\bibitem[Maciejewski and Sparke(1999)]{macie4}Maciejewski, W., Sparke, L., Orbits Supporting Bars within Bars, 1999, arXiv:astro-ph/9911281v1
-\bibitem[Moulton(1914)]{moulton}Moulton, F., An Introduction to Celestial Mechanics, 1914, Palmyrin Library TODO
 \bibitem[Trenti and Hut(2008)]{trenti}Trenti, M., Hut, P., Gravitational N-body Simulations, 2008, arXiv:0806.3950v1
-\bibitem[Malhotra(1998)]{malhotra}Malhotra, R., Orbital Resonances and Chaos in the Solar System, 1998, Solar system Formation and Evolution ASP Conference Series, Vol. 149, 1998
-\bibitem[Manos and Athanassoula(2011)]{manos}Manos, T., Athanassoula, E., Regular and chaotic orbits in barred galaxies - I.
-Applying the SALI/GALI method to explore their distribution in several models, 2011, arXiv:1102.1157v2
-\bibitem[Saha and Maciejewski(2013)]{saha}Saha, K., Maciejewski, W., Spontaneous formation of double bars in dark matter dominated galaxies, 2013, arXiv:1304.7108v1
-\bibitem[Sellwood and Wilkinson(2006)]{sellwood}Sellwood, J., Wilkinson, A., Dynamics of Barred Galaxies, 2006, arXiv:astro-ph/0608665v1
-\bibitem[Collins(2004)]{collins}Collins, G., The Foundations Of Celestial Mechanics, 2004, Pachart Publishing House TODO
+\bibitem[Wozniak(2015)]{wozniak}Wozniak, H., How can double-barred galaxies be long-lived?, 2015, A and A 575, A7 (2015)
 \end{thebibliography}

### 2017-07-19 workspace/DoubleBars/2017results/1-1-102-1/DoubleBars_v2.f90

@@ -24,7 +24,7 @@
 
      step = 1.e-7 !vary step size
      start=0.
-     end=35 !vary end time
+     end=25 !vary end time
 
       m1=.5   !blue
       m2=.5  !red
@@ -41,12 +41,12 @@
      ys(7) = 0         ! m2,y
      ys(8) = 1	!m2,vy
 
-     ys(9) = -.1021    ! m3,x
+     ys(9) = -.102    ! m3,x
      ys(10) = 0       ! m3,vx
      ys(11) = 0        ! m3,y
      ys(12) = -.1		!m3, vy
 
-     ys(13) = .1021     ! m4,x
+     ys(13) = .102     ! m4,x
      ys(14) = 0      ! m4,vx
      ys(15) = 0        ! m4,y
      ys(16) = .1		!m4, vy

### 2017-07-19 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Refs

@@
    29 & .12 & .12 & -1.1 & 1.1 & -.12 & .12 & -.015 & .015\\
-   30 & .1 & .1 & -1 & 1 &-1. & .1 & -.1 & .1\\
+   30 & .1 & .1 & -1 & 1 & -.1 & .1 & -.1 & .1\\
+   31 & .1 & .1 & -1 & 1 & -.102 & .102 & -.1 & .1\\
+   31 & .1 & .1 & -1 & 1 & -.1021 & .1021 & -.1 & .1\\
   \end{tabular}
@@
 \label{fig:config30i}
-\end{figure}l
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./2017results/1-1-102-1/Orbit.eps}
+\caption{Configuration 31}
+\label{fig:config31}
+\end{figure}
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./2017results/1-1-102-1/Inner.eps}
+\caption{Configuration 31 - Inner Bar}
+\label{fig:config31i}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./2017results/1-1-1021-1/Orbit.eps}
+\caption{Configuration 32}
+\label{fig:config32}
+\end{figure}
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./2017results/1-1-1021-1/Inner.eps}
+\caption{Configuration 32 - Inner Bar}
+\label{fig:config32i}
+\end{figure}
 
@@
 \section{Disussion}
-TODO
+The final three runs (configurations 30, 31, and 32) were by far the most stable. The orbital periods vary slightly with time so it is difficult to give accurate figures, so
+I am using the periods of the first orbit. The following table shows the periods of the two binaries together with the ratios of the corresponding periods.
+\begin{table}[ht!]
+  \centering
+  \caption{Periods}
+  \label{tab:periods}
+  \begin{tabular}{ccccc}
+   Config. & Inner & Outer & Ratio & Length of simulation\\
+    \hline
+   30 & .51 & 11.56 & 1:22.7 & 25\\
+   31 & .54 & 11.5 & 1:21.3 & 25\\
+   32 & .55 & 11.5 & 1:20.9 & 35\\
+  \end{tabular}
+\end{table}
+The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilises. Run 32 was by far the most stable run and the orbits persisted until the integrator
+errored with a 'Step size underflow'. 1:21 could possibly be an orbital resonance.
 
@@
 \bibitem[Debattista and Shen(2007)]{debattista}Debattista, V., Shen, J., Long-Lived Double-Barred Galaxies from Psuedobulges, 2007, The Astrophysical Journal, 654: L127–L130, 2007 January 10
-
-Kimematic PROPERTIES OF DOUBLE-BARRED GALAXIES: SIMULATIONS VS. INTEGRAL-FIELD
-OBSERVATIONS
-\bibitem[Du et. al.(2015)]{du}Du, M., Shen, J., Debattista, V., Forming Double-Barred Galaxies from Dynamically Cool Inner Disks, 2015, The Astrophysical Journal, 804:139 (10pp), 2015 May 10 
-\bibitem[Erwin(2011)]{erwin3}Erwin, P., Double-Barred Galaxies, 2011, Mem. S.A.It. Suppl. Vol. 18, 145
-\bibitem[Erwin and Sparke(2002)]{erwin2}Erwin, P., Sparke, L., Double Bars, Inner Disks, and Nuclear Rings in Early-Type Disk Galaxies, 2002, The Astronomical Journal, 124:65–77, 2002 July
+\bibitem[Du et. al.(2015)]{du1}Du, M., Shen, J., Debattista, V., Forming Double-Barred Galaxies from Dynamically Cool Inner Disks, 2015, The Astrophysical Journal, 804:139 (10pp), 2015 May 10 
+\bibitem[Du et. al.(2016)]{du2}Du, M., Debattista, V., Shen, J., Kimematic Properties of Double-Barred Galaxies: Simulations vs. Integral-Field Observations, 2016, arXiv:1607.00585v1
+\bibitem[Erwin(2003)]{erwin3}Erwin, P., Double-Barred Galaxies I. A Catalog of Barred Galaxies with Stellar Secondary Bars and Inner Disks, 2003, arXiv:astro-ph/0310806v2
+\bibitem[Erwin(2011)]{erwin2}Erwin, P., Double-Barred Galaxies, 2011, Mem. S.A.It. Suppl. Vol. 18, 145
+\bibitem[Erwin and Sparke(2002)]{erwin1}Erwin, P., Sparke, L., Double Bars, Inner Disks, and Nuclear Rings in Early-Type Disk Galaxies, 2002, The Astronomical Journal, 124:65–77, 2002 July
 \bibitem[Garzon(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1
 \bibitem[Heggie and Hut(2002)]{heggie}Heggie D., Hut, P., The Gravitational Million-Body Problem, 2002, TODO
-\bibitem[Maciejewski(2003)]{macie1}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1
-\bibitem[Maciejewski and Athanassoula(2008)]{macie1}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
-\bibitem[Maciejewski and Small(2010)]{macie2}Maciejewski, W., Small, E., Orbital Support of Fast and Slow Inner Bars in Double Barred Galaxies, 2010, arXiv:1006.4574v1
-\bibitem[Maciejewski and Sparke(1998)]{macie3}Maciejewski, W., Sparke, L., Bars within Bars in Galaxies, 1998, arXiv:astro-ph/9812228v1
-\bibitem[Maciejewski and Sparke(1999)]{macie4}Maciejewski, W., Sparke, L., Orbits Supporting Bars within Bars, 1999, arXiv:astro-ph/9911281v1
+\bibitem[Lorenzo-Caceres et.
+al.(2002)]{lorenzo1}de Lorenzo-Caceres, A., Vazdekis, A., Aguerri, K., A., L., Corsini, E., M., Debattista, V., P., Constraining the formation of inner bars.
+Photometry, kinematics and stellar populations in NGC357⋆, 2002, arXiv:1111.1718v1
+\bibitem[Lorenzo-Caceres et.
+al.(2013)]{lorenzo2}de Lorenzo-Caceres, A., Falcon-Barroso, J., Vazdekis, A., Distinct stellar populations in the inner bars of double-barred galaxies, 2013, arXiv:1302.5701v1
+\bibitem[Maciejewski(2002)]{macie1}Maciejewski, W., Constraints on nested bars – implications for gas inflow, 2002, arXiv:astro-ph/0202110v1
+\bibitem[Maciejewski(2003)]{macie2}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1
+\bibitem[Maciejewski(2008)]{macie3}Maciejewski, W., Orbits in corotating and counterrotating double bars, 2008, arXiv:0801.1471v1
+\bibitem[Maciejewski and Athanassoula(2008)]{macie4}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
+\bibitem[Maciejewski and Small(2010)]{macie5}Maciejewski, W., Small, E., Orbital Support of Fast and Slow Inner Bars in Double Barred Galaxies, 2010, arXiv:1006.4574v1
+\bibitem[Maciejewski and Sparke(1998)]{macie6}Maciejewski, W., Sparke, L., Bars within Bars in Galaxies, 1998, arXiv:astro-ph/9812228v1
+\bibitem[Maciejewski and Sparke(1999)]{macie7}Maciejewski, W., Sparke, L., Orbits Supporting Bars within Bars, 1999, arXiv:astro-ph/9911281v1
 \bibitem[Malhotra(1998)]{malhotra}Malhotra, R., Orbital Resonances and Chaos in the Solar System, 1998, Solar system Formation and Evolution ASP Conference Series, Vol. 149, 1998
@@
 Applying the SALI/GALI method to explore their distribution in several models, 2011, arXiv:1102.1157v2
+\bibitem[Moiseev(2001)]{moiseev}Moiseev, A., V., Velocity dispersion of stars and gas motion in double-barred galaxies, 2001, arXiv:astro-ph/0111220v1
 \bibitem[Moulton(1914)]{moulton}Moulton, F., An Introduction to Celestial Mechanics, 1914, Palmyrin Library TODO
@@
 \bibitem[Shen and Debattista(2009)]{shen2}Shen, J., Debattista, V., Observable Properties of Double-Barred Galaxies in N-Body Simulations, 2009, The Astrophysical Journal, 690:758–772, 2009 January 1
-
 \bibitem[Steves and Roy(1998)]{steves}Steves, B., Roy, A., Some special restricted four-body problems-I.

### 2017-07-20 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Refs2

@@
 \end{abstract}
-{\bf Keywords:} TODO
 \section{Introduction}
@@
 
-\subsection{N-Body Simulations}
+\citep{erwin3}, \citep{erwin2}, \citep{erwin1}, \citep{lorenzo1}, \citep{lorenzo2}, \citep{moiseev}, \citep{saha}, \citep{sellwood}
 
-\subsection{Orbits and Resonances}
+\subsection{N-Body Simulations and the Four-Body Problem}
+\citep{aarseth}, \citep{alvarez}, \citep{heggie}, \citep{binney}, \citep{collins}, \citep{moulton}, \citep{steves}, \citep{szell}, \citep{trenti}
+
+\subsection{Orbits, Loops, Resonances and Double-Barred Galaxy Simulations}
+\citep{debattista}, \citep{du1}, \citep{du2}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}
+\citep{macie5}, \citep{macie6}, \citep{macie7}, \citep{malhotra}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{wozniak}
 
@@
 \end{figure}
-
+l
 \begin{figure}[H]
@@
 
+\citep{garzon}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}, \citep{macie5}
+\citep{macie6}, \citep{macie7}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{debattista},
+
 \newpage

### 2017-08-13 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
latest

@@
 Double-barred galaxies, where a smaller inner bar is nested inside a larger outer bar, were first observed in 1975 by de Vaucouleurs, however it was only in the 1990s that double-barred galaxies
-were recognsied as being a distinct category of galaxies.
+were recognised as being a distinct category of galaxies.
 The current estimates of double-barred galaxy frequencies are $\approx30\%$ of barred galaxies and $\approx20\%$ of all galaxies \citep{erwin3}.
@@
 semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \cite{erwin2}.
+\cite{erwin1} believe that innder bars probably rotate independently of their outer bars and suggest that they are relatively long-lived strucutures.
+Inner bar size has been found to correlate slightly with outer bar size \citep{erwin3}.
+The median size ratio is $\approx0.12$ with an upper limit of $\approx0.25$ which is consistent with the theory that
+inner bars cannot be too large without disrupting the orbit of the outer bar.
 
-\citep{erwin3}, \citep{erwin2}, \citep{erwin1}, \citep{lorenzo1}, \citep{lorenzo2}, \citep{moiseev}, \citep{saha}, \citep{sellwood}
+It has been suggested that nested bars may provide a mechanism for fueling active galactic nuclei (Shlosman 1989) however more recently \citep{erwin3} believe that inner bars only play a minor role in this
+activity.
+\citep{erwin1} compared single and double-barred galaxies and found that there was no significant difference in nuclear activity.
+To investigate this theory further \cite{lorenzo1} carried out a
+study of NGC 357.
+This study aimed to present the first detailed morhpological, kinematical and stellar population analysis of the bulge, inner and outer bars of a double-barred galaxy.
+They concluded that
+the bulge and inner bar show similar stellar population properties whilst the outer bar was less metal-rich.
+This result implied that, for the case of NGC 357, the outer bar was formed in a shorter timescale than
+the inner structures and so disagrees with traditional idea that gas flown along the outer structure triggers star formation causing the formation of the inner structure.
+This was strengthened by a further study of
+four double-barred galaxies by \cite{lorenzo2} where they concluded that the stellar populations of inner bars are younger and more metal-rich than the outer bars.
+This suggests that at present inner bars play a
+moderate or even minor role in the morphological evolution of double-barred galaxies \citep{lorenzo2}.
+
+Through the use of computer modeeling \cite{saha} showed that double bars can form without gas in a dark matter dominated halo.
+They produced a model where the inner bar was rotating at almost as slowly as the
+outer bar. The route to the formation of double bars maybe very different to that of a single bar.
+
+Photos!
+
+ \citep{moiseev}, , \citep{sellwood}
 
 \subsection{N-Body Simulations and the Four-Body Problem}
-\citep{aarseth}, \citep{alvarez}, \citep{heggie}, \citep{binney}, \citep{collins}, \citep{moulton}, \citep{steves}, \citep{szell}, \citep{trenti}
-
-\subsection{Orbits, Loops, Resonances and Double-Barred Galaxy Simulations}
-\citep{debattista}, \citep{du1}, \citep{du2}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}
-\citep{macie5}, \citep{macie6}, \citep{macie7}, \citep{malhotra}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{wozniak}
-
-\section{Methods}
-Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
-The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
-Please see the Compilers subsection for a few notes on compilers - may or may not be relevant to the project.
+Studies of three or more bodies interacting gravitationally dates back to the time of Newton.
+However it is only with the widespread use of computers that orbits of more than three bodies have been extensively studied.
+Today
+N-body simulations are an essential tool in the study of solar system dynamics and galactic dynamics \citep{binney}.
+Celestial mechanics during the preceding centuries has mainly focussed on the three body problem
+due to the complexities involved in adding more bodies.
+It is estimated that within the Milky Way approximately two thirds of all stars exist in binary systems and a further one fifth of these are in triple systems whilst a
+further fifth still exist in systems with more than three bodies \citep{steves}.
+\cite{steves} estimate that $\approx2\times10^9$ quadruple systems exist in the Milky Way galaxy.
+They found that quadruple systems exist in two
+forms: a double binary or linear.
 \newline
@@
 \end{equation}
+The units have been selected so that the gravitational constant (G) is equal to one.
+
+The four body problem, specifically, has been studied extensively by \cite{steves}.
+Here they model four equal mass bodies orbiting in circular coplanar orbits.
+This model they term the Caledonian problem, because
+their research is being carried out at Glasgow Caledonian University.
+Approximately a century ago the Copenhagen problem was studied extensively in Copenhagen which consisted of a three body problem: two equal mass bodies and an infinitesimally small test particle.
+\cite{szell} also
+studied the Caledonian problem and found that the stability of the system is dependent upon a constant termed the Szebehely constant.
+
+Methods for the study of N-body simulations have changed considerably since the advent of computers.
+In a study of N-body simulations \cite{trenti} mentions a pioneering attempt by Holmberg in 1941 which used
+light bulbs and galvanometers to track the evolution of 37 particles. Nowadays simulations can contain millions of particles \citep{heggie}.
+
+\citep{aarseth}, \citep{alvarez}, \citep{heggie}, \citep{collins}, \citep{moulton}
+
+\subsection{Orbits, Loops, Resonances and Double-Barred Galaxy Simulations}
+\citep{debattista}, \citep{du1}, \citep{du2}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}
+\citep{macie5}, \citep{macie6}, \citep{macie7}, \citep{malhotra}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{wozniak}
+
+\section{Methods}
+Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
+The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
+Please see the Compilers subsection for a few notes on compilers - may or may not be relevant to the project.
+
 The above equations were encoded into Fortran. A full code listing is supplied in a separate file.
@@
 errored with a 'Step size underflow'. 1:21 could possibly be an orbital resonance.
+Fourier analysis?
+Rewrite in Python, automate, introduce machine laerning algoriythm
 
@@
 \bibitem[Wozniak(2015)]{wozniak}Wozniak, H., How can double-barred galaxies be long-lived?, 2015, A and A 575, A7 (2015)
+Numerical Recipes
 \end{thebibliography}

### 2017-08-14 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
writeup!

--- 2017-07-19 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
@@
 \end{abstract}
-{\bf Keywords:} TODO
 \section{Introduction}
 \subsection{Double-Barred Galaxies}
-Double-barred galaxies, where a smaller inner bar is nested inside a larger outer bar, were first observed in 1975 by de Vaucouleurs, however it was only in the 1990s that double-barred galaxies
-were recognsied as being a distinct category of galaxies.
+Double-barred galaxies, where a smaller inner bar is nested inside a larger outer bar, were first observed in 1975 by de Vaucouleurs \citep{vaucouleurs}, however it was only in the 1990s that double-barred galaxies
+were recognised as being a distinct category of galaxies.
 The current estimates of double-barred galaxy frequencies are $\approx30\%$ of barred galaxies and $\approx20\%$ of all galaxies \citep{erwin3}.
@@
 semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \cite{erwin2}.
+\cite{erwin1} believe that innder bars probably rotate independently of their outer bars and suggest that they are relatively long-lived strucutures.
+Inner bar size has been found to correlate slightly with outer bar size \citep{erwin3}.
+The median size ratio is $\approx0.12$ with an upper limit of $\approx0.25$ which is consistent with the theory that
+inner bars cannot be too large without disrupting the orbit of the outer bar.
 
-\subsection{N-Body Simulations}
+\begin{figure}[H]
+\centering
+\includegraphics[width=.9\textwidth]{./eso0128d.eps}
+\caption{This is a schematic drawing of the various components of a double-barred galaxy. Credit: ESO}
+\label{fig:doublepic}
+\end{figure}
 
-\subsection{Orbits and Resonances}
+It has been suggested that nested bars may provide a mechanism for fueling active galactic nuclei \citep{shlosman} however more recently \citep{erwin3} believe that inner bars only play a minor role in this
+activity.
+\citep{erwin1} compared single and double-barred galaxies and found that there was no significant difference in nuclear activity.
+To investigate this theory further \cite{lorenzo1} carried out a
+study of NGC 357.
+This study aimed to present the first detailed morphological, kinematical and stellar population analysis of the bulge, inner and outer bars of a double-barred galaxy.
+They concluded that
+the bulge and inner bar show similar stellar population properties whilst the outer bar was less metal-rich.
+This result implied that, for the case of NGC 357, the outer bar was formed in a shorter timescale than
+the inner structures and so disagrees with traditional idea that gas flown along the outer structure triggers star formation causing the formation of the inner structure.
+This was strengthened by a further study of
+four double-barred galaxies by \cite{lorenzo2} where they concluded that the stellar populations of inner bars are younger and more metal-rich than the outer bars.
+This suggests that at present inner bars play a
+moderate or even minor role in the morphological evolution of double-barred galaxies \citep{lorenzo2}.
 
-\section{Methods}
-Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
-The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
-Please see the Compilers subsection for a few notes on compilers - may or may not be relevant to the project.
+Through the use of computer modeeling \cite{saha} showed that double bars can form without gas in a dark matter dominated halo.
+They produced a model where the inner bar was rotating at almost as slowly as the
+outer bar. The route to the formation of double bars maybe very different to that of a single bar.
+%\citep{moiseev}, \citep{sellwood}
+
+\subsection{N-Body Simulations and the Four-Body Problem}
+Studies of three or more bodies interacting gravitationally dates back to the time of Newton.
+However it is only with the widespread use of computers that orbits of more than three bodies have been extensively studied.
+Today
+N-body simulations are an essential tool in the study of solar system dynamics and galactic dynamics \citep{binney}.
+Celestial mechanics during the preceding centuries has mainly focussed on the three body problem
+due to the complexities involved in adding more bodies.
+It is estimated that within the Milky Way approximately two thirds of all stars exist in binary systems and a further one fifth of these are in triple systems whilst a
+further fifth still exist in systems with more than three bodies \citep{steves}.
+\cite{steves} estimate that $\approx2\times10^9$ quadruple systems exist in the Milky Way galaxy.
+They found that quadruple systems exist in two
+forms: a double binary or linear.
 \newline
-The classical equations of motion for an n-body problem are:
+The classical equations of motion for an N-body problem are:
 \begin{equation}
@@
 \end{equation}
-where $\mathbf{r_i}=\left(x_i,y_i\right)$ and $\mathbf{r_{ij}}=\mathbf{r_j}-\mathbf{r_i}$. So for a four body problem 
-this equation results in the following:
+where $\mathbf{r_i}=\left(x_i,y_i\right)$ and $\mathbf{r_{ij}}=\mathbf{r_j}-\mathbf{r_i}$. 
+The units have been selected so that the gravitational constant (G) is equal to one.
+In simulations it is natural to choose units where G = 1 and the units of length nand mass are conventionally chosen
+so that the total mass of the system is M = 1 \citep{heggie}.
+
+The four body problem, specifically, has been studied extensively by \cite{steves}.
+Here they model four equal mass bodies orbiting in circular coplanar orbits.
+This model they term the Caledonian problem, because
+their research is being carried out at Glasgow Caledonian University.
+Approximately a century ago the Copenhagen problem was studied extensively in Copenhagen which consisted of a three body problem: two equal mass bodies and an infinitesimally small test particle.
+\cite{szell} also
+studied the Caledonian problem and found that the stability of the system is dependent upon a constant termed the Szebehely constant.
+The Caledonian Symmetric Double Binary Problem (CSDBP) is relevant in
+studying the stability and evolution of symmetric quadruple stellar clusters and exoplanetary systems of two planets orbiting a binary pair of stars \citep{alvarez}.
+
+Methods for the study of N-body simulations have changed considerably since the advent of computers.
+In a study of N-body simulations \cite{trenti} mentions a pioneering attempt by Holmberg in 1941 which used
+light bulbs and galvanometers to track the evolution of 37 particles. However it wasn't until the early seventies and the advent of computers that this field really got started. 
+Nowadays simulations can up to a million particles \citep{heggie} which is of particular relevance to the study of globular
+clusters. I refer the reader to \cite{aarseth} for much more detailed explanation of the field of N-body simulations.
+ %\citep{collins}, \citep{moulton}
+
+\subsection{Orbits, Loops, Resonances and Double-Barred Galaxy Simulations}
+Double-barred galaxy simulations and models have taken many forms.
+\cite{du1} explored the formation and evolution of double-barred galaxies (which is not very well understood) by showing that a dynamically
+cool inner disk embedded in a hotter outer disk can generate a secondary bar while the outer disk forms a large-scale primary bar.
+\cite{macie8} used hydrodynamical simulations to study gas inflow in barred galaxies
+and the effect this has on secondary bars.
+They concluded that the secondary bar can prevent, rather than enhance, the inflow.
+More recently, \cite{debattista} used collisionless N-body simulations to show
+that inner bars can form spontaneously without requiring gas.
+They also found that secondary bars rotate faster that primary bars and that the secondary bars pulsate.
+This is considered further in the Discussion.
+
+In an attempt to explain the dynamics of the orbits of nested bars \cite{macie7} introduced the concept of `loops'. They wanted to investigate how two independently rotating bars can maintain
+themselves as gravitating systems.
+A gravitational potential consisting of two concentric bisymmetric bars that are independently rotating is not stationary in any reference frame \cite{macie7}.
+In the potential
+of a single bar particles would stay on a stable closed periodic orbit but with nested bars orbits will not be closed because the particles would be experience two separate forces.
+The `loop' therefore was
+introduced as an extension of the concept of orbit.
+Particles in a `loop' would return to their original positions every time the bars come back to the same relative orientation \citep{macie7}.
+This lead to
+the conclusion that loops supporting the inner bar are thicker with parallel bars and thinner with perpendicular bars \citep{macie3}.
+This also suggets that the two bars do not rotate through each other as rigid bodies
+but the inner bar should pulsate and that the bars should spend more time nearly orthognal\citep{macie3}. These predictions were confirmed by \citep{debattista} and \citep{shen1}. 
+This was further confirmed by \cite{macie5} who found
+that the inner bar should end well within its corotation radius and that the inner bar extends further out for lower pattern speeds than higher ones; the inner bar pulsates; and that the angular velocity of the inner
+bar is not uniform and that faster inner bars rotate more coherently.
+
+\cite{wozniak} also carried out double-barred galaxy simulations and achieved long-lived inner bars lasting for up to 7 Gyr.
+These were N-body/hydrodynamical simulations.
+They found that the two bar lengths, ratio of
+pattern speeds, as well as the age of the inner stellar bar populations in the simulations matched well with observations.
+They concluded that because, unlike previous simulations, there was no overlap between primary bar
+inner Lindblad resonance (ILR) and the inner bar corotation (CR) or any kind of resonance overlap, then some other kind of mechanism must feed the central waves \citep{wozniak}.
+It is suggested that star
+formation may be responsible for bringing energy into the inner bar.
+
+\cite{garzon} studied the dynamical evolution of two isolated bars, within the same galaxy, under their mutual gravitational interaction.
+They considered two cases purely analytically: rotation of rigid bodies and
+rotation of deformable rotation. They found that the bars oscillate in size according to the relative angle between the bars. This is broadly in agreement with \citep{debattista}.
+What follows is an attempt to verify the calculations in \citep{garzon} but using a simpler model of a double-barred galaxy using a double binary.
+ %\citep{du2}, \citep{macie1}, \citep{macie2},  \citep{macie4}
+ %\citep{malhotra}, \citep{manos}, \citep{shen2}
+
+\section{Methods}
+The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
+Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
+For a four body problem Equation 1 results in the following:
 \begin{equation}
@@
 
-\subsection{Compilers}
-As I had converted OdeInt into Fortran 90 I wanted to make sure I hadn't introduced any errors. I therefore compared the output of the Numerical Recipes
-version of OdeInt with mine. I was compiling with GFortran and the Fortran 77 version was compiled with g77. They were different! So naturally concerned 
-I then compiled the Fortran 77 version with GFortran. The results were identical to mine, which satisified me that I hadn't introduced any errors.
-
 \section{Results}
@@
 errored with a 'Step size underflow'. 1:21 could possibly be an orbital resonance.
+Fourier analysis?
+Rewrite in Python, automate, introduce machine laerning algoriythm
+
+\citep{garzon}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}, \citep{macie5}
+\citep{macie6}, \citep{macie7}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{debattista}, \citep{malhotra}
 
@@
 \bibitem[Aarseth(2003)]{aarseth}Aarseth, S., Gravitational N-Body Simulations, 2003, Cambridge University Press
-\bibitem[Alvarez–Ramirez and Medina(2014)]{alvarez}Alvarez–Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014, TODO
+\bibitem[Alvarez–Ramirez and Medina(2014)]{alvarez}Alvarez–Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014
 \bibitem[Binney and Tremaine(2008)]{binney}Binney, J., Tremaine, S., Galactic Dynamics, 2008, Princeton University Press
-\bibitem[Collins(2004)]{collins}Collins, G., The Foundations Of Celestial Mechanics, 2004, Pachart Publishing House TODO
+%\bibitem[Collins(2004)]{collins}Collins, G., The Foundations Of Celestial Mechanics, 2004, Pachart Publishing House
+\bibitem[de Vaucouleurs(1975)]{vaucouleurs}de Vaucouleurs, G., 1975, ApJS,29, 193
 \bibitem[Debattista and Shen(2007)]{debattista}Debattista, V., Shen, J., Long-Lived Double-Barred Galaxies from Psuedobulges, 2007, The Astrophysical Journal, 654: L127–L130, 2007 January 10
 \bibitem[Du et. al.(2015)]{du1}Du, M., Shen, J., Debattista, V., Forming Double-Barred Galaxies from Dynamically Cool Inner Disks, 2015, The Astrophysical Journal, 804:139 (10pp), 2015 May 10 
-\bibitem[Du et. al.(2016)]{du2}Du, M., Debattista, V., Shen, J., Kimematic Properties of Double-Barred Galaxies: Simulations vs. Integral-Field Observations, 2016, arXiv:1607.00585v1
+%\bibitem[Du et. al.(2016)]{du2}Du, M., Debattista, V., Shen, J., Kimematic Properties of Double-Barred Galaxies: Simulations vs. Integral-Field Observations, 2016, arXiv:1607.00585v1
 \bibitem[Erwin(2003)]{erwin3}Erwin, P., Double-Barred Galaxies I. A Catalog of Barred Galaxies with Stellar Secondary Bars and Inner Disks, 2003, arXiv:astro-ph/0310806v2
@@
 \bibitem[Garzon(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1
-\bibitem[Heggie and Hut(2002)]{heggie}Heggie D., Hut, P., The Gravitational Million-Body Problem, 2002, TODO
+\bibitem[Heggie and Hut(2003)]{heggie}Heggie D., Hut, P., The Gravitational Million-Body Problem, 2003, Cambridge University Press
 \bibitem[Lorenzo-Caceres et.
@@
 al.(2013)]{lorenzo2}de Lorenzo-Caceres, A., Falcon-Barroso, J., Vazdekis, A., Distinct stellar populations in the inner bars of double-barred galaxies, 2013, arXiv:1302.5701v1
-\bibitem[Maciejewski(2002)]{macie1}Maciejewski, W., Constraints on nested bars – implications for gas inflow, 2002, arXiv:astro-ph/0202110v1
-\bibitem[Maciejewski(2003)]{macie2}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1
+\bibitem[Maciejewski et. al.(2001)]{macie8}Maciejewski, W., Teuben, P., J., Sparke, L., S., Stone, J., M., Gas Inflow in Barred Galaxies - Effects of Secondary Bars, arXiv:astro-ph/0109431v1
+%\bibitem[Maciejewski(2002)]{macie1}Maciejewski, W., Constraints on nested bars – implications for gas inflow, 2002, arXiv:astro-ph/0202110v1
+%\bibitem[Maciejewski(2003)]{macie2}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1
 \bibitem[Maciejewski(2008)]{macie3}Maciejewski, W., Orbits in corotating and counterrotating double bars, 2008, arXiv:0801.1471v1
-\bibitem[Maciejewski and Athanassoula(2008)]{macie4}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
+%\bibitem[Maciejewski and Athanassoula(2008)]{macie4}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
 \bibitem[Maciejewski and Small(2010)]{macie5}Maciejewski, W., Small, E., Orbital Support of Fast and Slow Inner Bars in Double Barred Galaxies, 2010, arXiv:1006.4574v1
@@
 \bibitem[Moiseev(2001)]{moiseev}Moiseev, A., V., Velocity dispersion of stars and gas motion in double-barred galaxies, 2001, arXiv:astro-ph/0111220v1
-\bibitem[Moulton(1914)]{moulton}Moulton, F., An Introduction to Celestial Mechanics, 1914, Palmyrin Library TODO
+%\bibitem[Moulton(1914)]{moulton}Moulton, F., An Introduction to Celestial Mechanics, 1914, Palmyrin Library 
+\bibitem[Press et. al.(1992)]{press}Press, W., H., Teukolsky, S., A., Vetterling, W., T., Flannery, B., P., Numerical Recipes in Fortran 77 The Art of Scientific Computing, 2nd edition
 \bibitem[Saha and Maciejewski(2013)]{saha}Saha, K., Maciejewski, W., Spontaneous formation of double bars in dark matter dominated galaxies, 2013, arXiv:1304.7108v1
@@
 \bibitem[Shen and Debattista(2009)]{shen2}Shen, J., Debattista, V., Observable Properties of Double-Barred Galaxies in N-Body Simulations, 2009, The Astrophysical Journal, 690:758–772, 2009 January 1
+\bibitem[Shlosman et. al.(1989)]{shlosman}Shlosman, I., Frank, J., Begeman, M.,C., 1989, Nature, 338, 45
 \bibitem[Steves and Roy(1998)]{steves}Steves, B., Roy, A., Some special restricted four-body problems-I.

### 2017-08-14 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
writeup!

@@
 \section{Methods}
+Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
 The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
-Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
-For a four body problem Equation 1 results in the following:
+For a four body problem 
+this equation results in the following:
 \begin{equation}
@@
 \end{table}
+
+\subsection{Compilers}
+As I had converted OdeInt into Fortran 90 I wanted to make sure I hadn't introduced any errors. I therefore compared the output of the Numerical Recipes
+version of OdeInt with mine. I was compiling with GFortran and the Fortran 77 version was compiled with g77. They were different! So naturally concerned 
+I then compiled the Fortran 77 version with GFortran. The results were identical to mine, which satisified me that I hadn't introduced any errors.
 

### 2017-08-16 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Test
Results

--- 2017-08-14 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
@@
 \begin{abstract}
-TODO
+TODOdd
 \end{abstract}
@@
 Today
-N-body simulations are an essential tool in the study of solar system dynamics and galactic dynamics \citep{binney}.
-Celestial mechanics during the preceding centuries has mainly focussed on the three body problem
+N-body simulations are an essential tool in the study of solar system dynamics and galactic dynamics. The orbit-integration problems we have to address
+vary in complexity from following a single particle in a given, smooth galactic potential, to tens of thousands of interacting stars in a globular cluster, to billions of dark-matter particles in a 
+simulation of cosmological clustering \citep{binney}. Celestial mechanics during the preceding centuries has mainly focussed on the three body problem
 due to the complexities involved in adding more bodies.
@@
  %\citep{malhotra}, \citep{manos}, \citep{shen2}
+\newpage
+\section{Methods}
+The model was created using OdeInt from Numerical Recipes \citep{press}. I first converted OdeInt to Fortran 90 for this purpose.
+Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
+OdeInt performs integration using the Runge-Kutta method. The Runge-Kutta method is an extension of the Euler method of solving differential equations. However the Euler method
+is an example of first-order integration whilst Runge-Kutta is second and higher order and so is mcuh more accurate.
+The OdeInt uses a subroutine called RK4 which utilises fourth-order Runge-Kutta integration.
+It also uses a technique known as adaptive stepsize.
+This involves adjusting the size of the integration steps to ensure the same level of accuracy however rapidly the function being integrated changes.
+Many small steps should tiptoe through treacherous terrain, while a few great strides should speed through smooth uninteresting countryside \citep{press}.
 
-\section{Methods}
-The model was created using OdeInt from Numerical Recipes. I first converted OdeInt to Fortran 90 for this purpose.
-Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
-For a four body problem Equation 1 results in the following:
+For a four body problem, Equation 1 results in the following:
 \begin{equation}
@@
 \end{equation}
-The above equations were encoded into Fortran. A full code listing is supplied in a separate file.
+These equations of motion are second-order ordinary differential equations (ODE) but any second-oder ODE can be written as a system
+of two coupled first-order ODEs. As there are four bodies in the system and and x and y components for both of these we have a total of eight second-order 
+ODEs which then gives us sixteen first-order ODEs to solve.
+These equations were encoded into Fortran. 
+A full code listing is supplied in a separate file.
 \subsection{Initial Conditions}
 As a starting point both the outer binary and inner binary were placed in circular orbits and tested separately and then combined to ensure a stable
-starting point. The inner binary had initially a neglible mass.
+starting point. The inner binary had an initial neglible mass.
 The outer binary initial conditions were:
@@
 Plots of the two initial binaries are shown in Fig. \ref{fig:outerbinary} and 
-\ref{fig:innerbinary}. The scale of the inner binary is so small that it is difficult
-to see in GnuPlot.
+\ref{fig:innerbinary}. The velocity of the inner binary is high and therefore few points are seen in this initial plot.
 \begin{figure}[H]
@@
 \end{table}
-
+\newpage
 \section{Results}
 As can be seen from Fig. \ref{fig:config1} even a neglible mass inner binary causes the outer binary to begin to separate and become less tightly bound.
-Figures Fig. \ref{fig:config1} to \ref{fig:config18i} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
+Figures Fig. \ref{fig:config1} to \ref{fig:config32i} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
 In many cases the inner binary is only visible as a point at the centre of the outer binary.
-As the mass of the inner binary increased the system became increasingly unstablle and the effect on the outer binary became more pronounced.
-Even the 'stable' configurations listed in Table \ref{tab:variables} hold for only a few orbits.
-I was unable to recreate a stable system for a mass $>.08$ for the inner bar.
-The plots below reveal that increasing the mass of the inner binary causes the outer binary to separate and
-form wider orbits.
+As the mass of the inner binary increased the system became increasingly unstable and the effect on the outer binary became more pronounced.
+Even the `stable' configurations listed in Table \ref{tab:variables} hold for only a few orbits.
 
@@
 \end{figure}
-
 \begin{figure}[H]
@@
 \end{figure}
+\ref{fig:config1} and \ref{fig:config1i} show the inital orbits when the circular inner binary is placed within the outer binary.
 
@@
 \end{figure}
+\ref{fig:config2} to \ref{fig:config3i} show plots where just the mass of the inner binary has been gradually increased. Nothing else was changed.
+By \ref{fig:config3i}, the inner binary can clearly be seen to be pulled apart by the gravitational affect of the outer binary.
 
@@
 \end{figure}
-l
+To counteract the effect of the increased mass of the inner binary in \ref{fig:config4} and \ref{fig:config4i} the initial separation of the 
+inner binary has been increased.
+
 \begin{figure}[H]
@@
 \end{figure}
+In \ref{fig:config5} to \ref{fig:config10i} the initial velocity of the outer binary has also been gradually increased to also counteract the effect
+of the increased mass of the inner binary. In order for these systems to be relatively stable the initial velocity of the inner binary was reduced. At this point 
+both the inner and outer binaries had started to separate and become increasingly elliptical.
 
@@
 \end{figure}
+In order to bring the orbits back into approximate circularity it was necessary to again increase the velocity of the inner binary. This is illustrated
+in \ref{fig:config11} and \ref{fig:config11i}. The mass of the inner binary has also increased considerably from the previous configurations.
 
@@
 \end{table}
+Discuss mass of inner bar ration of outer bar
 The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilises. Run 32 was by far the most stable run and the orbits persisted until the integrator
@@
 \bibitem[Saha and Maciejewski(2013)]{saha}Saha, K., Maciejewski, W., Spontaneous formation of double bars in dark matter dominated galaxies, 2013, arXiv:1304.7108v1
-\bibitem[Sellwood and Wilkinson(2006)]{sellwood}Sellwood, J., Wilkinson, A., Dynamics of Barred Galaxies, 2006, arXiv:astro-ph/0608665v1
+%\bibitem[Sellwood and Wilkinson(2006)]{sellwood}Sellwood, J., Wilkinson, A., Dynamics of Barred Galaxies, 2006, arXiv:astro-ph/0608665v1
 \bibitem[Shen and Debattista(2008)]{shen1}Shen, J., Debattista, V., Long-lived double-barred galaxies in N-body simulations, 2008, Mem. S.A.It. Vol. 00, 169

### 2017-08-17 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Results

@@
 \begin{abstract}
-TODOdd
+TODO
 \end{abstract}
@@
 forms: a double binary or linear.
-\newline
+
 The classical equations of motion for an N-body problem are:
@@
 \section{Methods}
-The model was created using OdeInt from Numerical Recipes \citep{press}. I first converted OdeInt to Fortran 90 for this purpose.
+The model was created using OdeInt from Numerical Recipes \citep{press}. OdeInt was first converted to Fortran 90 for this purpose.
 Fortran 90 was selected as the language of choice in order to avoid the fixed line limitations of Fortran 77.
@@
 \end{equation}
-These equations of motion are second-order ordinary differential equations (ODE) but any second-oder ODE can be written as a system
-of two coupled first-order ODEs. As there are four bodies in the system and and x and y components for both of these we have a total of eight second-order 
-ODEs which then gives us sixteen first-order ODEs to solve.
+These equations of motion are second-order ordinary differential equations (ODEs) but any second-oder ODE can be written as a system
+of two coupled first-order ODEs. As there are four bodies in the system, and they all have x and y positional components, this gives eight second-order 
+ODEs which leaves us with a total of sixteen simultanious first-order ODEs to solve.
 These equations were encoded into Fortran. 
@@
 separation of the inner binary. In many of the configurations the systems were inherently unstable and produced collisions with some 
-or all of the bodies being ejected from the system. These were rejected and only 'stable' configurations were considered.
+or all of the bodies being ejected from the system. These were rejected and only `stable' configurations were considered.
 Table \ref{tab:variables} shows the perturbations made to the system - only the variables in this table changed and all the other initial
@@
 \end{table}
+
 \newpage
@@
 
+Figs. \ref{fig:config12} to \ref{fig:config13i} show the mass of the inner binary slowly increasing and adjustments made to both velocity of outer binary
+and separation of inner binary.
+
 \begin{figure}[H]
@@
 
+As the mass of the inner binary was increased, it became necessary to reduce the velocity of the inner binary, Figs. \ref{fig:config14} to \ref{fig:config15i}.
+
 \begin{figure}[H]
@@
 
+Holding the mass of the inner binary at .05 the system showed signs of more stability if the velocity of the inner binary was further reduced, Figs. \ref{fig:config16} to \ref{fig:config18i}.
+
 \begin{figure}[H]
@@
 
+Further reduction of the velocity of the inner binary allowed further increase of inner binary mass without losing stability, Figs. \ref{fig:config19} to \ref{fig:config21i}.
+
 \begin{figure}[H]
@@
 
+With the mass of the inner binary now a quite considerable .1, the system showed more signs of stability if the velocity of the inner binary was further reduced, Figs.
+\ref{fig:config22} to \ref{fig:config25i}.
+
 \begin{figure}[H]
@@
 \end{figure}
+
+In Figs.
+\ref{fig:config26} to \ref{fig:config29i} the mass of the inner binary was increased to .12, and various adjustments made to inner separation, velocity and outer velocity to maintain stability.
 
@@
 
+Having increased the mass of the inner binary to .12 on the previous three plots it was decided, that as inner bars are generally $\approx12\%$ of the mass of the outer bar, the mass of the inner binary could stay at .1
+as this is 20\% of the mass of the outer binary, Figs. \ref{fig:config30} and \ref{fig:config30i}.
+
 \begin{figure}[H]
@@
 \end{figure}
+
+Finally, Figs. \ref{fig:config31} to \ref{fig:config32i} show the mass of the inner binary held at .1, but with minor adjustments made to the inner binary separation to enhance stability.
+The final plot was very stable and allowed the simulation to run for far longer than all the other runs. This may be a resonance, please see Discussion. Although this final system was
+very stable, even minor deviations from the initial conditions immediately resulted in destabilisation of the system, and chaos.
 
@@
 \end{table}
-Discuss mass of inner bar ration of outer bar
+Discuss mass of inner bar ratio of outer bar
 The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilises. Run 32 was by far the most stable run and the orbits persisted until the integrator
 errored with a 'Step size underflow'. 1:21 could possibly be an orbital resonance.
-Fourier analysis?
-Rewrite in Python, automate, introduce machine laerning algoriythm
+%Fourier analysis?
+%Rewrite in Python, automate, introduce machine laerning algoriythm
 
-\citep{garzon}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}, \citep{macie5}
-\citep{macie6}, \citep{macie7}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{debattista}, \citep{malhotra}
+\subsection{Future Work}
+As this system has a multi-dimensional parameter space, finding stable configurations was very time consuming. I suggest further studies in this area try to automate this process.
+It may be possible for an automated process to automatically reject systems in which there are collisions, or if any of the bodies exits the system, or if the system fails 
+to persist for a certain number of orbits.
+
+%\citep{garzon}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}, \citep{macie5}
+%\citep{macie6}, \citep{macie7}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{debattista}, \citep{malhotra}
 

### 2017-08-20 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Temp

@@
 semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \cite{erwin2}.
-\cite{erwin1} believe that innder bars probably rotate independently of their outer bars and suggest that they are relatively long-lived strucutures.
+\cite{erwin1} believe that inner bars probably rotate independently of their outer bars and suggest that they are relatively long-lived structures.
 Inner bar size has been found to correlate slightly with outer bar size \citep{erwin3}.
@@
 
-Through the use of computer modeeling \cite{saha} showed that double bars can form without gas in a dark matter dominated halo.
+Through the use of computer modeling \cite{saha} showed that double bars can form without gas in a dark matter dominated halo.
 They produced a model where the inner bar was rotating at almost as slowly as the
 outer bar. The route to the formation of double bars maybe very different to that of a single bar.
-%\citep{moiseev}, \citep{sellwood}
+%\citep{moiseev}
 
@@
 Studies of three or more bodies interacting gravitationally dates back to the time of Newton.
-However it is only with the widespread use of computers that orbits of more than three bodies have been extensively studied.
+However, it is only with the widespread use of computers that orbits of more than three bodies have been extensively studied.
 Today
@@
 vary in complexity from following a single particle in a given, smooth galactic potential, to tens of thousands of interacting stars in a globular cluster, to billions of dark-matter particles in a 
-simulation of cosmological clustering \citep{binney}. Celestial mechanics during the preceding centuries has mainly focussed on the three body problem
+simulation of cosmological clustering \citep{binney}. Celestial mechanics during the preceding centuries has mainly focused on the three-body problem
 due to the complexities involved in adding more bodies.
@@
 The units have been selected so that the gravitational constant (G) is equal to one.
-In simulations it is natural to choose units where G = 1 and the units of length nand mass are conventionally chosen
+In simulations, it is natural to choose units where G = 1 and the units of length and mass are conventionally chosen
 so that the total mass of the system is M = 1 \citep{heggie}.
 
-The four body problem, specifically, has been studied extensively by \cite{steves}.
+The four-body problem, specifically, has been studied extensively by \cite{steves}.
 Here they model four equal mass bodies orbiting in circular coplanar orbits.
@@
 their research is being carried out at Glasgow Caledonian University.
-Approximately a century ago the Copenhagen problem was studied extensively in Copenhagen which consisted of a three body problem: two equal mass bodies and an infinitesimally small test particle.
+Approximately a century ago the Copenhagen problem was studied extensively in Copenhagen which consisted of a three-body problem: two equal mass bodies and an infinitesimally small test particle.
 \cite{szell} also
@@
 In a study of N-body simulations \cite{trenti} mentions a pioneering attempt by Holmberg in 1941 which used
-light bulbs and galvanometers to track the evolution of 37 particles. However it wasn't until the early seventies and the advent of computers that this field really got started. 
+light bulbs and galvanometers to track the evolution of 37 particles. However, it wasn't until the early seventies and the advent of computers that this field really got started. 
 Nowadays simulations can up to a million particles \citep{heggie} which is of particular relevance to the study of globular
 clusters. I refer the reader to \cite{aarseth} for much more detailed explanation of the field of N-body simulations.
- %\citep{collins}, \citep{moulton}
 
@@
 What follows is an attempt to verify the calculations in \citep{garzon} but using a simpler model of a double-barred galaxy using a double binary.
- %\citep{du2}, \citep{macie1}, \citep{macie2},  \citep{macie4}
- %\citep{malhotra}, \citep{manos}, \citep{shen2}
+
 \newpage
@@
 OdeInt performs integration using the Runge-Kutta method. The Runge-Kutta method is an extension of the Euler method of solving differential equations. However the Euler method
-is an example of first-order integration whilst Runge-Kutta is second and higher order and so is mcuh more accurate.
-The OdeInt uses a subroutine called RK4 which utilises fourth-order Runge-Kutta integration.
+is an example of first-order integration whilst Runge-Kutta is second and higher order and so is much more accurate.
+The OdeInt uses a subroutine called RK4 which utilizes fourth-order Runge-Kutta integration.
 It also uses a technique known as adaptive stepsize.
@@
 
-For a four body problem, Equation 1 results in the following:
+For a four-body problem, Equation 1 results in the following:
 \begin{equation}
@@
 \end{equation}
-These equations of motion are second-order ordinary differential equations (ODEs) but any second-oder ODE can be written as a system
+These equations of motion are second-order ordinary differential equations (ODEs) but any second-order ODE can be written as a system
 of two coupled first-order ODEs. As there are four bodies in the system, and they all have x and y positional components, this gives eight second-order 
-ODEs which leaves us with a total of sixteen simultanious first-order ODEs to solve.
+ODEs which leaves us with a total of sixteen simultaneous first-order ODEs to solve.
 These equations were encoded into Fortran. 
@@
 As a starting point both the outer binary and inner binary were placed in circular orbits and tested separately and then combined to ensure a stable
-starting point. The inner binary had an initial neglible mass.
+starting point. The inner binary had an initial negligible mass.
 The outer binary initial conditions were:
@@
 \end{figure}
-I then proceeded to perturb the system by gradually increasing the mass of the inner binary. As the system became unstable I componensated for
+I then proceeded to perturb the system by gradually increasing the mass of the inner binary. As the system became unstable I compensated for
 the increasing mass of the inner binary by increasing the velocity of the outer binary and also decreasing the velocity and increasing the binary 
@@
 \end{table}
-
-\newpage
+\clearpage
 \section{Results}
-As can be seen from Fig. \ref{fig:config1} even a neglible mass inner binary causes the outer binary to begin to separate and become less tightly bound.
+As can be seen from Fig. \ref{fig:config1} even a negligible mass inner binary causes the outer binary to begin to separate and become less tightly bound.
 Figures Fig. \ref{fig:config1} to \ref{fig:config32i} show plots of the double binary using the parameters given in Table \ref{tab:variables}.
-In many cases the inner binary is only visible as a point at the centre of the outer binary.
 As the mass of the inner binary increased the system became increasingly unstable and the effect on the outer binary became more pronounced.
 Even the `stable' configurations listed in Table \ref{tab:variables} hold for only a few orbits.
-
+Figs. \ref{fig:config1} and \ref{fig:config1i} show the initial orbits when the circular inner binary is placed within the outer binary. The placement of
+the inner binary can clearly be seen to affect the orbit of the outer binary even though the inner binary mass is very low.
 \begin{figure}[H]
@@
 \end{figure}
-\ref{fig:config1} and \ref{fig:config1i} show the inital orbits when the circular inner binary is placed within the outer binary.
 
+Figs. \ref{fig:config2} to \ref{fig:config3i} show plots where just the mass of the inner binary has been gradually increased. Nothing else was changed.
+By Fig. \ref{fig:config3i}, the inner binary has been pulled apart by the gravitational affect of the outer binary.
 \begin{figure}[H]
@@
 \end{figure}
-\ref{fig:config2} to \ref{fig:config3i} show plots where just the mass of the inner binary has been gradually increased. Nothing else was changed.
-By \ref{fig:config3i}, the inner binary can clearly be seen to be pulled apart by the gravitational affect of the outer binary.
 
+To counteract the effect of the increased mass of the inner binary in Figs. \ref{fig:config4} and \ref{fig:config4i} the initial separation of the 
+inner binary has been increased. This has a noticeable stabilizing effect on the inner binary which has regained approximate circularity.
 \begin{figure}[H]
@@
 \end{figure}
-To counteract the effect of the increased mass of the inner binary in \ref{fig:config4} and \ref{fig:config4i} the initial separation of the 
-inner binary has been increased.
 
+In figs. \ref{fig:config5} to \ref{fig:config10i} the initial velocity of the outer binary has also been gradually increased to also counteract the effect
+of the increased mass of the inner binary. In order for these systems to be relatively stable the initial velocity of the inner binary was reduced. At this point 
+both the inner and outer binaries had started to separate and the inner binary became increasingly elliptical.
 \begin{figure}[H]
@@
 \end{figure}
-In \ref{fig:config5} to \ref{fig:config10i} the initial velocity of the outer binary has also been gradually increased to also counteract the effect
-of the increased mass of the inner binary. In order for these systems to be relatively stable the initial velocity of the inner binary was reduced. At this point 
-both the inner and outer binaries had started to separate and become increasingly elliptical.
 
+In order to bring the orbits back into approximate circularity it was necessary to again increase the velocity of the inner binary. This is illustrated
+in Figs. \ref{fig:config11} and \ref{fig:config11i}. The mass of the inner binary has also increased considerably from the previous configurations.
 \begin{figure}[H]
@@
 \end{figure}
-In order to bring the orbits back into approximate circularity it was necessary to again increase the velocity of the inner binary. This is illustrated
-in \ref{fig:config11} and \ref{fig:config11i}. The mass of the inner binary has also increased considerably from the previous configurations.
 
+Figs. \ref{fig:config12} to \ref{fig:config13i} show the mass of the inner binary being slowly increased with adjustments made to both velocity of outer binary
+and separation of inner binary.
 \begin{figure}[H]
@@
 
-Figs. \ref{fig:config12} to \ref{fig:config13i} show the mass of the inner binary slowly increasing and adjustments made to both velocity of outer binary
-and separation of inner binary.
-
+As the mass of the inner binary was increased, it became necessary to reduce the velocity of the inner binary, Figs. \ref{fig:config14} to \ref{fig:config15i}.
+However, the inner binary was showing signs of increasingly chaotic orbits.
 \begin{figure}[H]
@@
 
-As the mass of the inner binary was increased, it became necessary to reduce the velocity of the inner binary, Figs. \ref{fig:config14} to \ref{fig:config15i}.
-
+Whilst holding the mass of the inner binary at .05, the system showed signs of more stability if the velocity of the inner binary was further reduced, Figs. \ref{fig:config16} to \ref{fig:config18i}.
 \begin{figure}[H]
@@
 
-Holding the mass of the inner binary at .05 the system showed signs of more stability if the velocity of the inner binary was further reduced, Figs. \ref{fig:config16} to \ref{fig:config18i}.
-
+Further reduction of the velocity of the inner binary allowed further increase of inner binary mass without losing stability, Figs. \ref{fig:config19} to \ref{fig:config21i}. The simulation is
+now stable enough for the outer binary to complete one orbit and proceed with a second.
 \begin{figure}[H]
@@
 
-Further reduction of the velocity of the inner binary allowed further increase of inner binary mass without losing stability, Figs. \ref{fig:config19} to \ref{fig:config21i}.
-
+With the mass of the inner binary now a quite considerable .1, the system showed more signs of stability if the velocity of the inner binary was further reduced, Figs.
+\ref{fig:config22} to \ref{fig:config25i}.
 \begin{figure}[H]
@@
 
-With the mass of the inner binary now a quite considerable .1, the system showed more signs of stability if the velocity of the inner binary was further reduced, Figs.
-\ref{fig:config22} to \ref{fig:config25i}.
-
+In Figs.
+\ref{fig:config26} to \ref{fig:config29i} the mass of the inner binary was increased to .12, and various adjustments made to inner separation, velocity and outer velocity to maintain stability.
+Both binaries, however, were now on highly elliptical orbits.
 \begin{figure}[H]
@@
 
-In Figs.
-\ref{fig:config26} to \ref{fig:config29i} the mass of the inner binary was increased to .12, and various adjustments made to inner separation, velocity and outer velocity to maintain stability.
-
+Having increased the mass of the inner binary to .12 on the previous three plots it was decided, that as inner bars are generally $\approx12\%$ of the mass of the outer bar, the mass of the inner binary could stay at .1
+as this is 20\% of the mass of the outer binary, Figs. \ref{fig:config30} and \ref{fig:config30i}.
 \begin{figure}[H]
@@
 
-Having increased the mass of the inner binary to .12 on the previous three plots it was decided, that as inner bars are generally $\approx12\%$ of the mass of the outer bar, the mass of the inner binary could stay at .1
-as this is 20\% of the mass of the outer binary, Figs. \ref{fig:config30} and \ref{fig:config30i}.
-
+Finally, Figs. \ref{fig:config31} to \ref{fig:config32i} show the mass of the inner binary held at .1, but with minor adjustments made to the inner binary separation to enhance stability.
+The final plot was very stable and allowed the simulation to run for far longer than all the other runs. This may be a resonance, please see Discussion. Although this final system was
+very stable, even minor deviations from the initial conditions immediately resulted in destabilization of the system, and chaos.
 \begin{figure}[H]
@@
 
-Finally, Figs. \ref{fig:config31} to \ref{fig:config32i} show the mass of the inner binary held at .1, but with minor adjustments made to the inner binary separation to enhance stability.
-The final plot was very stable and allowed the simulation to run for far longer than all the other runs. This may be a resonance, please see Discussion. Although this final system was
-very stable, even minor deviations from the initial conditions immediately resulted in destabilisation of the system, and chaos.
-
 \newpage
-\section{Disussion}
-The final three runs (configurations 30, 31, and 32) were by far the most stable. The orbital periods vary slightly with time so it is difficult to give accurate figures, so
-I am using the periods of the first orbit. The following table shows the periods of the two binaries together with the ratios of the corresponding periods.
+\section{Discussion}
+The final three runs (configurations 30, 31, and 32) were by far the most stable. The orbital periods vary slightly with time so it is difficult to give accurate figures, so the 
+periods of the first orbit were used to calculate the ratio of the periods between inner and outer binaries.
+The following table shows the periods of the two binaries together with the ratios of the corresponding periods.
 \begin{table}[ht!]
@@
 \end{table}
-Discuss mass of inner bar ratio of outer bar
-The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilises. Run 32 was by far the most stable run and the orbits persisted until the integrator
-errored with a 'Step size underflow'. 1:21 could possibly be an orbital resonance.
-%Fourier analysis?
-%Rewrite in Python, automate, introduce machine laerning algoriythm
+
+Configuration 32 was by far the longest running simulation showing almost three complete orbits of the outer binary and many orbits of the inner binary. It showed no signs of destabilizing and the 
+simulation only ended because of a `Step size underflow' in the integrator. This error occurs when the adaptive stepsize becomes too small and the integrator is effectively not making any
+progress with the integration and so ends the simulation.
+The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilizes.
+1:21 could therefore possibly be an orbital resonance.
+Attempts at modifying the initial conditions of configuration 32 even slightly resulted in chaotic systems.
+
+Unfortunately, a 1:21 resonance doesn't represent the observational data of double-barred galaxies.
+\cite{font} found that the inner bar rotates more rapidly than the outer bar by a factor between 3.3 and 3.6.
+Whilst \cite{macie4} predicted values for the ratio of bar angular rates between 1.6 and 2.7.
 
@@
 As this system has a multi-dimensional parameter space, finding stable configurations was very time consuming. I suggest further studies in this area try to automate this process.
-It may be possible for an automated process to automatically reject systems in which there are collisions, or if any of the bodies exits the system, or if the system fails 
-to persist for a certain number of orbits.
+It may be possible for an automated process to automatically reject systems in which there are collisions, or if any of the bodies exit the system, or if the system fails 
+to persist for a certain number of orbits. In this way, it may be possible to find further orbital resonances.
 
@@
 %\citep{macie6}, \citep{macie7}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{debattista}, \citep{malhotra}
+ %\citep{du2} \citep{manos}, \citep{shen2}
 
@@
 \bibitem[Binney and Tremaine(2008)]{binney}Binney, J., Tremaine, S., Galactic Dynamics, 2008, Princeton University Press
-%\bibitem[Collins(2004)]{collins}Collins, G., The Foundations Of Celestial Mechanics, 2004, Pachart Publishing House
 \bibitem[de Vaucouleurs(1975)]{vaucouleurs}de Vaucouleurs, G., 1975, ApJS,29, 193
@@
 \bibitem[Erwin and Sparke(2002)]{erwin1}Erwin, P., Sparke, L., Double Bars, Inner Disks, and Nuclear Rings in Early-Type Disk Galaxies, 2002, The Astronomical Journal, 124:65–77, 2002 July
+\bibitem[Font el. al.(2014)]{font}Font, J., Beckman, J., E., Zaragoza-Cardiel, J., Fathi, K., Epinat, B., Amram, P., The ratio of pattern speeds in double-barred galaxies, arXiv:1408.4274v1
 \bibitem[Garzon(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1
@@
 \bibitem[Maciejewski(2008)]{macie3}Maciejewski, W., Orbits in corotating and counterrotating double bars, 2008, arXiv:0801.1471v1
-%\bibitem[Maciejewski and Athanassoula(2008)]{macie4}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
+\bibitem[Maciejewski and Athanassoula(2008)]{macie4}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1
 \bibitem[Maciejewski and Small(2010)]{macie5}Maciejewski, W., Small, E., Orbital Support of Fast and Slow Inner Bars in Double Barred Galaxies, 2010, arXiv:1006.4574v1
@@
 \bibitem[Malhotra(1998)]{malhotra}Malhotra, R., Orbital Resonances and Chaos in the Solar System, 1998, Solar system Formation and Evolution ASP Conference Series, Vol. 149, 1998
-\bibitem[Manos and Athanassoula(2011)]{manos}Manos, T., Athanassoula, E., Regular and chaotic orbits in barred galaxies - I.
+%\bibitem[Manos and Athanassoula(2011)]{manos}Manos, T., Athanassoula, E., Regular and chaotic orbits in barred galaxies - I.
 Applying the SALI/GALI method to explore their distribution in several models, 2011, arXiv:1102.1157v2
 \bibitem[Moiseev(2001)]{moiseev}Moiseev, A., V., Velocity dispersion of stars and gas motion in double-barred galaxies, 2001, arXiv:astro-ph/0111220v1
-%\bibitem[Moulton(1914)]{moulton}Moulton, F., An Introduction to Celestial Mechanics, 1914, Palmyrin Library 
 \bibitem[Press et. al.(1992)]{press}Press, W., H., Teukolsky, S., A., Vetterling, W., T., Flannery, B., P., Numerical Recipes in Fortran 77 The Art of Scientific Computing, 2nd edition
 \bibitem[Saha and Maciejewski(2013)]{saha}Saha, K., Maciejewski, W., Spontaneous formation of double bars in dark matter dominated galaxies, 2013, arXiv:1304.7108v1
-%\bibitem[Sellwood and Wilkinson(2006)]{sellwood}Sellwood, J., Wilkinson, A., Dynamics of Barred Galaxies, 2006, arXiv:astro-ph/0608665v1
 \bibitem[Shen and Debattista(2008)]{shen1}Shen, J., Debattista, V., Long-lived double-barred galaxies in N-body simulations, 2008, Mem. S.A.It. Vol. 00, 169

### 2017-08-21 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Draft

@@
 \DeclareGraphicsExtensions{.eps}
-%\setlength\abovecaptionskip{-5pt}
 \author{Howard Kinsman}
@@
 The current estimates of double-barred galaxy frequencies are $\approx30\%$ of barred galaxies and $\approx20\%$ of all galaxies \citep{erwin3}.
-\cite{erwin2} surveyed 38 barred galaxies and found a total of 10 double-barred galaxies with no preference for Hubble type. They found that inner bars were about 12\% the size of the outer bars with
-semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \cite{erwin2}.
+\cite{erwin2} surveyed 38 barred galaxies and found a total of 10 double-barred galaxies with no preference for Hubble type. He found that inner bars were about 12\% the size of the outer bars with
+semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \citep{erwin2}.
 \cite{erwin1} believe that inner bars probably rotate independently of their outer bars and suggest that they are relatively long-lived structures.
@@
 
-It has been suggested that nested bars may provide a mechanism for fueling active galactic nuclei \citep{shlosman} however more recently \citep{erwin3} believe that inner bars only play a minor role in this
+It has been suggested that nested bars may provide a mechanism for fueling active galactic nuclei \citep{shlosman} however more recently, \citep{erwin3}, it was found that inner bars only play a minor role in this
 activity.
-\citep{erwin1} compared single and double-barred galaxies and found that there was no significant difference in nuclear activity.
+\cite{erwin1} compared single and double-barred galaxies and found that there was no significant difference in nuclear activity.
 To investigate this theory further \cite{lorenzo1} carried out a
@@
 
-Methods for the study of N-body simulations have changed considerably since the advent of computers.
-In a study of N-body simulations \cite{trenti} mentions a pioneering attempt by Holmberg in 1941 which used
+Methods for the study of N-body simulations have changed considerably over the years. In a study of N-body simulations \cite{trenti} mentions a pioneering attempt by Holmberg in 1941 which used 
 light bulbs and galvanometers to track the evolution of 37 particles. However, it wasn't until the early seventies and the advent of computers that this field really got started. 
@@
 themselves as gravitating systems.
-A gravitational potential consisting of two concentric bisymmetric bars that are independently rotating is not stationary in any reference frame \cite{macie7}.
+A gravitational potential consisting of two concentric bisymmetric bars that are independently rotating is not stationary in any reference frame \citep{macie7}.
 In the potential
@@
 These were N-body/hydrodynamical simulations.
-They found that the two bar lengths, ratio of
+He found that the two bar lengths, ratio of
 pattern speeds, as well as the age of the inner stellar bar populations in the simulations matched well with observations.
-They concluded that because, unlike previous simulations, there was no overlap between primary bar
+He concluded that because, unlike previous simulations, there was no overlap between primary bar
 inner Lindblad resonance (ILR) and the inner bar corotation (CR) or any kind of resonance overlap, then some other kind of mechanism must feed the central waves \citep{wozniak}.
@@
 rotation of deformable rotation. They found that the bars oscillate in size according to the relative angle between the bars. This is broadly in agreement with \citep{debattista}.
-What follows is an attempt to verify the calculations in \citep{garzon} but using a simpler model of a double-barred galaxy using a double binary.
+What follows is an attempt to verify the calculations in \cite{garzon} but using a simpler model of a double-barred galaxy using a double binary.
 
@@
 \end{figure}
-I then proceeded to perturb the system by gradually increasing the mass of the inner binary. As the system became unstable I compensated for
-the increasing mass of the inner binary by increasing the velocity of the outer binary and also decreasing the velocity and increasing the binary 
+The system was then perturbed by gradually increasing the mass of the inner binary. As the system became unstable the increasing mass of the inner binary 
+was compensated for by increasing the velocity of the outer binary and also decreasing the velocity and increasing the binary 
 separation of the inner binary. In many of the configurations the systems were inherently unstable and produced collisions with some 
@@
 Attempts at modifying the initial conditions of configuration 32 even slightly resulted in chaotic systems.
+The concept of a resonance is very familiar to us all: pushing someone on a playground swing at its natural
+frequency will rapidly increase the swing amplitude \citep{malhotra}.
+Resonances occur naturally in the Solar system and there are three main types of resonance: spin-orbit, secular and mean motion \citep{malhotra}.
+The most obvious example of a spin-orbit resonance is the Earth/Moon system where the Moon is spin-locked and only one side of the Moon is visible from the Earth.
+Secular resonances occur where the precession of two
+orbits is synchronised.
+However, the 1:21 resonance, is an example of a mean-motion resonance where the orbital period of two bodies are close to a ratio of small integer \citep{malhotra}.
+Orbital resonances
+generally results in destabilizing a system however occasionally they may be stable and remain in resonance.
+In the standard model of double-barred galaxies \citep{macie7} it is assumed that resonant coupling between bars is present: CR of the inner bar coincides with the ILR of the outer bar.
+Although \cite{macie4}
+found that resonant coupling between the bars may not be the dominant factor reducing chaos in the system.
 
@@
 
-%\citep{garzon}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}, \citep{macie5}
-%\citep{macie6}, \citep{macie7}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{debattista}, \citep{malhotra}
- %\citep{du2} \citep{manos}, \citep{shen2}
-
 \newpage
@@
 \bibitem[Du et. al.(2015)]{du1}Du, M., Shen, J., Debattista, V., Forming Double-Barred Galaxies from Dynamically Cool Inner Disks, 2015, The Astrophysical Journal, 804:139 (10pp), 2015 May 10 
-%\bibitem[Du et. al.(2016)]{du2}Du, M., Debattista, V., Shen, J., Kimematic Properties of Double-Barred Galaxies: Simulations vs. Integral-Field Observations, 2016, arXiv:1607.00585v1
 \bibitem[Erwin(2003)]{erwin3}Erwin, P., Double-Barred Galaxies I. A Catalog of Barred Galaxies with Stellar Secondary Bars and Inner Disks, 2003, arXiv:astro-ph/0310806v2
@@
 \bibitem[Font el. al.(2014)]{font}Font, J., Beckman, J., E., Zaragoza-Cardiel, J., Fathi, K., Epinat, B., Amram, P., The ratio of pattern speeds in double-barred galaxies, arXiv:1408.4274v1
-\bibitem[Garzon(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1
+\bibitem[Garzon and Lopez-Corredoira(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1
 \bibitem[Heggie and Hut(2003)]{heggie}Heggie D., Hut, P., The Gravitational Million-Body Problem, 2003, Cambridge University Press
@@
 \bibitem[Malhotra(1998)]{malhotra}Malhotra, R., Orbital Resonances and Chaos in the Solar System, 1998, Solar system Formation and Evolution ASP Conference Series, Vol. 149, 1998
-%\bibitem[Manos and Athanassoula(2011)]{manos}Manos, T., Athanassoula, E., Regular and chaotic orbits in barred galaxies - I.
-Applying the SALI/GALI method to explore their distribution in several models, 2011, arXiv:1102.1157v2
 \bibitem[Moiseev(2001)]{moiseev}Moiseev, A., V., Velocity dispersion of stars and gas motion in double-barred galaxies, 2001, arXiv:astro-ph/0111220v1

### 2017-09-07 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Final

@@
 \begin{abstract}
-TODO
+A brief literature review of double-barred galaxies and N-body simulations followed by an attempt at modelling a double-barred galaxy using a 4-body simulation of a double binary.
+The aim of the project was to
+find orbital resonances which may be responsible for the stability of double-barred galaxies. A 1:21 resonance was found but this is not representative of the observed pattern speeds of nested bars.
 \end{abstract}
@@
 
-It has been suggested that nested bars may provide a mechanism for fueling active galactic nuclei \citep{shlosman} however more recently, \citep{erwin3}, it was found that inner bars only play a minor role in this
+It has been suggested that nested bars may provide a mechanism for fuelling active galactic nuclei \citep{shlosman} however more recently, \citep{erwin3}, it was found that inner bars only play a minor role in this
 activity.
@@
 the bulge and inner bar show similar stellar population properties whilst the outer bar was less metal-rich.
-This result implied that, for the case of NGC 357, the outer bar was formed in a shorter timescale than
+This result implied that, for the case of NGC 357, the outer bar was formed in a shorter time-scale than
 the inner structures and so disagrees with traditional idea that gas flown along the outer structure triggers star formation causing the formation of the inner structure.
@@
 
-Through the use of computer modeling \cite{saha} showed that double bars can form without gas in a dark matter dominated halo.
+Through the use of computer modelling \cite{saha} showed that double bars can form without gas in a dark matter dominated halo.
 They produced a model where the inner bar was rotating at almost as slowly as the
@@
 themselves as gravitating systems.
-A gravitational potential consisting of two concentric bisymmetric bars that are independently rotating is not stationary in any reference frame \citep{macie7}.
+A gravitational potential consisting of two concentric bi-symmetric bars that are independently rotating is not stationary in any reference frame \citep{macie7}.
 In the potential
@@
 the conclusion that loops supporting the inner bar are thicker with parallel bars and thinner with perpendicular bars \citep{macie3}.
-This also suggets that the two bars do not rotate through each other as rigid bodies
-but the inner bar should pulsate and that the bars should spend more time nearly orthognal\citep{macie3}. These predictions were confirmed by \citep{debattista} and \citep{shen1}. 
+This also suggests that the two bars do not rotate through each other as rigid bodies
+but the inner bar should pulsate and that the bars should spend more time nearly orthogonal\citep{macie3}. These predictions were confirmed by \cite{debattista} and \cite{shen1}. 
 This was further confirmed by \cite{macie5} who found
-that the inner bar should end well within its corotation radius and that the inner bar extends further out for lower pattern speeds than higher ones; the inner bar pulsates; and that the angular velocity of the inner
+that the inner bar should end well within its co-rotation radius and that the inner bar extends further out for lower pattern speeds than higher ones; the inner bar pulsates; and that the angular velocity of the inner
 bar is not uniform and that faster inner bars rotate more coherently.
@@
 He concluded that because, unlike previous simulations, there was no overlap between primary bar
-inner Lindblad resonance (ILR) and the inner bar corotation (CR) or any kind of resonance overlap, then some other kind of mechanism must feed the central waves \citep{wozniak}.
+inner Lindblad resonance (ILR) and the inner bar co-rotation (CR) or any kind of resonance overlap, then some other kind of mechanism must feed the central waves \citep{wozniak}.
 It is suggested that star
@@
 They considered two cases purely analytically: rotation of rigid bodies and
-rotation of deformable rotation. They found that the bars oscillate in size according to the relative angle between the bars. This is broadly in agreement with \citep{debattista}.
+rotation of deformable rotation. They found that the bars oscillate in size according to the relative angle between the bars. This is broadly in agreement with \cite{debattista}.
 What follows is an attempt to verify the calculations in \cite{garzon} but using a simpler model of a double-barred galaxy using a double binary.
@@
 The OdeInt uses a subroutine called RK4 which utilizes fourth-order Runge-Kutta integration.
-It also uses a technique known as adaptive stepsize.
+It also uses a technique known as adaptive step-size.
 This involves adjusting the size of the integration steps to ensure the same level of accuracy however rapidly the function being integrated changes.
@@
 Plots of the two initial binaries are shown in Fig. \ref{fig:outerbinary} and 
-\ref{fig:innerbinary}. The velocity of the inner binary is high and therefore few points are seen in this initial plot.
+\ref{fig:innerbinary}.
 \begin{figure}[H]
@@
 To counteract the effect of the increased mass of the inner binary in Figs. \ref{fig:config4} and \ref{fig:config4i} the initial separation of the 
-inner binary has been increased. This has a noticeable stabilizing effect on the inner binary which has regained approximate circularity.
+inner binary has been increased. This has a noticeable stabilising effect on the inner binary which has regained approximate circularity.
 \begin{figure}[H]
@@
 The final plot was very stable and allowed the simulation to run for far longer than all the other runs. This may be a resonance, please see Discussion. Although this final system was
-very stable, even minor deviations from the initial conditions immediately resulted in destabilization of the system, and chaos.
+very stable, even minor deviations from the initial conditions immediately resulted in destabilisation of the system, and chaos.
 \begin{figure}[H]
@@
 
-Configuration 32 was by far the longest running simulation showing almost three complete orbits of the outer binary and many orbits of the inner binary. It showed no signs of destabilizing and the 
-simulation only ended because of a `Step size underflow' in the integrator. This error occurs when the adaptive stepsize becomes too small and the integrator is effectively not making any
+Configuration 32 was by far the longest running simulation showing almost three complete orbits of the outer binary and many orbits of the inner binary. It showed no signs of destabilising and the 
+simulation only ended because of a `Step size underflow' in the integrator. This error occurs when the adaptive step-size becomes too small and the integrator is effectively not making any
 progress with the integration and so ends the simulation.
-The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilizes.
+The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilises.
 1:21 could therefore possibly be an orbital resonance.
@@
 Orbital resonances
-generally results in destabilizing a system however occasionally they may be stable and remain in resonance.
+generally results in destabilising a system however occasionally they may be stable and remain in resonance.
 In the standard model of double-barred galaxies \citep{macie7} it is assumed that resonant coupling between bars is present: CR of the inner bar coincides with the ILR of the outer bar.
@@
 Whilst \cite{macie4} predicted values for the ratio of bar angular rates between 1.6 and 2.7.
+With these studies in mind it may be worthwhile exploring the double binary model further and searching for stable
+resonances around the 1:2 or 1:3 parameter space - see next section.
 
 \subsection{Future Work}
-As this system has a multi-dimensional parameter space, finding stable configurations was very time consuming. I suggest further studies in this area try to automate this process.
+As this system has a multidimensional parameter space, finding stable configurations was very time consuming. I suggest further studies in this area try to automate this process.
 It may be possible for an automated process to automatically reject systems in which there are collisions, or if any of the bodies exit the system, or if the system fails 

### 2017-09-13 workspace/DoubleBars/Final/MScProjectDoubleBars.tex
Done!

--- 2017-09-07 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
@@
 The aim of the project was to
-find orbital resonances which may be responsible for the stability of double-barred galaxies. A 1:21 resonance was found but this is not representative of the observed pattern speeds of nested bars.
+find orbital resonances which may be responsible for the stability of double-barred galaxies and attempt at verifying the calculations carried out by \cite{garzon}. 
+A 1:21 resonance was found but this is not representative of the observed pattern speeds of nested bars.
 \end{abstract}
@@
 To counteract the effect of the increased mass of the inner binary in Figs. \ref{fig:config4} and \ref{fig:config4i} the initial separation of the 
-inner binary has been increased. This has a noticeable stabilising effect on the inner binary which has regained approximate circularity.
+inner binary has been increased by .003 to .004. This has a noticeable stabilising effect on the inner binary which has regained approximate circularity.
 \begin{figure}[H]
@@
 
-In figs. \ref{fig:config5} to \ref{fig:config10i} the initial velocity of the outer binary has also been gradually increased to also counteract the effect
+In Figs. \ref{fig:config5} to \ref{fig:config10i} the initial velocity of the outer binary has also been gradually increased to also counteract the effect
 of the increased mass of the inner binary. In order for these systems to be relatively stable the initial velocity of the inner binary was reduced. At this point 
@@
 In order to bring the orbits back into approximate circularity it was necessary to again increase the velocity of the inner binary. This is illustrated
-in Figs. \ref{fig:config11} and \ref{fig:config11i}. The mass of the inner binary has also increased considerably from the previous configurations.
+in Figs. \ref{fig:config11} and \ref{fig:config11i}. The mass of the inner binary has also increased considerably from the previous configurations to .025.
 \begin{figure}[H]
@@
 Further reduction of the velocity of the inner binary allowed further increase of inner binary mass without losing stability, Figs. \ref{fig:config19} to \ref{fig:config21i}. The simulation is
-now stable enough for the outer binary to complete one orbit and proceed with a second.
+now stable enough for the outer binary to complete one orbit and proceed with a second. It was found that by increasing the mass and separation of the inner binary by .01, and increasing the 
+velocity of the outer binary by .05, whilst steadily reducing the velocity of the inner binary resulted in the fairly stable configurations shown here.
 \begin{figure}[H]
@@
 Secular resonances occur where the precession of two
-orbits is synchronised.
+orbits are synchronised.
 However, the 1:21 resonance, is an example of a mean-motion resonance where the orbital period of two bodies are close to a ratio of small integer \citep{malhotra}.
@@
 resonances around the 1:2 or 1:3 parameter space - see next section.
+A comparison with the calculations carried out by \cite{garzon} was not completed because the results here seem to be dominated by free
+oscillations rather than forced oscillations generated by the inner bar \citep{macie1}.
 
 \subsection{Future Work}
+As time constraints prevented the Garzon comparison from being carried out, a future project would need to minimise the free oscillations in order to compare it with Garzon's prediction.
 As this system has a multidimensional parameter space, finding stable configurations was very time consuming. I suggest further studies in this area try to automate this process.
@@
 \bibitem[Aarseth(2003)]{aarseth}Aarseth, S., Gravitational N-Body Simulations, 2003, Cambridge University Press
-\bibitem[Alvarez–Ramirez and Medina(2014)]{alvarez}Alvarez–Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014
+\bibitem[Alvarez-Ramirez and Medina(2014)]{alvarez}Alvarez-Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014
 \bibitem[Binney and Tremaine(2008)]{binney}Binney, J., Tremaine, S., Galactic Dynamics, 2008, Princeton University Press
@@
 \bibitem[Maciejewski et. al.(2001)]{macie8}Maciejewski, W., Teuben, P., J., Sparke, L., S., Stone, J., M., Gas Inflow in Barred Galaxies - Effects of Secondary Bars, arXiv:astro-ph/0109431v1
-%\bibitem[Maciejewski(2002)]{macie1}Maciejewski, W., Constraints on nested bars – implications for gas inflow, 2002, arXiv:astro-ph/0202110v1
-%\bibitem[Maciejewski(2003)]{macie2}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1
 \bibitem[Maciejewski(2008)]{macie3}Maciejewski, W., Orbits in corotating and counterrotating double bars, 2008, arXiv:0801.1471v1
+\bibitem[Maciejewski(2017)]{macie1}Maciejewski, W., 11th September 2017 [email]
 \bibitem[Maciejewski and Athanassoula(2008)]{macie4}Maciejewski, W., Athanassoula, A., Regular motions in double bars. II. Survey of trajectories and 23 models, 2008, arXiv:0805.3967v1

### 2017-09-13 workspace/DoubleBars/Final/DoubleBars.f90

@@ -1,0 +1,278 @@
+ program doublebars
+   implicit none
+   integer, parameter :: NVAR=16 ,NMAX=50,KMAXX=10000
+   integer :: kmax=0,i,j
+   real :: ys(NVAR),xp(KMAXX),yp(NMAX,KMAXX),m1,m2,m3,m4,m, start, end, step, energy(KMAXX)
+  real :: period(4, KMAXX)
+   real, parameter :: PI=4*ATAN(1.)
+
+   call initialconditions
+   call odeint(ys,start,end,step,1.e-8,0.0)
+
+   do i=1,kmax
+      call calcenergy
+      write(9,33) xp(i), (yp(j,i),j=1,NVAR),i,energy(i)
+      if(xp(i)==end) then
+          exit
+      end if
+   end do
+ 33 format(17f10.4,i5,f10.4)
+
+ contains
+
+   subroutine initialconditions
+
+     step = 1.e-7 !vary step size
+     start=0.
+     end=35 !vary end time
+
+      m1=.5   !blue
+      m2=.5  !red
+      m3=0.1    !black
+      m4=0.1   !yellow
+
+     ys(1) = -.5     ! m1,x
+     ys(2) = 0  ! m1,vx
+     ys(3) = 0       ! m1,y
+     ys(4) = -1		!m1,vy
+     
+     ys(5) = .5         ! m2,x
+     ys(6) = 0          ! m2,vx
+     ys(7) = 0         ! m2,y
+     ys(8) = 1	!m2,vy
+
+     ys(9) = -.1021    ! m3,x
+     ys(10) = 0       ! m3,vx
+     ys(11) = 0        ! m3,y
+     ys(12) = -.1		!m3, vy
+
+     ys(13) = .1021     ! m4,x
+     ys(14) = 0      ! m4,vx
+     ys(15) = 0        ! m4,y
+     ys(16) = .1		!m4, vy
+
+   end subroutine initialconditions
+
+   subroutine derivs(x,y,dydx)
+     real, intent(in) :: x,y(NVAR)
+     real, intent(out) :: dydx(NVAR)
+     real :: dist1,dist2,dist3, dist4, dist5, dist6,dist7,dist8,dist9,dist10,dist11,dist12,r1,r2,r3,r4,r5,r6
+
+    dist1 = y(5)-y(1)  !x2-x1
+    dist2 = y(7)-y(3) !y2-y1
[…]
